\setboolean{@twoside}{true} 
\clearpage
\pagestyle{dictionarystyle}

\chapter{Dictionary}
\label{chap:dictionary}
\vspace{10.5cm}

\subsection{A}

\begin{multicols*}{2}

    \dictentry{ADL}{}

    \dictentry{Allocate}{to assign, allot, distribute, or set apart for 
    a particular purpose}

    \dictentry{Atomic}{Objects of atomic types are the only C++ objects 
    that are free from data races; that is, if one thread writes to an 
    atomic object while another thread reads from it, the behavior is 
    well-defined. In addition, accesses to atomic objects may establish 
    inter-thread synchronization and order non-atomic memory accesses 
    as specified by \textbf{\texttt{std::memory\char`_order}}.}
    
\end{multicols*}

\subsection{B}
\begin{multicols*}{2}

    \dictentry{branch}{a mean for teams to develop features giving them 
    a separate workspace for their code. The branch to create are part 
    of the branching strategy, which could be different depending on 
    the specific svn used.}

\end{multicols*}

\subsection{C}
\begin{multicols*}{2}

    \dictentry{constexpr}{a C++ specifier which declares that it is 
    possible to evaluate the value of the function or variable 
    at compile time.}

\end{multicols*}

\subsection{D}

\begin{multicols*}{2}

    \dictentry{data race}{}

    \dictentry{deadlock}{any situation in which no member of some group 
    of entities can proceed because each waits for another member, 
    including itself, to take action, such as sending a message or, 
    more commonly, releasing a lock}

    \dictentry{duck typing}{a concept related to \textit{dynamic typing}, 
    where the type or the class of an object is less important than the 
    methods it defines}
    
    \dictentry{dynamic binding}{}

\end{multicols*}

\subsection{E}
\begin{multicols*}{2}

    \dictentry{encapsulation}{preventing unauthorized access to some piece of information or functionality}

\end{multicols*}

%~ \subsection{F}
%~ \begin{multicols*}{2}\end{multicols*}

%~ \subsection{G}
%~ \begin{multicols*}{2}\end{multicols*}

\subsection{H}

\begin{multicols*}{2}

    \dictentry{Heap}{a large pool of memory used to satisfy memory 
    request by allocating part.}

\end{multicols*}

\subsection{I}

\begin{multicols*}{2}

    \dictentry{Idiom}{a group of code fragments sharing an equivalent 
    semantic role (meaning), which recurs frequently across software 
    projects often expressing a special feature in one or more 
    programming languages or libraries.}

\end{multicols*}

%~ \subsection{J}
%~ \begin{multicols*}{2}\end{multicols*}

%~ \subsection{K}
%~ \begin{multicols*}{2}\end{multicols*}

%~ \subsection{L}
%~ \begin{multicols*}{2}\end{multicols*}

\subsection{M}
\begin{multicols*}{2}

    \dictentry{method chaining}{}

    \dictentry{most vexing parsing}{}
\end{multicols*}

%~ \subsection{N}
%~ \begin{multicols*}{2}\end{multicols*}

\subsection{O}

\begin{multicols*}{2}

    \dictentry{}{region of storage with associated semantics}

    \dictentry{Overloading}{a compile time polymorphism achieved 
    specifying more than one function with the same name but different 
    signature in the same scope.}
    
    \dictentry{Overridding}{the act of redefining a method of a base 
    class in a derived class, using the same name, return type and 
    parameters.}

\end{multicols*}

\subsection{P}

\begin{multicols*}{2}

    \dictentry{Polymorphysm}
              {ability of different objects to be accessed by a common 
              interface.}
                      
\end{multicols*}

%~ \subsection{Q}
%~ \begin{multicols*}{2}\end{multicols*}

\subsection{R}

\begin{multicols*}{2}
    
    \dictentry{Race Condition}{undesirable situation which occurs when 
    two instructions from different threads access the same memory 
    location, at least one of these accesses is a write and there is no 
    synchronization that is mandating any particular order among these 
    accesses.}
    
    \dictentry{RAII}{acronym for Resource Acquisition Is Initialization.}
    
    \dictentry{RBV optimization}{see~\cite{isocpp-rbv-optimization}.}
    
    \dictentry{RVO}{acronym for Return Value Optimization.}

\end{multicols*}

\subsection{S}
    
    \dictentry{static typing}{}
    
\begin{multicols*}{2}
    
    \dictentry{SFINAE}{acronym for \textbf{Substitution} \textbf{Failure} 
    \textbf{Is} \textbf{Not} \textbf{An} \textbf{Error}, a technique 
    used in template programming which discards the specialization from 
    the overloaded set avoiding a compile error, when substituting the 
    explicitly specified or deduced type for the template parameter fails.}
    
    \dictentry{Stack}{LIFO-ordered contiguous region of temporary memory 
    whose size is known to the compiler, which manages to allocate and 
    deallocate blocks of memory for the stack variables like functions' 
    local data. . If a region of memory lies on the thread's stack, 
    that memory is said to have been allocated on the stack, i.e. 
    stack-based memory allocation (SBMA).}

\end{multicols*}


\newpage{}
%~ \subsection{T}
%~ \begin{multicols*}{2}\end{multicols*}

%~ \subsection{U}
%~ \begin{multicols*}{2}\end{multicols*}

\subsection{V}
\begin{multicols*}{2}
    
    \dictentry{virtual member function}{a mean to allow derived classes 
    to replace the implementation provided by the base class. The 
    compiler makes sure to call the correct replacement when 
    the object is of the derived class and also when the object is 
    accessed through a pointer to the base rather than to the derived 
    class.}
    
\end{multicols*}

%~ \subsection{W}
%~ \begin{multicols*}{2}\end{multicols*}

%~ \subsection{X}
%~ \begin{multicols*}{2}\end{multicols*}

%~ \subsection{Y}
%~ \begin{multicols*}{2}\end{multicols*}

%~ \subsection{Z}
%~ \begin{multicols*}{2}\end{multicols*}



\setboolean{@twoside}{false} 
\pagestyle{basicstyle}


