\chapter{Event Driven I/O}
\label{chap:eventdrivenio}

Non-blocking sockets (Chapter~\ref{chap:linuxnetworking}) return immediately with 
\texttt{EAGAIN} when there is nothing to do. That alone is not enough: you still need a way to 
\highlight{wait until an fd is ready} without spinning forever or blocking on one socket at a time.
\newline
\highlightbf{I/O multiplexing} answers ``which of my fds can I read/write now?'' Classic APIs are 
\texttt{select} and \texttt{poll}; Linux's scalable answer is \texttt{epoll}. Busy-polling and 
kernel bypass come later (Chapters~\ref{chap:lowlatency} and~\ref{chap:kernelbypass}).

\section{\texttt{select()}}
\label{sec:select}
\texttt{select(2)}~\cite{man-select} is the oldest portable multiplexer. Interviewers still ask 
about it to contrast with \texttt{epoll}.

\subsection{How It Works}
\label{subsec:select-how-it-works}
You pass three bitmasks (\texttt{fd\char`_set}): readable, writable, and exceptional conditions, 
plus an optional timeout. The kernel blocks (or returns after the timeout) and \highlight{rewrites} 
the sets so only ready fds remain set.
\begin{lstlisting}[language=C++]
#include <sys/select.h>

fd_set rd;
FD_ZERO(&rd);
FD_SET(listen_fd, &rd);
int nfds = listen_fd + 1;

timeval tv{};
tv.tv_sec = 0;
tv.tv_usec = 1000;  // 1 ms

int n = select(nfds, &rd, nullptr, nullptr, &tv);
if (n > 0 && FD_ISSET(listen_fd, &rd)) {
    // accept / recv is safe to attempt
}
\end{lstlisting}
You must \highlight{rebuild} the \texttt{fd\char`_set} before every call (the kernel mutates it). 
\texttt{nfds} is one past the highest fd number you care about.

\subsection{Limits}
\label{subsec:select-limits}
\begin{itemize}
    \item \highlightbf{\texttt{FD\char`_SETSIZE}} --- typically 1024. Fds above that cannot be 
          tracked without rebuilding libc / using non-portable tricks.
    \item \highlightbf{Linear cost} --- the kernel walks your interest set (cost linear in the 
          number of fds); user space also copies and scans bitmasks. Cost grows with the 
          \highlight{number of monitored fds}, not only with ready ones.
    \item \highlightbf{Stateful rebuild} --- every wait starts from scratch; no persistent 
          interest list in the kernel.
\end{itemize}
Fine for tiny tools; the wrong model for thousands of connections.

\section{\texttt{poll()}}
\label{sec:poll}
\texttt{poll(2)}~\cite{man-poll} replaces bitmasks with an array of \texttt{pollfd} 
(\texttt{fd}, \texttt{events}, \texttt{revents}).

\subsection{Improvements over \texttt{select()}}
\label{subsec:poll-improvements}
\begin{itemize}
    \item No \texttt{FD\char`_SETSIZE} hard limit (array length is yours).
    \item Clearer event flags (\texttt{POLLIN}, \texttt{POLLOUT}, \texttt{POLLERR}, \ldots).
    \item Input \texttt{events} and output \texttt{revents} are separate --- you do not rebuild 
          interest from scratch the same awkward way as \texttt{select}.
\end{itemize}
\begin{lstlisting}[language=C++]
#include <poll.h>

pollfd pfd{};
pfd.fd = sock;
pfd.events = POLLIN | POLLOUT;

int n = poll(&pfd, 1, /* timeout_ms */ 1);
if (n > 0 && (pfd.revents & POLLIN)) {
    // readable
}
\end{lstlisting}

\subsection{Still Linear}
\label{subsec:poll-still-linear}
\texttt{poll} still scales linearly in the number of fds: each wait copies/scans the whole 
array. Better than \texttt{select} for large fd numbers, but not the HFT/server default on Linux.

\section{\texttt{epoll()}}
\label{sec:epoll}
\texttt{epoll}~\cite{man-epoll} is Linux-specific and built for many fds. Interest is registered 
\highlight{once} and stored in the kernel; waits return only \highlight{ready} events.
\newline
Three syscalls:
\begin{itemize}
    \item \texttt{epoll\char`_create1} --- create an epoll instance (returns an epoll fd).
    \item \texttt{epoll\char`_ctl} --- add/mod/del interest for a target fd.
    \item \texttt{epoll\char`_wait} --- block until events (or timeout).
\end{itemize}

\begin{lstlisting}[language=C++]
#include <sys/epoll.h>
#include <unistd.h>

int epfd = epoll_create1(0);
epoll_event ev{};
ev.events = EPOLLIN;
ev.data.fd = sock;
epoll_ctl(epfd, EPOLL_CTL_ADD, sock, &ev);

epoll_event out[64];
int n = epoll_wait(epfd, out, 64, /* timeout_ms */ 1);
for (int i = 0; i < n; ++i) {
    if (out[i].events & EPOLLIN) {
        // handle out[i].data.fd
    }
}
\end{lstlisting}
\texttt{ev.data} is a union (\texttt{fd}, pointer, \texttt{u64}, \ldots) --- store whatever your 
event loop needs to dispatch without another lookup.

\subsection{Level Triggered}
\label{subsec:epoll-level-triggered}
\highlightbf{Default} mode (no \texttt{EPOLLET}). While the condition remains true 
(e.g.\ data still in the receive buffer), \texttt{epoll\char`_wait} keeps reporting the fd.
\newline
Pros: easier to use; missing a byte once does not permanently silence the fd. 
Cons: if you do not drain carefully, you can get woken repeatedly for the same readiness 
(extra wake-ups). Pair with non-blocking I/O and read until \texttt{EAGAIN}.

\subsection{Edge Triggered}
\label{subsec:epoll-edge-triggered}
Add \texttt{EPOLLET}: you are notified on a \highlight{state transition} (e.g.\ went from 
empty to readable), not continuously while ready.
\newline
Rules that interviewers expect:
\begin{itemize}
    \item Fds \highlight{must} be non-blocking.
    \item On notification, \highlight{drain} until \texttt{EAGAIN}/\texttt{EWOULDBLOCK} 
          (read/write loops). Stopping early can leave data buffered with \highlight{no further 
          edge} until more arrives --- classic ET bug.
    \item Often combined with \texttt{EPOLLONESHOT} (re-arm explicitly after handling) in 
          multi-threaded reactors.
\end{itemize}
LT is safer for learning; ET is common in high-performance loops once the drain discipline 
is solid.

\subsection{Epoll Scalability}
\label{subsec:epoll-scalability}
\begin{itemize}
    \item Interest list lives in the kernel --- \texttt{epoll\char`_ctl} is roughly constant 
          time per registration change; not rebuilt every wait.
    \item \texttt{epoll\char`_wait} cost tracks \highlight{ready} fds (plus your userspace 
          handling), not the full interest set size.
    \item One thread can monitor tens of thousands of sockets; the limit becomes your 
          processing, not the wait API.
\end{itemize}
That is why Linux network servers and many trading gateways use \texttt{epoll} (or go further 
with kernel bypass). \texttt{select}/\texttt{poll} remain useful vocabulary and for tiny 
portability niches.

\section{Event Loop Architecture}
\label{sec:event-loop-architecture}
A typical \highlight{reactor} loop:
\begin{enumerate}
    \item Create \texttt{epfd}; register listening and connected sockets (non-blocking).
    \item \texttt{epoll\char`_wait} (timeout 0 for pure poll, small ms, or \texttt{-1} to sleep).
    \item For each ready event: \texttt{accept} new clients, or \texttt{recv}/\texttt{send} 
          on existing ones; on ET, loop until \texttt{EAGAIN}.
    \item Handle \texttt{EPOLLERR} / \texttt{EPOLLHUP}: close and \texttt{EPOLL\char`_CTL\char`_DEL}.
    \item Repeat.
\end{enumerate}
Design notes for low-latency stacks:
\begin{itemize}
    \item Keep the hot path on a \highlight{pinned} core (Chapter~\ref{chap:linuxprocesses}); 
          avoid blocking syscalls inside the loop.
    \item Prefer \highlight{one} epoll thread owning the sockets, or shard by connection with 
          clear ownership --- shared epoll across many threads needs oneshot/re-arm discipline.
    \item Timeout 0 + spinning is busy-poll territory (Chapter~\ref{chap:lowlatency}); a small 
          positive timeout trades a bit of latency for less wasted power when idle.
    \item Still kernel-mediated I/O: packet path goes through the networking stack. Cutting 
          that cost is Chapter~\ref{chap:kernelbypass}.
\end{itemize}

\begin{lstlisting}[language=C++]
// Sketch: single-threaded LT epoll loop (errors omitted)
void run_loop(int epfd)
{
    epoll_event out[64];
    for (;;) {
        const int n = epoll_wait(epfd, out, 64, 1);
        for (int i = 0; i < n; ++i) {
            const int fd = out[i].data.fd;
            if (out[i].events & (EPOLLERR | EPOLLHUP)) {
                epoll_ctl(epfd, EPOLL_CTL_DEL, fd, nullptr);
                close(fd);
                continue;
            }
            if (out[i].events & EPOLLIN) {
                // read until EAGAIN if using EPOLLET; else one or more reads
                handle_readable(fd);
            }
            if (out[i].events & EPOLLOUT) {
                handle_writable(fd);
            }
        }
    }
}
\end{lstlisting}

Interview line: \texttt{select}/\texttt{poll} scan all interest (linear in fds); \texttt{epoll} keeps 
interest in-kernel and returns ready set. Prefer LT until you need ET; with ET always use 
non-blocking fds and drain to \texttt{EAGAIN}.
