\chapter{Memory Model and Synchronization}
\label{chap:memorymodel}

Chapter~\ref{chap:concurrency} lists the C++ APIs (\texttt{std::atomic}, mutexes, 
\texttt{memory\char`_order} enumerators). This chapter explains the \highlight{rules}: when two 
threads may see each other's writes, which orders are enough for a handoff, and how to 
build a small \highlight{lock-free SPSC} queue used on HFT hot paths.

\section{C++ Memory Model}
\label{sec:cpp-memory-model}
The C++ memory model defines which program executions are allowed when threads share memory. 
Compilers and CPUs may \highlight{reorder} loads and stores unless synchronisation forbids it. 
Portable concurrent code relies on atomics, mutexes, or other synchronising operations --- 
not on ``it works on my machine.''

\subsection{Data Race}
\label{subsec:data-race}
A \highlight{data race} occurs when two threads access the \highlight{same memory location}, 
at least one access is a \highlight{write}, and the accesses are \highlight{not ordered} by 
the memory model (no mutex, no atomic with adequate ordering, no other happens-before edge).
\newline
In C++, a data race is \highlightbf{undefined behavior}. Fix it with atomics, locks, or by 
giving each thread private data.

\begin{lstlisting}[language=C++]
#include <thread>

int g = 0;  // shared, non-atomic

void bad()
{
    std::thread a([] { g = 1; });
    std::thread b([] { g = 2; });  // data race with a
    a.join();
    b.join();
}
\end{lstlisting}

\subsection{Race Condition}
\label{subsec:race-condition}
A \highlight{race condition} is a broader engineering term: the program's correctness depends on 
\highlight{timing} or interleaving. It may or may not be a formal data race.
\newline
Example: both threads use a mutex (no data race) but the business logic still breaks if 
updates run in the wrong order. Interview tip: say \highlight{data race} when you mean UB; 
say \highlight{race condition} for logic/timing bugs.

\subsection{Happens-Before}
\label{subsec:happens-before}
\highlight{Happens-before} is the relation that makes writes in one thread \highlight{visible} 
to another. If A happens-before B, then B is guaranteed to see the side effects of A 
(for the relevant memory).
\newline
Typical ways to create happens-before between threads:
\begin{itemize}
    \item Unlock of a mutex happens-before a later lock of the \highlight{same} mutex.
    \item Atomic \texttt{release} store happens-before an atomic \texttt{acquire} load that 
          \highlight{reads the value written} by that store (on the same atomic object).
    \item Thread construction / \texttt{join} edges with the thread's entry and exit.
\end{itemize}
Without a happens-before edge, another thread may see stale or torn values.

\subsection{Sequential Consistency}
\label{subsec:sequential-consistency}
See Chapter~\ref{chap:concurrency}, Section~\ref{subsec:memory-order} for \texttt{memory\char`_order\char`_seq\char`_cst}. 
This subsection covers the formal model only.
\newline
Under \highlight{sequential consistency}, there is a single total order of all seq\_cst 
atomic operations that all threads agree on, and it is consistent with each thread's 
program order. It is the \highlight{easiest mental model} and the default for many atomic 
operations --- but often \highlight{stronger (and costlier)} than a producer/consumer handoff needs.

\section{Memory Orders}
\label{sec:memory-orders-theory}

\subsection{C++ Reference}
\label{subsec:memory-order-reference}
See Chapter~\ref{chap:concurrency}, Section~\ref{subsec:memory-sync-ordering} for API listings and 
Section~\ref{subsec:memory-order} for semantics.

\subsection{Release-Acquire Pairing}
\label{subsec:release-acquire-pairing}
The standard pattern for publishing data from a producer to a consumer:
\begin{enumerate}
    \item Producer writes the payload (non-atomic or relaxed atomics).
    \item Producer performs a \highlight{release} store on a flag / index (``data is ready'').
    \item Consumer performs an \highlight{acquire} load on that flag / index.
    \item If the consumer sees the producer's value, it may safely read the payload.
\end{enumerate}
The release-acquire pair creates happens-before: all writes \highlight{before} the release 
become visible to the consumer \highlight{after} the matching acquire.

\begin{lstlisting}[language=C++]
#include <atomic>
#include <thread>

int payload = 0;
std::atomic<bool> ready{false};

void producer()
{
    payload = 42;                                  // (1)
    ready.store(true, std::memory_order_release);  // (2)
}

void consumer()
{
    while (!ready.load(std::memory_order_acquire)) {  // (3)
        // spin or wait
    }
    // (4) safe to read payload: sees 42
    int x = payload;
}
\end{lstlisting}

\noindent
\texttt{memory\char`_order\char`_acq\char`_rel} is for read-modify-write that both acquire and release 
(e.g.\ some lock implementations). \texttt{memory\char`_order\char`_relaxed} only guarantees 
atomicity of that location --- \highlight{no} synchronisation with other memory.

\subsection{When to Weaken Ordering}
\label{subsec:when-to-weaken-ordering}
\begin{itemize}
    \item \highlightbf{seq\_cst} --- default / when unsure; simplest reasoning; may cost more on weak HW.
    \item \highlightbf{release / acquire} --- message passing, SPSC indices, ``publish flag'' patterns.
    \item \highlightbf{relaxed} --- pure counters, statistics, when no other data depends on the atomic.
    \item Prefer \highlight{correct first}; measure before weakening further on x86 (often strong in practice, 
          but portable code must still use the right orders).
\end{itemize}

\section{Lock-Free Ring Buffer (SPSC)}
\label{sec:spsc-ring-buffer}
A \highlight{single-producer single-consumer} ring buffer moves messages between two threads 
without a mutex. It is a standard HFT building block (feed thread $\rightarrow$ strategy thread).

\subsection{SPSC Model}
\label{subsec:spsc-model}
Exactly \highlight{one} thread pushes, exactly \highlight{one} thread pops. That restriction allows 
a simple design with two indices and no CAS loops for the common path. Multi-producer or 
multi-consumer queues need different algorithms (and are harder to get right).

\subsection{Head and Tail Indices}
\label{subsec:spsc-head-tail}
Use a fixed-capacity array. The producer advances a \highlight{tail} (write position); the consumer 
advances a \highlight{head} (read position). Indices are \texttt{std::atomic} (or indexed modulo capacity).
\newline
Only the producer writes \texttt{tail}; only the consumer writes \texttt{head}. Each thread may 
\highlight{read} the other's index.

\subsection{Producer and Consumer Ordering}
\label{subsec:spsc-producer-consumer-ordering}
\begin{itemize}
    \item Producer: write slot, then \texttt{tail.store(..., release)}.
    \item Consumer: \texttt{head}-side acquire load of \texttt{tail} (or acquire load when observing 
          new data), then read slot.
\end{itemize}
Release-acquire on the indices publishes the slot contents to the consumer.

\subsection{Full and Empty Conditions}
\label{subsec:spsc-full-empty}
With capacity $N$ (often power of two):
\begin{itemize}
    \item \highlight{Empty} when \texttt{head == tail} (consumer has caught up).
    \item \highlight{Full} when \texttt{tail - head == N} (no free slot) --- exact formula depends on 
          whether you leave one slot empty or use an explicit size counter.
\end{itemize}
Never let producer and consumer update the \highlight{same} index.

\subsection{Cache-Line Layout}
\label{subsec:spsc-cache-line-layout}
See Section~\ref{subsec:false-sharing-solutions} for padding and alignment as a general technique. 
This subsection covers SPSC-specific head/tail placement only.
\newline
Put \texttt{head} and \texttt{tail} on \highlight{separate cache lines} (\texttt{alignas(64)}; 
see Section~\ref{subsec:false-sharing-solutions} for the portable constant) so the 
producer's stores to \texttt{tail} do not invalidate the consumer's line for \texttt{head} 
(false sharing).

\subsection{Implementation}
\label{subsec:spsc-implementation}
Sketch (power-of-two capacity, one empty slot to distinguish full/empty):

\begin{lstlisting}[language=C++]
#include <atomic>
#include <cstdint>
#include <vector>

template <typename T, std::size_t N>
class SpscRing {
    static_assert((N & (N - 1)) == 0, "N power of two");
    static_assert(N >= 2);

    alignas(64) std::atomic<std::size_t> head_{0};  // consumer
    alignas(64) std::atomic<std::size_t> tail_{0};  // producer
    T buf_[N];

public:
    bool try_push(const T& v)
    {
        const std::size_t t = tail_.load(std::memory_order_relaxed);
        const std::size_t next = (t + 1) & (N - 1);
        if (next == head_.load(std::memory_order_acquire)) {
            return false;  // full
        }
        buf_[t] = v;
        tail_.store(next, std::memory_order_release);
        return true;
    }

    bool try_pop(T& out)
    {
        const std::size_t h = head_.load(std::memory_order_relaxed);
        if (h == tail_.load(std::memory_order_acquire)) {
            return false;  // empty
        }
        out = buf_[h];
        head_.store((h + 1) & (N - 1), std::memory_order_release);
        return true;
    }
};
\end{lstlisting}

\noindent
Interview talking points: SPSC only; release/acquire on indices; cache-line separation; 
preallocate \texttt{buf\_} (no \texttt{new} on the hot path).

\section{Synchronization}
\label{sec:synchronization-theory}

\subsection{C++ Reference}
\label{subsec:sync-reference}
See Chapter~\ref{chap:concurrency}, Section~\ref{subsec:mutual-exclusion} for 
\texttt{std::mutex}, \texttt{std::shared\char`_mutex}, and related types.

\subsection{Spinlock}
\label{subsec:spinlock}
A \highlight{spinlock} waits by spinning on an atomic flag instead of sleeping in the kernel. 
On a \highlight{dedicated core} with very short critical sections, it can beat a mutex 
(no context switch). On a busy machine, spinning wastes CPU and can hurt latency elsewhere.
\newline
Rough rule: mutex for most code; spinlock only when you control the core and the critical 
section is tiny. See also busy polling (Section~\ref{sec:busy-polling}).

\begin{lstlisting}[language=C++]
#include <atomic>

struct Spinlock {
    std::atomic_flag flag = ATOMIC_FLAG_INIT;

    void lock()
    {
        while (flag.test_and_set(std::memory_order_acquire)) {
            // spin
        }
    }

    void unlock()
    {
        flag.clear(std::memory_order_release);
    }
};
\end{lstlisting}

\section{Performance Considerations}
\label{sec:sync-performance}

\subsection{Lock Contention}
\label{subsec:lock-contention}
\highlight{Contention} means many threads fight for the same lock. Hold times grow, cores wait, 
and caches bounce the lock word. Fixes: finer-grained locks, shard data, or remove sharing 
(SPSC / thread-local).

\subsection{Lock Convoy}
\label{subsec:lock-convoy}
A \highlight{lock convoy} happens when threads repeatedly wake, find the lock taken, and sleep 
again --- often with priority or scheduling effects that serialize progress. Symptom: high 
context-switch rate around one mutex. Prefer shorter critical sections and less sharing.

\subsection{Synchronization Scalability}
\label{subsec:scalability}
Ask: does adding cores increase useful work, or only increase waiting on one lock? 
Designs that scale typically \highlight{partition} work (per-core queues, sharding) instead of 
one global mutex. HFT systems often accept \highlight{single-threaded} hot paths plus SPSC 
handoffs rather than fine-grained locking everywhere.
