\chapter{Memory Mapping and Shared Memory}
\label{chap:memorymapping}

Processes do not share heap pointers across address spaces (Chapter~\ref{chap:linuxprocesses}). 
They still need fast ways to move bytes: classic \texttt{read}/\texttt{write}, or 
\highlight{map} pages into the virtual address space with \texttt{mmap}. 
Mapped regions matter for IPC, large files, and low-copy data paths; page behavior ties to the 
TLB (Section~\ref{sec:tlb}) and NUMA first-touch (Chapter~\ref{chap:numa}).

\section{File I/O}
\label{sec:file-io}
The traditional path copies between kernel buffers and user buffers on every call. 
Fine for configuration and logs; often too heavy for hot market-data or shared state.

\subsection{\texttt{open()}}
\label{subsec:open}
\texttt{open(2)}~\cite{man-open} returns a file descriptor for a path. Flags select access 
(\texttt{O\char`_RDONLY}, \texttt{O\char`_RDWR}, \ldots) and behavior (\texttt{O\char`_CREAT}, 
\texttt{O\char`_CLOEXEC}, \ldots). Always check for \texttt{-1} / \texttt{errno}.

\subsection{\texttt{read()}}
\label{subsec:read}
Copies up to \texttt{count} bytes from the fd into a user buffer. May return fewer bytes than 
requested (partial read). Return 0 means end-of-file for regular files.

\subsection{\texttt{write()}}
\label{subsec:write}
Copies from a user buffer toward the fd. May be partial; loop until all intended bytes are 
queued (same discipline as \texttt{send} in Chapter~\ref{chap:linuxnetworking}).

\subsection{\texttt{close()}}
\label{subsec:close-file-io}
Drops the fd. After the last close, kernel resources for that open file description are released. 
Mapped pages that still reference the file can outlive the fd until \texttt{munmap}.

\subsection{\texttt{lseek()}}
\label{subsec:lseek}
Moves the file offset used by subsequent \texttt{read}/\texttt{write}. Irrelevant for 
\texttt{mmap} access (you index the mapped pointer instead), but still used when mixing 
mapped and unmapped I/O on the same file.

\begin{lstlisting}[language=C++]
#include <fcntl.h>
#include <unistd.h>

int fd = open("feed.bin", O_RDONLY);
if (fd < 0) { /* handle error */ }
char buf[4096];
ssize_t n = read(fd, buf, sizeof(buf));
close(fd);
\end{lstlisting}

\section{Memory Mapping}
\label{sec:memory-mapping}
\texttt{mmap} maps a file (or anonymous memory) into the calling process's virtual address space. 
After success you touch memory with ordinary loads/stores; the kernel fills pages on demand.

\subsection{\texttt{mmap()}}
\label{subsec:mmap}
\texttt{mmap(2)}~\cite{man-mmap}:
\begin{lstlisting}[language=C++]
#include <sys/mman.h>
#include <fcntl.h>
#include <unistd.h>

int fd = open("book.bin", O_RDWR);
void* p = mmap(/* addr */ nullptr,
               /* length */ 4096,
               PROT_READ | PROT_WRITE,
               MAP_SHARED,
               fd,
               /* offset */ 0);
if (p == MAP_FAILED) { /* handle error */ }
// use p as a typed pointer into the file
close(fd);  // mapping can outlive the fd
\end{lstlisting}
Important flags:
\begin{itemize}
    \item \highlightbf{\texttt{PROT\char`_READ} / \texttt{PROT\char`_WRITE} / \texttt{PROT\char`_EXEC}} 
          --- allowed access; mismatches fault.
    \item \highlightbf{\texttt{MAP\char`_SHARED}} --- writes are visible to other mappers of the 
          same object and (for files) can be written back to the file.
    \item \highlightbf{\texttt{MAP\char`_PRIVATE}} --- copy-on-write private view; writes do not 
          update the underlying file or other processes.
    \item \highlightbf{\texttt{MAP\char`_ANONYMOUS}} --- no file; pages backed by RAM/swap 
          (fd is typically \texttt{-1}). Used for large private arenas or, with care, shared 
          anonymous maps via inheritance after \texttt{fork}.
\end{itemize}
Offset and length should be page-aligned where the man page requires it. Prefer 
\texttt{MAP\char`_FAILED} checks --- never compare only to \texttt{nullptr}.

\subsection{\texttt{munmap()}}
\label{subsec:munmap}
\texttt{munmap(2)}~\cite{man-munmap} removes the mapping. Accessing the old pointer afterward is 
undefined (likely \texttt{SIGSEGV}). For file-backed \texttt{MAP\char`_SHARED} data you need on 
disk, \texttt{msync} can flush dirty pages before unmap if durability matters.

\section{Concepts}
\label{sec:memory-mapping-concepts}
See Section~\ref{sec:address-space} for virtual address space and pages.

\subsection{Memory-Mapped Files}
\label{subsec:memory-mapped-files}
A file-backed map lets you treat file contents as a byte array (or struct overlay) in memory. 
Benefits versus repeated \texttt{read}:
\begin{itemize}
    \item Fewer explicit syscalls on the hot path (page faults still enter the kernel when cold).
    \item OS page cache and your process share physical frames for the same file pages.
    \item Natural fit for large read-mostly datasets (reference data, sometimes capture files).
\end{itemize}
Costs: harder error handling (faults instead of \texttt{read} return codes), alignment/struct-layout 
care, and accidental large working sets that thrash the TLB/cache.

\subsection{Shared Memory}
\label{subsec:shared-memory}
To share a region between unrelated processes, map the \highlight{same} object with 
\texttt{MAP\char`_SHARED}:
\begin{itemize}
    \item \highlightbf{POSIX shared memory:} \texttt{shm\char`_open}~\cite{man-shm-open} creates/opens 
          a name under \texttt{/dev/shm}; then \texttt{ftruncate} + \texttt{mmap(..., MAP\char`_SHARED, ...)}.
    \item \highlightbf{File-backed:} both processes \texttt{mmap} the same path with 
          \texttt{MAP\char`_SHARED}.
\end{itemize}
\begin{lstlisting}[language=C++]
#include <sys/mman.h>
#include <sys/stat.h>
#include <fcntl.h>
#include <unistd.h>

int fd = shm_open("/hft_ring", O_CREAT | O_RDWR, 0600);
ftruncate(fd, 4096);
void* p = mmap(nullptr, 4096, PROT_READ | PROT_WRITE, MAP_SHARED, fd, 0);
\end{lstlisting}
Readers and writers still need a \highlight{protocol} (atomics, seqlocks, SPSC rings --- 
Chapter~\ref{chap:memorymodel}). Mapping alone does not make concurrent access safe.
\newline
HFT use: publisher maps a ring or book snapshot region; consumers map read-only or shared and 
poll without socket copies. Pin producers/consumers with NUMA in mind (Chapter~\ref{chap:numa}).

\subsection{Lazy Loading}
\label{subsec:lazy-loading}
\texttt{mmap} usually does \highlight{not} pull the whole file into RAM up front. The kernel 
establishes virtual mappings; physical pages appear when first touched (\highlight{demand paging}). 
That keeps startup cheap and RAM use proportional to the working set --- until you touch 
everything and fault it all in.

\subsection{Page Faults}
\label{subsec:page-faults}
A \highlight{page fault} occurs when the CPU accesses a virtual address that is not currently 
resident (or lacks permission). The kernel may:
\begin{itemize}
    \item Load the page from file/swap (major fault --- slow),
    \item Map an already-cached page (minor fault --- cheaper),
    \item Or deliver \texttt{SIGSEGV} on illegal access.
\end{itemize}
On a latency-critical path, unexpected major faults are jitter. Warm important maps 
(touch pages at startup), size for the working set, and watch TLB pressure 
(Section~\ref{sec:tlb}).

\subsection{Copy-on-Write}
\label{subsec:copy-on-write}
With \texttt{MAP\char`_PRIVATE}, the first write to a shared-looking page allocates a 
\highlight{private} copy for that process; other mappers keep the old contents. 
\texttt{fork} uses the same idea for the child's address space: pages stay shared read-only 
until either side writes.
\newline
Interview line: \texttt{MAP\char`_SHARED} shares updates; \texttt{MAP\char`_PRIVATE} is COW and 
does not update the file. Shared memory still needs synchronization on top of the map.
