\chapter{Algorithms}
\label{chap:algorithms}
\vspace{10.5cm}

\section{Time Complexity}
Time complexity for an algorithm has tree different facets
\begin{itemize}
    \item The \highlightit{worst-case} complexity of the algorithm is 
          the function deﬁned by the maximum number of steps taken in 
          any instance of size n. 
    \item The \highlightit{best-case} complexity of the algorithm is 
          the function deﬁned by the minimum number of steps taken in 
          any instance of size n. 
    \item The \highlightit{average-case} complexity or expected time of 
          the algorithm, which is the function deﬁned by the average 
          number of steps over all instances of size n. 
\end{itemize}
\newpage
\subsection{Big O notation}
\begin{equation}
    \label{eqn:bigOnotation}
    f(n) = O(g(n)) \quad \text{if} \quad 0 \le f(n) \le c g(n) \forall n \ge n_0      
\end{equation}

For all values $n >= n_0$, $f(n)$ is on or below $c g(n)$.
\newline
A constant multiple of $g(n)$ is an \highlightit{asimptotic upper bound} for $f(n)$.

\subsection{Omega notation}
\begin{equation}
    \label{eqn:omeganotation}
    f(n) = \Omega(g(n)) \quad \text{if} \quad 0 \le c g(n) \le f(n) \forall n \ge n_0
\end{equation}

For all values $n>=n_0$, $f(n)$ is on or abow $c g(n)$.
\newline
A constant multiple of $g(n)$ is an \highlightit{asimptotic lower bound} for $f(n)$.

\subsection{Theta notation}
\begin{equation}
    \label{eqn:thetanotation}
    f(n) = \Theta(g(n)) \quad \text{if} \quad 0 \le c_1 g(n) \le f(n) \le c_2 g(n) \forall n \ge n_0     
\end{equation}
 
For all values $n >= n_0$, $f(n)$ is equal to $g(n)$ within a constant factor.
\newline
$g(n)$ is an \highlightit{asimptotically tight bound} for $f(n)$.

\subsection{Meaning of previous notations - Insertion sort case}
\textbf{Big O notation}
\newline
describes an \highlightit{upper bound}.
When we use it to \highlightit{bound the worst-case running time} of an 
algorithm, we have a bound on the running time of the algorithm on \highlightit{every input}.
\newline
Thus, the $O(n^2)$ bound on worst-case running time of \highlightit{insertion sort} 
also applies to its running time on every input.
\newline
\textbf{Omega notation} 
\newline
describes a \highlightit{lower bound}.
When we say that the running time (no modifier) of an algorithm is $\Omega(g(n))$, 
we mean that 
\highlightit{no matter what particular input of size n is chosen} for
each value of n, the running time on that input is at least a constant 
times $g(n)$, for sufficiently large $n$.
\newline
Equivalently, we are giving a \highlightit{lower bound on the best-case running time} 
of an algorithm. For example, the best-case running time of insertion 
sort is $\Omega(n)$, which implies that the running time of insertion sort 
is $\Omega(n)$.
\newline
The \highlightit{running time of insertion sort therefore belongs to both $\Omega(n)$ 
and $O(n^2)$}, since it falls anywhere between a linear function of n and 
a quadratic function of n.
Moreover, these bounds are asymptotically as tight as possible: 
for instance, the running time of insertion sort is not $\Omega(n^2)$, 
since there exists an input for which 
insertion sort runs in $\Theta(n)$ time 
(e.g., when the input is already sorted). 
It is not contradictory, however, 
to say that the worst-case running time of insertion sort is $\Omega(n^2)$, 
since there exists an input that causes the algorithm to take $\Omega(n^2)$ time.
\newline
\textbf{Theta notation}
\newline{}
describes a \highlightit{tight bound}. The $\Theta(n^2)$ notation bound on the worst-case running time of insertion sort, however, 
does not imply a $\Theta(n^2)$ bound 
on the running time of insertion sort on every input. 
For example, when the input is already sorted, 
insertion sort runs in $\Theta(n)$ time.
Technically, 
it is an abuse to say that the running time of insertion sort is $O(n^2)$, 
since for a given n, 
the actual running time varies, depending on the particular input of size n. 
When we say "the running time is $O(n^2)$", 
we mean that 
there is a function $f(n)$ that is $O(n^2)$ 
such that for any value of n, 
no matter what particular input of size n is chosen, 
the running time on that input is bounded from above by the value $f(n)$.
Equivalently, we mean that the worst-case running time is $O(n^2)$.

\newpage
\subsection{Running times of well known algorithms}

\begin{table}[h]
    \begin{tabularx}{\textwidth}{|l|l|l|l|X|}
    %~ \begin{tabular}{|l|l|l|l|X|}
        \hline
        Algorithm & Best & Average & Worst & Worst Space Complexity  \\
        \hline
        Quicksort & $O(n\log{n})$ & $O(n\log{n})$ & $O(n^{2})$&$O(\log{n})$ \\
        \hline
        Mergesort & $O(n\log{n})$ & $O(n\log{n})$&$O(n\log{n})$ &$O(n)$ \\
        \hline
        Timsort & $O(n)$  &$O(n\log{n})$& $O(n\log{n})$& $O(n)$\\
        \hline
        Heapsort & $O(n\log{n})$& $O(n\log{n})$& $O(n\log{n})$& $O(1)$\\
        \hline
        Bubble Sort & $O(n)$ &$O(n^{2})$ & $O(n^{2})$&$O(1)$\\
        \hline
        Insertion Sort & $O(n)$ &$O(n^{2})$ & $O(n^{2})$&$O(1)$\\
        \hline
        Selection Sort & $O(n^{2})$ & $O(n^{2})$& $O(n^{2})$& $O(1)$  \\
        \hline
        Shell Sort & $O(n)$ & $O((n\log{n})^{2})$ & $O((n\log{n})^{2})$& $O(1)$\\
        \hline
        Bucket Sort & $O(n+k)$ & $O(n+k)$ & $O(n^{2})$ & $O(n)$\\
        \hline
        Radix Sort & $O(nk)$ & $O(nk)$ & $O(nk)$ & $O(n+k)$\\
        \hline    
    %~ \end{tabular}
    \end{tabularx}
\end{table}

Verbal description of the previous sorting algorithms:
\begin{itemize}

    \item \href{https://en.wikipedia.org/wiki/Quicksort}{\highlight{Quicksort}} \\
    Efficient, general-purpose sorting algorithm, \highlight{slightly faster} 
    than \highlight{merge sort} and \highlight{heapsort} for \highlight{randomized data}, 
    particularly on larger distributions. \\
    It selects a \highlight{pivot element} from the array and partitions 
    the other elements into \highlight{two sub-arrays}, according to whether 
    they are \highlight{less than or greater than the pivot}. For this reason, 
    it is sometimes called \highlight{partition-exchange sort}.
    \newline
    The sub-arrays are then \highlight{sorted recursively}.
    
    \item \href{https://en.wikipedia.org/wiki/Merge_sort}{\highlight{Mergesort}} \\
    First, divide the list into the smallest unit (1 element), 
    then \highlight{compare each element} \highlight{with the adjacent} 
    list to sort and merge the two adjacent lists.
    \newline 
    Finally, all the elements are sorted and merged.
        
    \item \href{https://en.wikipedia.org/wiki/Timsort}{\highlight{Timsort}} \\
    a hybrid, stable sorting algorithm, derived from merge sort 
    and insertion sort, designed to perform well on many kinds of 
    real-world data. It was implemented by Tim Peters in 2002 for 
    use in the Python programming language. The algorithm finds 
    subsequences of the data that are already ordered (runs) and uses 
    them to sort the remainder more efficiently. This is done by 
    merging runs until certain criteria are fulfilled. Timsort was 
    Python's standard sorting algorithm from version 2.3 to version 3.10
    
    \item \href{https://en.wikipedia.org/wiki/Heapsort}{\highlight{Heapsort}} \\
    comparison-based sorting algorithm which can be thought of as 
    "an implementation of selection sort using the right data structure."[3] 
    Like selection sort, heapsort divides its input into a sorted and an 
    unsorted region, and it iteratively shrinks the unsorted region by 
    extracting the largest element from it and inserting it into the 
    sorted region. Unlike selection sort, heapsort does not waste time 
    with a linear-time scan of the unsorted region; rather, heap sort 
    maintains the unsorted region in a heap data structure to efficiently 
    find the largest element in each step.
    
    \item \href{https://en.wikipedia.org/wiki/Bubble_sort}{\highlight{Bubble Sort}} \\
    a simple sorting algorithm that repeatedly steps through the input 
    list element by element, comparing the current element with the one 
    after it, swapping their values if needed. These passes through the 
    list are repeated until no swaps have to be performed during a pass, 
    meaning that the list has become fully sorted
    
    \item \href{https://en.wikipedia.org/wiki/Insertion_sort}{\highlight{Insertion Sort}} \\
    iterates, consuming one input element each repetition, and grows a 
    sorted output list. At each iteration, insertion sort removes one 
    element from the input data, finds the location it belongs within 
    the sorted list, and inserts it there. It repeats until no input 
    elements remain. 
    
    
    \item \href{https://en.wikipedia.org/wiki/Selection_sort}{\highlight{Selection Sort}} \\
    repeatedly identiﬁes the smallest remaining unsorted element 
    and puts it at the end of the sorted portion of the array
    
    \item \href{https://en.wikipedia.org/wiki/Shellsort}{\highlight{Shell Sort}} \\
    
    
    \item \href{https://en.wikipedia.org/wiki/Bucket_sort}{\highlight{Bucket Sort}} \\
    Each bucket is then sorted individually, either using a different 
    sorting algorithm, or by recursively applying the bucket sorting 
    algorithm. It is a distribution sort, a generalization of pigeonhole 
    sort that allows multiple keys per bucket, and is a cousin of radix 
    sort in the most-to-least significant digit flavor. Bucket sort can 
    be implemented with comparisons and therefore can also be considered 
    a comparison sort algorithm. The computational complexity depends on 
    the algorithm used to sort each bucket, the number of buckets to use, 
    and whether the input is uniformly distributed. 
    
    \item \href{https://en.wikipedia.org/wiki/Radix_sort}{\highlight{Radix Sort}} \\

\end{itemize}

\subsection{Methods for solving the recurrence}
There are three methods for solving recurrences:
\begin{itemize}
    \item the substitution method
    \item the recursion-tree method
    \item the master theorem
\end{itemize}


\section{Space Complexity}
