\chapter{C++}
\label{chap:c++}
\section{C++: General Knowledge and Good Practices}
\label{sec:c++gen-knowledge}
\subsection{Rule of three/five/zero}
%~ References in bibliography:~\cite{fernandes-ruleofzero}, 
%~ \cite{meyers-concern-ruleof0}, \cite{cppcoreguidelines-C67}, \cite{cppcoreguidelines-C21}.

\subsubsection{Rule of three}
\label{subsubsec:c++ruleof3}
If a class requires a \highlightit{user-defined destructor}, a user-defined 
\highlightit{copy constructor}, or a user-defined 
\highlightit{copy assignment operator}, it almost certainly requires all 
three.

Because C++ copies and copy-assigns objects of user-defined types in 
various situations (passing/returning by value, manipulating a 
container, etc), these special member functions will be called, 
if accessible, and if they are not user-defined, they are 
implicitly-defined by the compiler.

The implicitly-defined special member functions are typically incorrect 
if the class manages a resource whose handle is an object of non-class 
type (raw pointer, POSIX file descriptor, etc), whose destructor does 
nothing and copy constructor/assignment operator performs a 
"shallow copy" (copy the value of the handle, without duplicating the 
underlying resource).

\begin{lstlisting}[language=C++]
class Buffer {
    char* data_;
    std::size_t size_;
public:
    explicit Buffer(std::size_t n);
    ~Buffer();                          // free data_
    Buffer(const Buffer& other);        // deep copy
    Buffer& operator=(const Buffer& other); // deep copy / self-assign safe
};
\end{lstlisting}

Classes that manage non-copyable resources through copyable handles 
may have to declare copy assignment and copy constructor private 
and not provide their definitions or define them as deleted. 
This is another application of the rule of three: deleting one 
and leaving the other to be implicitly-defined will most likely result 
in errors.

\subsubsection{Rule of five}
\label{subsubsec:c++ruleof5}
Because the presence of a user-defined destructor, copy-constructor, 
or copy-assignment operator prevents the implicit definition of the 
move constructor and the move assignment operator, \highlightit{any class 
for which move semantics are desirable}, has to declare all five special 
member functions.

\begin{lstlisting}[language=C++]
class Buffer {
    char* data_;
    std::size_t size_;
public:
    explicit Buffer(std::size_t n);
    ~Buffer();
    Buffer(const Buffer&);
    Buffer& operator=(const Buffer&);
    Buffer(Buffer&&) noexcept;            // steal pointer
    Buffer& operator=(Buffer&&) noexcept; // steal pointer
};
\end{lstlisting}

Unlike Rule of Three, 
failing to provide move constructor and move assignment 
is usually not an error, 
but a missed optimization opportunity.

\subsubsection{Rule of zero}
\label{subsubsec:c++ruleof0}
Classes that have custom 
destructors, copy/move constructors or copy/move assignment operators 
should deal exclusively with ownership 
(which follows from the Single Responsibility Principle). 
Other classes should not have custom 
destructors, copy/move constructors or copy/move assignment operators.
(see~\cite{fernandes-ruleofzero})

This rule also appears 
in the C++ Core Guidelines as C.20: 
If you can avoid defining default operations, do.

When a \highlightit{base class is intended for polymorphic use}, 
its destructor may have to be declared public and virtual. 
This blocks implicit moves (and deprecates implicit copies), 
and so the special member functions have to be declared as defaulted
(see~\cite{meyers-concern-ruleof0})

However, this makes the class prone to slicing, 
which is why polymorphic classes often 
define copy as deleted 
(see~\cite{cppcoreguidelines-C67}), 
which leads to the following generic wording for the Rule of Five:
If you define or =delete any default operation, 
define or =delete them all
(see~\cite{cppcoreguidelines-C21})

\subsection{The Most Vexing Parse}
\label{subsubsec:most-vex-parse}
\toclessnparagraph{Examples}
The term \highlight{"most vexing parse"} was first used by Scott Meyers 
in 2001. While unusual in C, the phenomenon was quite common in C++ until 
the introduction of uniform initialization in C++11.
\newline
This issue is about ambiguous code line like
\begin{lstlisting}[language=C++]
void f(double my_dbl)
{
  int i(int(my_dbl));
}
\end{lstlisting}
in which the second line can be interpreted both as \highlight{a declaration 
of a variable} \texttt{i} whose initial value is obtained by converting 
\texttt{my\char`_dbl} into an \texttt{int} and as \highlight{a function declaration} 
equivalent to
\begin{lstlisting}[]
int i(int my_dbl);
\end{lstlisting} 
A similar ambiguity arises for 
\begin{lstlisting}[language=C++]
struct Foo {};

struct FooKeeper {
  explicit FooKeeper(Foo f);
  int get_foo();
};

int main() {
  FooKeeper foo_keeper(Foo());
  return foo_keeper.get_foo();
}
\end{lstlisting}
in which the line
\begin{lstlisting}[language=C++]
TimeKeeper time_keeper(Timer());    
\end{lstlisting}
can be interpreted both as a variable definition and a function 
declaration. The C++ standard requires the second interpretation, which 
is not consistent with the previous code.
\toclessnparagraph{How to fix it}
It is possible to solve this parsing issue in different ways.
\newline
In the type conversion example
\begin{itemize}
    \item C-style cast
    \begin{lstlisting}[language=C++]
    int i((int)my_dbl);
    \end{lstlisting}
    
    \item C++ named cast (to prefer to the previous solution)
    \begin{lstlisting}[language=C++]
    int i(static_cast<int>(my_dbl));
    \end{lstlisting} 
\end{itemize}
In the variable declaration example, since C++11 the preferred method is 
to use uniform brace initialization
\begin{lstlisting}[language=C++]
//Any of the following work:
FooKeeper foo_keeper(Foo{});
FooKeeper foo_keeper{Foo()};
FooKeeper foo_keeper{Foo{}};
FooKeeper foo_keeper(     {});
FooKeeper foo_keeper{     {}};
\end{lstlisting}

Prior to C++11, extra parenthesis or copy-initialization solves the 
most vexing parse issue
\begin{lstlisting}[language=C++]
FooKeeper foo_keeper( /*Avoid MVP*/ (Foo()) );
FooKeeper foo_keeper = FooKeeper(Foo());
\end{lstlisting}

\subsection{General Topics}
\subsubsection{\href{https://isocpp.org/wiki/faq/intrinsic-types}{Built-in / Intrinsic / Primitive Data Types}}
A \highlight{POD type} is a C++ type that has an equivalent in C, and 
that uses the same rules as C uses for initialization, copying, layout, 
and addressing.

The actual definition of a POD type is recursive and gets a little gnarly. 
\newline
Here’s a slightly simplified definition of POD: 
\begin{itemize}
    \item a \highlight{POD type’s non-static data members} must be 
          \highlight{public} and can be of any of these types: 
          \highlighttt{bool}, \highlight{any numeric type} including the 
          various char variants, 
          \newline
          \highlight{any enumeration type}, \highlight{any data-pointer type} 
          (that is, any type convertible to \texttt{void*}), 
          \highlight{any pointer-to-function type}, or 
          \highlight{any POD type}, including \highlight{arrays of any of these}. 
    \item \highlight{data-pointers and pointers-to-function} are okay, 
          but \highlight{pointers-to-member} are not. 
    \item \highlight{references} are not allowed
    \item a POD type can’t have \highlight{constructors}, \highlight{virtual functions}, 
          \highlight{base classes}, or an 
          \newline
          \highlight{overloaded assignment operator}.
\end{itemize}

When \highlight{initializing non-static data members} of built-in / intrinsic / 
primitive types, it is always better to initialize all non-static data 
members in the constructor's \highlight{initialization list}.
\newline
In order to nitialize your built-in / intrinsic / primitive variable by 
an expression that the compiler doesn’t evaluate solely at compile-time, 
concern should be put on the initalization order fiasco, explained 
in~\ref{subsubsec:constructors}.
\newline
A function should return
\begin{itemize}
    \item \highlighttt{void} if there is no need for a return value;
    \item \highlight{local by value}
    \item \highlight{local by pointer}
    \item \highlight{data member by value}
    \item \highlight{data member by pointer}
    \item \highlight{data member by reference-to-nonconst}
    \item \highlight{data member by referenec-to-const}
    \item \highlighttt{shared\char`_ptr}
    \item local \highlighttt{unique\char`_ptr} or \highlighttt{shared\char`_ptr}
    \item \highlight{other}
    \item
\end{itemize}

\subsubsection{\href{https://isocpp.org/wiki/faq/input-output}{Input/Output via \texttt{<iostream>} and \texttt{<cstdio>}}}
There are several reasons why \texttt{<iostream>} should be preferred to 
\texttt{<cstdio>}
\begin{itemize}
    \item \highlight{type-safeness} is better guaranteed since \texttt{<iostream>} 
          lets the compiler know the I/O object statically, while \texttt{<cstdio>}
    \item \highlight{less error prone}
    \item \highlight{extensibility}
    \item \highlight{inheritance}
\end{itemize}

\textbf{check for invalid input}
\begin{lstlisting}[language=C++]
#include <iostream>
int main()
{
  std::cout << "Enter a number, or -1 to quit: ";
  int i = 0;
  while (std::cin >> i) {    // GOOD FORM
    if (i == -1) break;
    std::cout << "You entered " << i << '\n';
  }
  // ...
}
\end{lstlisting}

\textbf{skip invalid input}
\begin{lstlisting}[language=C++]
// ...
while ((std::cout << "How old are you? ")
     && (!(std::cin >> age) || age < 1 || age > 200)) {
std::cout << "That's not a number between 1 and 200; ";
std::cin.clear();
std::cin.ignore(std::numeric_limits<std::streamsize>::max(), '\n');
}
// ...
\end{lstlisting}

\textbf{Functioning of \texttt{(std::cin >> foo)}}

\textbf{stop processing past the end of file}

\textbf{Ending output lines with \texttt{std::endl} of \texttt{'\textbackslash n'}}

\textbf{Providing printing for an entire hierarchy of classes}

\textbf{\texttt{printOn() method vs. \texttt{friend} function}}

\textbf{providing input for a class}

\textbf{reopen \texttt{std::cin} and \texttt{std::cout} in binary mode}

\textbf{writing/reading objects to/from a data file}

\textbf{sending objects to another computer}

\textbf{printing a \texttt{char} as a number}

\subsubsection{\href{https://isocpp.org/wiki/faq/const-correctness}{Const Correctness}}
\label{subsubsec:c++const-corr}
\href{https://isocpp.org/wiki/faq/const-correctness}{ISOCpp FAQ about \highlight{\texttt{const} correctness}}.
\newline
Const correctness consists in using the keyword \highlighttt{const} to 
prevent \texttt{const} objects from being mutated.

\subsubsection{\texttt{const} and pointer combinations}
Reading it right-to-left
\begin{table}[h!t]
    \begin{tabularx}{\textwidth}{|l|X|}
        \hline
        \texttt{const X* p} & p is a pointer to an X that is constant.\\
        \hline
        \texttt{X const* p} & equivalent to \texttt{const X* p}.\\
        \hline
        \texttt{X* const p} & p is a \texttt{const} pointer to an X that 
                              is not \texttt{const}.\\
        \hline
        \texttt{const X* const p} &  p is a \texttt{const} pointer to an 
                                     X that is \texttt{const}.\\
        \hline
    \end{tabularx}
\end{table}

The second combination is equivalent to the first one: the 
\highlight{\texttt{const}-on-the-right style} requires to write the 
\texttt{const} specifier on the right of the object it makes 
\texttt{const}. This style is also called \highlight{consistent \texttt{const}} 
since it is consistent with the way of defining a const member function. 
Moreover, it is also more consistent with pointer declarations.
\subsubsection{const and reference combinations}
Reading it right-to-left
\begin{table}[h!t]
    \begin{tabularx}{\textwidth}{|l|X|}
        \hline
        \texttt{const X\& r} & r aliases an X object, which can not be changed via r \\
        \hline
        \texttt{X const\& r} & equivalent to \texttt{const X\& r} \\
        \hline
        \texttt{X\& const r} & nonsense functionally equivalent to \texttt{X\& r}\\
        \hline
        \texttt{const X\& const r} & nonsense functionally equivalent to \texttt{const X\& r}\\       
        \hline
    \end{tabularx}
\end{table}

The equivalence between the second and the first combination is due 
to the same reason valid for the pointers.
\newline
The last two combinations are nonsense and should be avoided. They are 
redundant as a reference is always constant, in the sense that you can 
never reseat a reference to make it refer to a different object.

\subsubsection{Logical and Physical State}
\label{subsubsec:logphystate}
When reasoning in terms of const-correctness, it's important to 
distinguish the logical and the physical state of an object.
\newline
On the inside, an object has a \highlightbf{physical state}, also called 
\highlightit{concrete} or \highlightit{bitwise} state. This state is the one 
visible to the programmer who implements the software. From the outside, 
the client has an access to the object restricted to public member 
functions and friend functions. What is visible is the \highlightbf{logical} 
or \highlightit{abstract} or \highlightit{meaningwise state}.
\newline
A lookup functionality for a collection-object is a good example to 
show how there could be bits in the object's physical state that haev 
no corresponding elements in the object's logical state. The lookup 
functionality might improve the performance using a cache to store 
the already looked-up items. The implementation of the cache is part of 
the object's physical state but the implementation is internal and 
the details will probably not be exposed to the users, i.e. it is not 
part of the object's logical state.

\subsubsection{\texttt{const}ness of \texttt{public} member functions}
The constness of a public method must make sense to the object's users, 
and those users can see only the object's logical state. Therefore, in 
this context the \texttt{const}ness to take care of is the logical one. 
\begin{itemize}    
    \item If a method changes any part of the object’s logical state, 
          it logically is a mutator; it should not be const even if 
          (as actually happens!) the method doesn’t change any physical 
          bits of the object’s concrete state.
    \item Conversely, a method is logically an inspector and should be 
          const if it never changes any part of the object’s logical 
          state, even if (as actually happens!) the method changes 
          physical bits of the object’s concrete state.
\end{itemize}

\subsubsection{return-by-reference and \texttt{const} member function}
A \texttt{const} member function must not change nor allow its caller to 
change the logical state of the \texttt{this} object it's been called 
on.
\newline
If an inspector method needs to return a data member of a \texttt{this} 
object, it should return by \texttt{const reference} or by copy. 
For example
\begin{lstlisting}[language=C++]
class Person
{
    public:
        const std::string& get_name() const;  // good one
        std::string& get_lastname() const;    // wrong one
        int get_age();                        // good one
}
\end{lstlisting}

Compilers won't always catch such a mistake, therefore it is 
responsibility of the programmer.

\subsubsection{\texttt{const} overloading}
\texttt{const} overloading is a way to achieve \texttt{const} 
correctness when there are an inspector and mutator method with the 
same name. One of the most common case is due to  the subscript 
operators, which usually come in pair as in the following
\begin{lstlisting}[language=C++]
class PersonList
{
    public:
        // Subscript operators often come in pairs
        const Person& operator[] (unsigned index) const;  
        Person&       operator[] (unsigned index);        
        // ...
};
\end{lstlisting}
Of course \texttt{const}-overloading can be used for other things than 
subscript operator.
\newline
Following some links to the \texttt{isocpp FAQs}
\begin{itemize}
    \item \href{https://isocpp.org/wiki/faq/operator-overloading#matrix-subscript-op}{Subscript operator for Matric class}
    \item \href{https://isocpp.org/wiki/faq/freestore-mgmt#multidim-arrays2}{Matrix class made simpler}
    \item \href{https://isocpp.org/wiki/faq/freestore-mgmt#multidim-arrays3}{Matrix using templates}
    \item \href{https://isocpp.org/wiki/faq/freestore-mgmt#multidim-arrays4}{Matrix using templates and \texttt{vector}s}
    \item \href{https://isocpp.org/wiki/faq/templates#class-templates}{Class Template's syntax and semantics}
\end{itemize}

\subsubsection{\texttt{mutable} keyword}
\texttt{mutable} keyword helps achieving the \texttt{const} correctness, 
when the code needs to make changes to a \texttt{const} object's 
physical state. An example could be again the cached previous lookup
in~\ref{subsubsec:logphystate}. In this case, the cache would be marked as 
\texttt{mutable} and the compiler would let the programmer change it 
even in case of a \texttt{const} object because it would know that 
a change of that attribute would reflect in a change in the physical 
state, leaving the logical state as it was before.
\newline
Using \texttt{const} is better than removing the constness with a 
\texttt{const\char`_cast}

\subsubsection{Error converting \quad \texttt{F** $\rightarrow$ const F**}}
\label{subsubsec:error-converting}
C++ allows the conversion \texttt{F** $\rightarrow$ F const*} but gives 
an error if a program tries to implicitly convert 
\texttt{F** $\rightarrow$ const F**}.
\newline
The reason why this conversion is not legal in C++ is that it would let 
silently and accidentally modify a \texttt{const Foo} object without a 
cast.
\begin{lstlisting}[language=C++]
class Foo
{
    public:
    void modify();  // make some modification to the this object
};

int main()
{
    const Foo x;
    Foo* p;
    const Foo** q = &p;  // q now points to p;
                         // this is (fortunately!) an error
    *q = &x;             // p now points to x
    p->modify();         // Ouch: modifies a const Foo!!
    // ...
}
\end{lstlisting}
In case the third line of the main were legal, \texttt{q} would be 
pointing at \texttt{p} and the next line it would change \texttt{p} to 
point at \texttt{x}. This would made the \texttt{const} qualifier 
disappear. The last line would then use \texttt{p} ability to modify 
its referent and the program would end up modifying a 
\texttt{const Foo}.
\newline
In case of necessity, the solution is to change \texttt{const Foo**} to 
\texttt{const Foo* const*}.

\subsubsection{\href{https://isocpp.org/wiki/faq/references}{References}}
A \highlight{reference} is an alias for an object. At the assembly 
implementation level, a reference \texttt{i} to an object \texttt{x} is 
typically the machine address of the object \texttt{x}. When the 
programmer writes an instruction like \texttt{i++}, the compiler generates 
the code to increment the value of the referenced object \texttt{x}.
\newline
Assigning to a reference changes the state of the referenced object.
\newline
In case a function returns a reference, the function call can appear 
on the left hand of an assignment operator:
\begin{lstlisting}[language=C++]
class Array {
public:
  int size() const;
  float& operator[] (int index);
  // ...
};
int main()
{
  Array a;
  for (int i = 0; i < a.size(); ++i)
    a[i] = 7;    // This line invokes Array::operator[](int)
  // ...
}
\end{lstlisting}
Method chaining like \highlighttt{object.method1().method2()}, in which 
the returned object from \texttt{method1()} is the \texttt{this} object 
for \texttt{method2()} are possible because of return by reference.
\newline
The method chaining is used in the \highlight{Named Parameter Idiom}, 
described in~\ref{subsec:named-parameter-idiom} (see also~\cite{isocpp-named-parameter-idiom}).

\subsubsection{\href{https://isocpp.org/wiki/faq/freestore-mgmt}{Memory Management}}
The best tecniques to prevent memory leakes rely on hiding allocation and 
deallocation inside more manageable types. Single objects can be managed 
through \texttt{make\char`_unique} and \texttt{make\char`_shared}, multiple 
objects can be managed through the container \texttt{vector} or 
\texttt{unordered\char`_map}. Their memory management frees the programmer 
from explicit memory management, macros, casts, overflow checks, 
explicit size limits, pointers and similia. 
\newline
Templates and the standard libraries make this use of containers, resource 
handles, etc., much easier than it was even a few years ago. The use of 
exceptions makes it close to essential.
\newline
In the following code, a \texttt{unique\char`_ptr} is used as a resource 
handle in order not to cope with allocation/deallocation responsibilities 
when returning an object allocated on the freestore from a function
\begin{lstlisting}[language=C++]
#include<memory>
#include<iostream>
using namespace std;
struct S {
  S() { cout << "make an S\n"; }
  ~S() { cout << "destroy an S\n"; }
  S(const S&) { cout << "copy initialize an S\n"; }
  S& operator=(const S&) { cout << "copy assign an S\n"; }
};
S* f()
{
  return new S;   // who is responsible for deleting this S?
};
unique_ptr<S> g()
{
  return make_unique<S>();    // explicitly transfer responsibility for deleting this S
}
int main()
{
  cout << "start main\n";
  S* p = f();
  cout << "after f() before g()\n";
//S* q = g(); // this error would be caught by the compiler
  unique_ptr<S> q = g();
  cout << "exit main\n";
//leaks *p
//implicitly deletes *q
}    
\end{lstlisting}
In case containers are not available, it is possible to use a memory leak 
detector as part of your standard development procedure, or plug in a 
garbage collector.
\newline
In C++11 a literal \highlighttt{nullptr} has been introduced and should 
be used as the null pointer value in place of the previous solutions 
(namely \texttt{NULL} and \texttt{0}).
\newline
The function of \highlighttt{delete} is deleting the data pointed by its 
operand. It must be used not more than once on the same pointer allocated 
with a \highlighttt{new}, assuming that pointer has not gone through a 
successive \texttt{new} assignment.
\newline
The \highlighttt{malloc()/free()} operations can not be mixed with 
\highlighttt{new/delete} operations. The reasons are that they do not 
operate on the same level: they allocate on \highlight{different heaps} 
and \highlight{the former work with raw memory} while 
\highlight{the latter work with constructed objects} whose proper 
construction and distruction are guaranteed. 
\newline
More specifically, \texttt{malloc()} is a function that takes a number 
(of bytes) as its argument; it returns a \texttt{void*} pointing to 
unitialized storage and this pointer needs to be converted to a proper 
type. \texttt{new} is an operator that takes a type and 
(optionally) a set of initializers for that type as its arguments; it 
returns a pointer to an (optionally) initialized object of its type.
\newline
\texttt{malloc()} reports memory exhaustion by returning \texttt{0}. 
\texttt{new} reports allocation and initialization errors by throwing 
exceptions (\texttt{bad\char`_alloc}).
\newline
Another important reason to use \texttt{new} and \texttt{delete} is their 
\highlight{overridability}.
\newline
In case there is a need to use \texttt{realloc()}, it is better to use 
\texttt{vector} containers. When \texttt{realloc()} has to copy the 
allocation, it uses a bitwise copy operation, which will tear many C++ 
objects to shreds. C++ objects should be allowed to copy themselves. 
\newline
C++ provides the programmer with 
\href{https://en.cppreference.com/w/cpp/memory/new/set_new_handler}{\texttt{set\char`_new\char`_handler}}, 
a function called by allocation functions whenever a memory allocation 
attempt fails.
\newline
In C++ there is not the need to check whether a pointer is \texttt{null} 
before \texttt{delete}-ing it. The process goes through two steps
\begin{lstlisting}[language=C++]
// Original code: delete p;
if (p) {    // or "if (p != nullptr)"
  p->~Fred();
  operator delete(p);
}   
\end{lstlisting}
in which \texttt{operator delete(p)} calls the memory deallocation primitive, 
\texttt{void operator delete(void* p)}
C++ does not guarantee that \texttt{delete} also null out its operand. 
One reason is that the operand of \texttt{delete} is not required to be 
an \texttt{lvalue}.
\newline
In case it is important to null out the pointer, a possible implementation 
can be
\begin{lstlisting}[language=C++]
template<class T> inline void destroy(T*& p) { delete p; p = 0; }
\end{lstlisting}
Passing a reference has also the effect to prevent the a call of this 
function template on an \texttt{rvalue}.
\newline
For what concerns the call to the destructor at the end of the scope, 
heap allocation should be used only if an object needs to live beyond the 
lifetime of the scope you create it in. Even then, \texttt{make\char`_unique} 
or \texttt{make\char`_shared}. In rare cases in which heap allocation is 
really needed and \texttt{new} is used to create an object, \texttt{delete} 
must be used to destroy the object, taking into account that code is 
no longer exception-safe.
\newline
In case an object must live in a scope only, heap allocation is not 
needed at all.
\newline
In case an array of objects is created with \texttt{new}, the \texttt{delete[]}
operator must be used to destroy the array. There are several ways a 
compiler can keep memory of the number of elements to destroy\\
https://isocpp.org/wiki/faq/compiler-dependencies\#num-elems-in-new-array-overalloc \\
https://isocpp.org/wiki/faq/compiler-dependencies\#num-elems-in-new-array-assocarray
\newline
In case of interest, there are several code example to
\begin{itemize}
    \item \href{https://isocpp.org/wiki/faq/freestore-mgmt#multidim-arrays}
               {allocating multidimensional arrays using \texttt{new}}
    
    \item \href{https://isocpp.org/wiki/faq/freestore-mgmt#multidim-arrays2}
               {allocating multidimensional arrays encapsulating pointers 
               in a class}
               
    \item \href{https://isocpp.org/wiki/faq/freestore-mgmt#multidim-arrays3}
               {reworking the previous example in terms of templates}
    
    \item \href{https://isocpp.org/wiki/faq/freestore-mgmt#multidim-arrays4}
               {building a multidimensiona array using \texttt{std::vector}}
    
\end{itemize}

In case a member function needs to \texttt{delete this} following 
requirements must be satisfied
\begin{itemize}
    \item \texttt{this} object was allocated via new (not by \texttt{new[]}, 
          nor by placement new, nor a local object on the stack, 
          nor a namespace-scope / global, nor a member of another 
          object)
    
    \item member function will be the last member function invoked on 
          this object.
    
    \item the rest of your member function (after the delete this line) 
          doesn’t touch any piece of this object (including calling any 
          other member functions or touching any data members). This 
          includes code that will run in destructors for any objects 
          allocated on the stack that are still alive.
    
    \item no one even touches the \texttt{this} pointer itself after the 
          \texttt{delete} this line. In other words, you must not examine 
          it, compare it with another pointer, compare it with \texttt{nullptr}, 
          print it, cast it, do anything with it.
    
\end{itemize}

It's also possible to use the \highlight{Named Constructor Idiom} 
(see~\ref{subsec:named-ctor-idiom} and~\cite{isocpp-named-ctor-idiom}) 
to force a class to be created via \texttt{new} rather than as local, 
namespace-scope, global or \texttt{static}
\begin{lstlisting}[language=C++]
class Fred {
public:
  // The create() methods are the "named constructors":
  static Fred* create()                 { return new Fred();     }
  static Fred* create(int i)            { return new Fred(i);    }
  static Fred* create(const Fred& fred) { return new Fred(fred); }
  // ...
private:
  // The constructors themselves are private or protected:
  Fred();
  Fred(int i);
  Fred(const Fred& fred);
  // ...
};
\end{lstlisting}
In case \texttt{Fred} class is expected to have derived classes, 
the constructors must be in 
\newline
\texttt{protected} section of the class.


\subsection{Classes and Inheritance}

\subsubsection{\href{https://isocpp.org/wiki/faq/classes-and-objects}{\textbf{Classes and Objects}}}
A \highlight{class} is the building block of Object Oriented programming. 
It defines a data type, that consists of a set of states and a set of 
operations which transition between those states. A class provides a set 
of (usually public) operations, and a set of (usually non-public) data 
bits representing the abstract values that instances of the type can have.
\newline
An object is a region of storage with associated semantics defined in the 
class it belongs to.
\newline
An interface is good when
\begin{itemize}
    \item unnecessary details are intentionally hidden. 
          \newline
          This reduces the user’s defect-rate.
    \item users don’t need to learn a new set of words and concepts. 
          \newline
          This reduces the user’s learning curve.
\end{itemize}
Encapsulation is the mechanism used to prevent unauthorized access to 
some piece of information or functionality of a class/object.
\newline
The key money-saving insight is to separate the volatile part of some 
chunk of software from the stable part.
\newline
The \highlight{volatile parts} are the implementation details. In case 
of a single class this part is encapsulated in the \highlight{private/protected} 
part of the class. Inheritance can also be used as a form of encapsulation.
\newline
The \highlight{stable parts} are the interfaces. Good ones provide 
simplified view in the vocabulary of users, and are designed from the 
outside-in. !!! PUT LINK TO OPERATOR OVERLOADING !!! 
\newline
If there is a single class, the interface is simply the class's \texttt{public} 
member functions and \texttt{friend} functions. The \texttt{private} 
and/or \texttt{protected} parts of a class contain the class’s implementation, 
which is typically where the data lives.
\newline
The clean interface separates the interface from its implementation, 
and encapsulation forces users to use the interface.
\newline
Encapsulation does not prevent a user to read the private and/or protected 
parts of a class, but does prevent the code from being dependent on the 
insides of the class.
\newline
A class method can directly access the non-\texttt{public} members of 
another instance of its class. This rule of C++ preserves the encapsulation. 
Otherwise there would be the need to provide a \texttt{public get\char`_method} 
for each \texttt{private} variable used by the assignment operator, 
the copy constructor, the comparison operators, the binary arithmetic 
and bitwise operators, and many others.
\newline
Methods and friends of a derived class can access the \texttt{protected} 
base class members of any of its own objects, but not others. Suppose 
the classes \texttt{D1} and \texttt{D2} inherit from class \texttt{B}, 
that has a \texttt{private x} member. In this scenario, the compiler 
let \texttt{D1}'s members and friend dirrectly access the \texttt{x} 
member of any \texttt{D1}. The compiler gives a compile-time error 
if \texttt{D1} member or friend tries to directly access the \texttt{x} 
member of anything it does not know is at least a \texttt{D1}, such as 
via a \texttt{B*} pointer, a \texttt{B\&} reference, a \texttt{B} object, 
a \texttt{D2*} pointer, a \texttt{D2\&} reference, a \texttt{D2} object.
\newline
The members and base classes of a struct are public by default, while in 
class, they default to private. For the other aspects are functionally 
equivalent.
\newline
To define an in-class constant useable in a compile time constant expression, 
there are three choices
\begin{lstlisting}[language=C++]
class X {
    constexpr int c1 = 42; // preferred since C++11
    static const int c2 = 7;
    enum { c3 = 19 };
    array<char,c1> v1;
    array<char,c2> v2;
    array<char,c3> v3;
    // ...
};    
\end{lstlisting}
In case the contant isn't needed in a compile time constant expression
\begin{lstlisting}[language=C++]
class Z {
    static char* p;     // initialize in definition
    const int i;        // initialize in constructor
public:
    Z(int ii) :i(ii) { }
};
char* Z::p = "hello, there";    
\end{lstlisting}
It is possible to bind a reference to a static data member (to take its 
address), if and only if it has an out-of-the-class definition
\begin{lstlisting}[language=C++]
class AE {
    // ...
public:
    static const int c6 = 7;
    static const int c7 = 31;
};
const int AE::c7;   // definition
    void byref(const int&);
int f()
{
            byref(AE::c6);                  // error: c6 not an lvalue
            byref(AE::c7);                  // ok
    const int* p1 = &AE::c6;    // error: c6 not an lvalue
    const int* p2 = &AE::c7;    // ok
    // ...
}
\end{lstlisting}

Like C, C++ doesn’t define layouts, just semantic constraints that must 
be met.
\newline
The size of an empty class isn't zero to ensure that two different objects 
will have different addresses. For the same reason, \texttt{new} always 
return pointers to different objects.
\newline
The Empty Base Optimization (see~\cite{cppreference-ebo}) is a mandatory 
requirement on class layout since C++11, and requires that the size of any 
object or member subobject is required to be at least 1 even if the type 
is an empty class type'

\subsubsection{\href{https://isocpp.org/wiki/faq/ctors}{\textbf{Constructors}}}
\label{subsubsec:constructors}

\subsubsection{\href{https://isocpp.org/wiki/faq/dtors}{\textbf{Destructors}}}
\label{subsubsec:destructors}
Destructors are used to realease any resource allocated by the object, 
like a semaphore, a file and so on. A very common case is 
\texttt{delete}-ing the object created using \texttt{new}.
\newline
Destructors are invocated on a local scope in reverse order of 
construction in the code.
\newline
The objects stored in an array are destructed in reverse order of 
construction as well.
\newline
The sub-objects are destructed in reverse order of construction as well. 
For example, having this structure
\begin{lstlisting}[language=C++]
struct derived: base0, base1
{
  member0 m0_;
  member1 m1_;
  ~derived()
  {
    local0 l0;
    local1 l1;
  }
}
\end{lstlisting}
when an \texttt{derived} object goes out of scope, destructors are called 
in this order: \texttt{~local1()}, \texttt{~local0()}, \texttt{~member1()}, 
\texttt{~member0()}, \texttt{~base1()}, \texttt{~base0()}
\newline
Destructors can not be overloaded.
\newline
Destructors need not to be called on local variables. In case it is needed 
to destroy a local object in a specific point of a local scope, it is 
possible to use an artificial block \texttt{\{\dots\}}. In case it's not 
possible to wrap the object with an artificial block, the solution is 
to create a function with similar duties.
\newline
If the object to destroy was allocated via \texttt{new}, a call to the 
destructor isn't needed unless it was a placement \texttt{new}
\begin{lstlisting}[language=C++]
void someCode()
{
  char memory[sizeof(Fred)];
  void* p = memory;
  Fred* f = new(p) Fred();
  // ...
  f->~Fred();   // Explicitly call the destructor for the placed object
}    
\end{lstlisting}
\href{https://isocpp.org/wiki/faq/dtors#placement-delete}{\textbf{Is there a placement delete?}}
\newline
From a class destructor, it is not needed to explicitly call destructor 
of neither member objects nor base classes.
\newline
\href{https://isocpp.org/wiki/faq/dtors#memory-pools}{\textbf{Is there a way to force new to allocate memory from a specific memory area?}}
\newline

\subsubsection{\href{https://isocpp.org/wiki/faq/assignment-operators}{\textbf{Assignment Operator}}}

\subsubsection{\href{https://isocpp.org/wiki/faq/operator-overloading}{\textbf{Operator Overloading}}}

\subsubsection{\href{https://isocpp.org/wiki/faq/friends}{\textbf{Friends}}}
\label{subsubsec:friend}
\href{https://en.cppreference.com/w/cpp/language/friend}{Cppreference documentation about \textbf{Friends}}
\newline
A \highlighttt{friend} can be a \highlighttt{function}, a \highlighttt{class}, 
and both their \highlighttt{template} version.
\newline
The friend declaration appears in a class body and grants a function or 
another class access to \highlighttt{private} and \highlighttt{protected} 
members of the class where the friend declaration appears. 
\newline
\highlighttt{friend} declaration helps enforcing encapsulation. Indeed, 
granting access to \texttt{private} and \texttt{protected} members 
removes the burden of creating public \highlighttt{getters} and 
\highlighttt{setters}, since in most of the cases these methods destroy 
encapsulation, exposing private and protected details which make no 
sense outside of the class.
\newline
The disadvantage of \texttt{friend} functions is that they require 
extra code when dynamic binding is needed. Indeed, the \texttt{friend} 
function should call a hidden (usually \texttt{protected}) \texttt{virtual} 
member function, as explained in details in~\ref{subsec:virtual-friend-function}.
\newline
Friendship \highlightit{is not inherited, transitive or reciprocal}:
\begin{itemize}
    \item a derived class of a friend class is not a friend class;
    \item a friend of a friend is not necessarily a friend;
    \item if \texttt{class A}\, declares \texttt{class B}\, its friend, 
    \texttt{B} objects have access to \texttt{A} objects' private data, 
    but the opposite is not true. 
\end{itemize}

Usually, it is better to rely on member function if possible and \texttt{friend} 
ones when it is really needed.

\subsubsection{\href{https://isocpp.org/wiki/faq/basics-of-inheritance}{\textbf{Inheritance - Basics}}}
\label{subsubsec:inheritance-basics}

\toclessnparagraph{\texttt{static} initialization}

\subsubsection{\href{https://isocpp.org/wiki/faq/virtual-functions}{Inheritance - \textbf{\texttt{virtual} member functions}}}.
\newline
A \href{https://en.cppreference.com/w/cpp/language/virtual}{virtual member function} 
is the way of C++ to allow derived classes to replace the implementation 
provided by the base class. The compiler makes sure the correct 
derived class' implementation is called, even when the object is 
accessed through a pointer to the base class.
\newline
The reason why the member function are not \texttt{virtual} by default 
is that many classes are not designed to be used as a base class. 
Moreover, the \texttt{virtual} function call mechanism needs requires 
an overhead of memory, typically one word per object.

\toclessnparagraph{Dynamic Binding and Static Typing}
A pointer to an object may actually point to a class that is derived 
from the class of the pointer. Thus there are two different types 
to consider: the (static) type of the pointer and the (dynamic) type 
of the actual pointed object.
\newline
\highlightbf{Static typing} means that the legality of a member function 
invocation is checked as early as possible, i.e. at compile time by the 
compiler itself. In this kind of check, the compiler uses the 
\highlightit{static type} of the pointer to determine whether the member 
function invocation is correct, meaning that the type pointed by the 
pointer can handle the member function.
\newline
\highlightbf{Dynamic binding} is the \texttt{virtual} function call 
mechanism, which checks the correctness of the function call at the last 
possible moment, i.e. at run time and based on the dynamic type of 
the object pointed to.

\toclessnparagraph{Member Function call: \texttt{virtual} vs. non-\texttt{virtual}}
The practical difference is that non-\texttt{virtual} member functinos 
are resolved statically, i.e. the compiler checks the 
\highlightit{type of the pointer or reference to the object} and selects 
the correct member function statically at compile time.
\newline
On the other side, a \texttt{virtual} member function call is resolved 
dynamically.

\toclessnparagraph{\texttt{virtual} Member Function Call - Overhead Estimation}
As explained, a non-\texttt{virtual} member function is resolved 
statically based on the type of the pointer or reference to the object.
\newline
In contrast, \highlightbf{dynamic binding} select the correct 
\texttt{virtual} member function at run time following a compiler-based 
variant of the following technique: for each virtual function, the 
compiler create in each object an hidden \highlightit{v-pointer} 
(virtual pointer) pointing to a global table called \highlightit{v-table} 
(virtual table). One single \highlightit{v-table} is created for each class
which contains at least a \highlighttt{virtual} function.
\newline
To resolve a call to a \highlightit{virtual} function, the run-time procedure 
follows the object's \highlightit{v-pointer} to the class's \highlightit{v-table}, 
and then follows the appropriate slot in the \highlightit{v-table} to the 
method code.
\newline
The \highlightbf{space-cost} overhead consists of an extra pointer per object 
needing to be dynamically binded and another extra pointer per 
\texttt{virtual} method. The \highlightbf{time-space} overhead compared to 
a normal function call consits of the two extra fetches to get the 
value of the \highlightit{v-pointer} and a second one to get the address of  
the method.
\newline
Suppose to have a \highlighttt{Base} class with $5$ \highlighttt{virtual} 
functions
\begin{lstlisting}[language=C++]
// Original C++ source code
class Base
{
    public:
        virtual return_type virt0( /*...arbitrary params...*/ );
        virtual return_type virt1( /*...arbitrary params...*/ );
        virtual return_type virt2( /*...arbitrary params...*/ );
        virtual return_type virt3( /*...arbitrary params...*/ );
        virtual return_type virt4( /*...arbitrary params...*/ );
        // ...
};
\end{lstlisting}

There are three steps the compiler goes through

\nsubparagraph{First step: building a static table}
Compiler build a static table containing 5 function-pointers, most 
likely while compiling the \texttt{.cpp} that define the \texttt{Base}'s 
first non-\texttt{inline} \texttt{virtual} function. 
\newline
Let's call this table \highlighttt{\textbf{Base::\char`_\char`_vtable}}. 
If a function pointer fits into one machine word, the \highlightit{v-table} 
adds $5$ hidden words of memory overall.
\begin{lstlisting}[language=C++]
// FunctionPtr is a generic pointer to a generic member function
FunctionPtr Base::__vtable[5] = {
    &Base::virt0, &Base::virt1, &Base::virt2, 
    &Base::virt3, &Base::virt4 };
\end{lstlisting}

\nsubparagraph{Second step: adding an hidden pointer}
An hidden pointer called \highlightit{v-pointer} is added by the compiler to 
\highlight{each object of \texttt{Base}} class as it was an hidden data member.

\begin{lstlisting}[language=C++]
// Original C++ source code
class Base 
{
    public:
    // ...
        FunctionPtr* __vptr;  // Supplied by the compiler, 
                              // hidden from the programmer
    // ...
};
\end{lstlisting}

\nsubparagraph{Third step: initializing \texttt{this->\char`_\char`_vptr}}
Compiler initializes \highlighttt{this->\char`_\char`_vptr} within each 
constructor in order to let each object's \highlighttt{v-pointer} to point 
at its class \highlighttt{v-table}. This step could be thought of as if 
the following instruction is added in each constructor's 
\highlightit{init list}.
\begin{lstlisting}[language=C++]
Base::Base( /*...arbitrary params...*/ )
  : __vptr(&Base::__vtable[0])  // Supplied by the compiler, 
                                // hidden from the programmer
  // ...
{
  // ...
}
\end{lstlisting}

When the code defines a \highlighttt{Der} class which inherits from the 
\highlighttt{Base} class, the compiler repeats the first and third steps, 
but \highlightit{skips the second step}.
\newline
In the \highlightbf{first step}, compiler creates another hidden 
\textit{v-table}, in which the compiler replaces the virtual methods 
overridden in the derived class. For example, in case \highlighttt{{Der}} 
overrides \highlight{\texttt{virt0()} \ldots \texttt{virt2()}} the \texttt{Der}'s 
\highlightit{v-table} looks like
\begin{lstlisting}[language=C++]
// Pseudo-code (not C++, not C)
// for a static table defined within file Der.cpp
FunctionPtr Der::__vtable[5] = {
    &Der::virt0, &Der::virt1, &Der::virt2,
    &Base::virt3, &Base::virt4 };
\end{lstlisting}

In the \highlightbf{third step}, the compilers for each \highlighttt{Der} 
object changes the \highlighttt{v-pointer} in order to make it point to its 
class \highlighttt{v-table}.

At this point, it is possible to call a virtual function with code like
\begin{lstlisting}[language=C++]
void mycode(Base* p)
{
    p->virt3();
}
\end{lstlisting}
The compiler rewrites the code like
\begin{lstlisting}[language=C++]
void mycode(Base* p)
{
    p->__vptr[3](p);
}
\end{lstlisting}
At this point
\begin{itemize}
    \item a \highlightit{first load} gets the \texttt{v-pointer} and stores it in a 
          register which we call \texttt{r1}
    \item a \highlightit{second load} gets the work at \texttt{r1 + 3*4} (if 
          function-pointers are 4-bytes long) and stores it in a second 
          register \texttt{r2}
    \item \highlightit{compiler calls} the code at location \texttt{r2}
\end{itemize}

Therefore, it is possible to state that:
\begin{itemize}
    \item a \highlightit{minimal space-overhead} is needed to manage \texttt{virtual} 
          functions
    \item \texttt{virtual} function calls are almost \highlightit{as fast as} 
          non-\texttt{virtual} function calls
    \item there is \highlightit{no chaining effect}, i.e. the fetch-fetch-call 
          mechanism is not dependent on how deep the inheritance gets.
\end{itemize}

\toclessnparagraph{Calling a Base Class \texttt{virtual} member function}
When a \texttt{virtual} function is called by its \highlightit{fully-qualified name}, 
the compiler skip the \texttt{virtual} call mechanism and uses the same 
mechanism used for non-\texttt{virtua} functions. Therefore, the 
fully-qualified \texttt{virtual} function is called by name rather than 
by slot-number.
\newline
To call the \texttt{virtual} function defined in the \texttt{Base} 
within the derived class, it is possible to write
\begin{lstlisting}[language=C++]
void Der::f()
{
    Base::f();
}
\end{lstlisting}

\toclessnparagraph{\texttt{virtual} destructor}
A \texttt{virtual} destructor is required any time someone 
\texttt{delete}s a derived-class object through a base-class pointer. 
In this case, indeed, the destructor invoked matches the type of the 
pointed object ignoring the type of the pointer itself.
\newline
In general, a \texttt{virtual} destructor is needed
\begin{itemize}
    \item if the class is meant to be \highlightit{derived from}
    \item \textit{and} if code uses \highlighttt{new Derived}
    \item \textit{and} if code uses \highlighttt{delete p} in which 
          \texttt{p} is a pointer to \texttt{Base} pointing to an actual 
          object of \texttt{Derived} class.
\end{itemize}
To be shorter, destructor needs to be \texttt{virtual} any time there 
is a \texttt{virtual} method in the class. Indeed
\begin{itemize}
    \item it's a general safety measure in case the class will be used 
    as a base class;
    \item it costs nothing since the 
\end{itemize}

\toclessnparagraph{\texttt{virtual} constructor}
\label{npar:virtual-ctor}
An idiom that allows to achieve a behavior that C++ does not really 
support.
\newline
\href{https://isocpp.org/wiki/faq/virtual-functions#virtual-ctors}{virtual-ctors} \\
\href{https://isocpp.org/wiki/faq/abcs#copy-of-abc-via-clone}{copy-of-abc-via-clone}



\subsubsection{\href{https://isocpp.org/wiki/faq/proper-inheritance}{\textbf{Inheritance - Proper Inheritance and Substitutability}}}
\label{subsubsec:proper-inheritance}

\toclessnparagraph{Converting \texttt{Derived** $\rightarrow$ Base**}}
Differently from the conversion \texttt{Derived* $\rightarrow$ Base*}, 
the conversion mentioned in the title does not compile, because it 
would be dangerous.
\newline
Even though a \texttt{Derived} object is a kind of 
\texttt{Base} object, in case the conversion in the title was possible 
the \highlighttt{Base**} could be dereferenced and it could be possible 
to make \highlighttt{Base*} point to an object of a 
\highlighttt{different derived class}.
\newline
This prohibition has similarities with~\ref{subsubsec:error-converting}.

\toclessnparagraph{container of \highlighttt{Thing} \texttt{==} kind-of container of \highlighttt{Everything}?}
As a consequence of~\ref{subsubsec:proper-inheritance}, a container 
of a certain \highlighttt{class Thing} objects is not a container of 
\highlighttt{class Anything} objects, even if \highlighttt{Thing} is a 
kind of \highlighttt{Anything}.

\toclessnparagraph{Is an array of \texttt{Derived} a kind-of array of \texttt{Base}?}
In the same perspective of the previous two paragraphs, the answer is negative 
and easy to explain: a \texttt{Derived} object is larger than a 
\texttt{Base} object and it would mess up the pointer aritmetics in case 
it was possible to store both the kinds of objects in the same array. 


\subsubsection{\href{https://isocpp.org/wiki/faq/abcs}{\textbf{Inheritance - Abstract Base Classes}}}
At the design level, an \highlighttt{ABC} corresponds to an abstract 
concept. At the programming language level, an \highlighttt{ABC} is a 
class containing one or more pure \highlighttt{virtual} functions, as 
in the following example
\begin{lstlisting}[language=C++]
class Shape
{
public:
    virtual void draw() const = 0;  // = 0 means "pure virtual"
    // ...
};
\end{lstlisting}
Even though it is possible to provide a definition of a pure 
\highlighttt{virtual} function, it could be confusing.

\toclessnparagraph{Copy Constructor and Assignment Operator for a class containing a pointer\\to a (abstract) base class}
If a class owns \ref{npar:virtual-ctor}


\subsubsection{\href{https://isocpp.org/wiki/faq/strange-inheritance}{\textbf{Inheritance - What your mother never told you}}}

\toclessnparagraph{\texttt{final} classes and \texttt{virtual} methods}

\toclessnparagraph{Calling \texttt{virtual} function from non-\texttt{virtual} functions of base class}


\subsubsection{\href{https://isocpp.org/wiki/faq/private-inheritance}{\textbf{Inheritance - \texttt{private} and \texttt{protected} inheritance}}}
Private and protected inheritance are expressed using respectively the 
keyword \highlighttt{private} and \highlighttt{protected} in place of 
\highlighttt{public} in the definition of the derived class.
\newline
\highlighttt{private} inheritance is a syntactic variant of the concepts 
of composition or aggregation. For example, the "\highlighttt{Car} 
\textbf{has-a} \highlighttt{Engine}" relationship can be implemented 
both using single composition and \highlighttt{private} inheritance. 
For the latter, an example is
\begin{lstlisting}[language=C++]
class Car : private Engine
{
    public:
    Car() : Engine(8) {}  // in case Engine has a ctor 
                          // which takes an int
    using Engine::start;
}
\end{lstlisting} 

\highlighttt{private} inheritance has both similarities and differences 
when compared to composition.
Some of the similarities are
\begin{itemize}
    \item a single \highlighttt{Engine} is present in both cases;
    \item in neither case a conversion \highlighttt{Car* $\rightarrow$ Engine*} 
    can be done by users;
    \item \highlighttt{Car} has a \highlighttt{start()} method that 
    calls \highlighttt{Engine}'s \highlighttt{start} method.
\end{itemize}

The differences are
\begin{itemize}
    \item simple-composition is needed in case multiple \highlighttt{Engine} 
    are contained in a \highlighttt{Car};
    \item \highlighttt{private} inheritance can lead to unnecessary 
    highlightbf{multiple inheritance}
    \item \highlighttt{private} inheritance allows member of \highlighttt{Car} 
    to convert \highlighttt{Car* $\rightarrow$ Engine*}
    \item \highlighttt{private} inheritance allows access to 
    \highlighttt{protected} members of the base class
    \item \highlighttt{private} inheritance allows \highlighttt{Car} to 
    override \highlighttt{Engine}'s \highlighttt{virtual} functions
    \item \highlighttt{private} inheritance make it simpler to implement 
    a \highlighttt{Car}'s \highlighttt{start()} method that calls the 
    \highlighttt{Engine}'s \highlighttt{start()} method.
\end{itemize}

\toclessnparagraph{Which should I prefer: composition or private inheritance?}
Composition should be generally preferred to \highlighttt{private} inheritance. 
The latter gives access to the internals and there could be many class 
to expose. This aspect could make the program harder to maintain and 
easy to break when something in the code changes. 
\newline
A scenario in which \highlighttt{private} inheritance is \textit{preferred} is 
when a \highlighttt{class Foo} uses code in \highlighttt{Class Bar} 
and the code in the latter class invokes member functions in the former 
class. The solution is to let \highlighttt{class Foo} call 
non-\texttt{virtual} in \highlighttt{class Bar}, and \highlighttt{class Bar} 
calls its own (pure) \highlighttt{virtual}s, which are overridden by 
\highlighttt{class Foo}.

\toclessnparagraph{Should I pointer-cast from a private derived class to its base class?}
It is generally not a good practice.
\newline
The conversion \highlighttt{PrivatelyDer* $\rightarrow$ Base*} (and 
likewise for the references) is safe and no cast is needed or 
recommended. However, users of \highlighttt{PrivatlyDer} should avoid 
this conversion, since a private decision is subject to change without 
notice.

\toclessnparagraph{How is \texttt{protected} inheritance related to \texttt{private} inheritance?}
Similarly, \highlighttt{protected} inheritance allows to override 
\highlighttt{virtual} functions in the base class and does not claim the 
derived is a kind-of-base class.
\newline
Dissimilarly, \highlighttt{protected} inheritance allows derived classes 
of derived classes to know about the inheritance relationship. The 
\texttt{benefit} is the capability of the 
\highlightit{protected derived class} to exploit the relationship to 
the base class. The \textit{cost} is the fact that changes the relationship 
in the protected derived class can potentially break further derived 
classes.

\toclessnparagraph{Access rules with \texttt{private} and \texttt{protected} inheritance}
Considering the following code
\begin{lstlisting}[language=C++]
class B                    { /*...*/ };
class D_priv : private   B { /*...*/ };
class D_prot : protected B { /*...*/ };
class D_publ : public    B { /*...*/ };
class UserClass            { B b; /*...*/ };
\end{lstlisting}
\begin{itemize}
    \item none of the derived classes can access anything \texttt{private} in \texttt{B};
    
    \item in \highlighttt{D\char`_priv}, the \highlighttt{public} and 
    \highlighttt{protected} of \highlighttt{B} parts are \highlighttt{private}
    
    \item in \highlighttt{D\char`_prot} the \texttt{public} and 
    \texttt{protected} parts of \highlighttt{B} are protected;
    
    \item in \highlighttt{D\char`_publ} the \highlighttt{public} parts 
    of \highlighttt{B} are public and the \highlighttt{protected} parts 
    are protected (\highlighttt{D\char`_publ} is-a-kind-of-a \highlighttt{B});
    
    \item in \highlighttt{class UserClass} can accesso only the 
    \highlighttt{public} parts of \highlighttt{B}.
\end{itemize}
The correct way to make a \highlighttt{public} member 
\highlighttt{f(int, float)} of \highlighttt{B} 
public in \highlighttt{D\char`_prot} or in \highlighttt{D\char`_priv} 
makes use of the \highlightbf{scope operator} as the in the following 
code
\begin{lstlisting}[language=C++]
class D_prot : protected B
{
public:
    using B::f;
}
\end{lstlisting}


\subsubsection{\href{https://isocpp.org/wiki/faq/multiple-inheritance}{\textbf{Inheritance - Multiple and Virtual Inheritance}}}
\toclessnparagraph{Do we really need multiple inheritance?}
Every modern language with \texttt{static} type checking and inheritance 
provides multiple inheritance. In C++, an \textit{ABC} serves usually as 
an interface and class can need many. Other language simply have a 
separate name for pure abstract classes interfaces. Using multiple 
inheritance is not really common but it is better than using 
workarounds.

\toclessnparagraph{Disciplines for using multiple inheritance}
\begin{itemize}
    \item Inheritance should be used if it removes 
    \highlighttt{if}/\highlighttt{switch} statements. Indeed, while 
    composition is meant for code reuse, inheritance's primary purpose 
    is dynamic binding and the flexibility which comes with it.
    \item \highlightit{pure ABC}s play a crucial role in the context of 
    \highlighttt{multiple inheritance}, since classes with as little 
    data as possible and almost all pure \texttt{virtual} methods help 
    avoiding inheritance for code reuse and using it for interface 
    sustainability' 
    \item The \textbf{bridge pattern} or \textbf{nested generalization} 
    are possible alternatives and a wise designer checks out all the 
    ways before choosing the best. 
\end{itemize}

\toclessnparagraph{Example for the guidelines}
\label{parag:mi-guideline-example}
Imagine a context in which there are multiple variants of a similar 
concepts, for example land, water, air and space vehicles, and different 
power sources, for example gas, wind, nuclear power and so on.
\newline
Before tieing up everything using multiple inheritance, there are two 
aspects to take into account choosing a good strategy:
\begin{itemize}
    \item users need to call methods on a \highlighttt{Vehicle}-reference 
    and expect the implementation of those methods to be specific to 
    a concrete vehicle, like \highlighttt{LandVehicle};
    
    \item users need a \highlighttt{Vehicle}-reference that refers to a 
    concrete vehicle, for example \highlighttt{GasPoweredVehicle}, and 
    want to call methods on that \highlighttt{Vehicle}-reference 
    expecting the implementation to get overridden by 
    \highlighttt{GasPoweredVehicle}.
    
\end{itemize}
In case both the conditions stand, multiple inheritance could be the 
correct choice. Better alternatives could still be available depending 
on a few more decision criteria.

In case there are $\textbf{N}$ geographies (land, water, air, space, e.c.) 
and $\textbf{M}$ power sources, there are at least 3 options.

\begin{itemize}
    \item \textbf{Bridge Pattern} \newline
    \highlighttt{Vehicle} and \highlighttt{Engine} are two pure 
    \highlighttt{ABC}s, concrete vehicles and engines are derived from. 
    \highlighttt{Vehicle} interface has an \highlighttt{Engine*} and 
    everything is matched at run-time.
    \newline
    This way the code grows \textit{gracefully}, because there are only 
    $\textbf{N+M}$ derived classes to implement. There are several 
    desadvantages as well. A first one is due to having at most 
    $\textbf{N+M}$ overrides and therefore the same number of concrete 
    algorithms and adding more could be very difficult.
    \newline
    \textbf{Bridge Pattern} is not able to solve nonsensical choices, 
    as a pedal powered space vehicles and to solve this issue extra 
    checks are needed as well.
    \newline
    Another issue is the absence of an interface for all the gas powered 
    vehicles, which are considered just a kind of \highlighttt{Vehicle}. 
    On the other side, all gas powered vehicles share their code in 
    derived class \highlighttt{GasPoweredEngine}.
    
    \item \textbf{Nested Generalization} \newline
    One of the hierarchies is choosen as primary, and suppose to choose 
    the geography as primary in the current example.
    \newline
    Concrete classes as \highlighttt{LandVehicle} and  
    \highlighttt{WaterVehicle} are derived from \highlighttt{Vehicle} 
    and each would have further derived classes, one per power source. 
    For example, \highlighttt{LandVehicle} would be the base class for 
    \highlighttt{GasPoweredLandVehicle}, \\
    \highlighttt{NuclearPoweredVehicle} and so on.
    \newline
    This way, the codebase does not grow gracefully because $\textbf{NxM}$ 
    different derived classes need to be implemented, the advantage is 
    the possibility to have the same number of different algorithms.
    \newline 
    This approach provides with a fine granular control removing the 
    nonsensical combination because the combinations are decided in the 
    design and should be reasonable.
    \newline
    Analougusly to \textit{Bridge Pattern}, there is not a common base 
    class for the specific vehicles, and finally it is not obvious how 
    to share code between different vehicles with same power source.

    \item \textbf{Multiple Inheritance} \newline
    There are two different hierarchies as in the \textit{Bridge Pattern}, 
    the \highlighttt{Engine*} is no longer a data member of the vehicle 
    classes though, and in its place roughly $\textbf{NxM}$ derived classes 
    are implemented below both the hierarchy of the geographies and the 
    hierarchy of egines.
    \newline
    In this context, the concept of \texttt{Engine} changes and vehicle 
    classes should be renamed, e.g. \highlighttt{GasPoweredEngine} would 
    be renamed to \highlighttt{GasPoweredVehicle}. Moreover,
    \highlighttt{GasPoweredLandVehicle} multiply inherits from 
    \highlighttt{GasPoweredVehicle} and \highlighttt{LandVehicle}. At 
    the cost of not having a graceful growth, the gain is a fine 
    granular control simply because nonsesincal combinations like 
    \highlighttt{PedalPoweredSpaceVehicle} are not created for the user. 
    \newline 
    In this model, it is also possible for any gas powered vehicle to 
    share a common interface. Unlike the \textit{Bridge Pattern}, the 
    derived class can override the shared code to replace parts that 
    are not ideal for the new class.
\end{itemize}

\toclessnparagraph{Visualization of tradeoffs}
Given $\textbf{N}$ and $\textbf{M}$, the tradoffs of the approaches 
described in~\ref{parag:mi-guideline-example} 
are summarized in table~\ref{tabular:tradeoffs}
\begin{table}[h!t]
    \begin{tabular}{|l|c|c|c|c|c|c|}
    \hline
    \multirow{2}{*}{}  & \textbf{Grow} & \textbf{Low Code} &\textbf{Fine-Grained} &\textbf{Static Detect} &\textbf{Polimorphic } &\textbf{Share} \\
                         & \textbf{Gracefully} & \textbf{Bulk} &\textbf{Control} &\textbf{Bad Combos} &\textbf{on both sides} &\textbf{Code} \\
    \hline 
    Bridge & \cellcolor{codegreen} & \cellcolor{codegreen}& & & & \cellcolor{codegreen}\\
    \hline                  
    Nested Gen. & & & \cellcolor{codegreen} & \cellcolor{codegreen}  & & \\
    \hline          
    Multiple In.  & & & \cellcolor{codegreen} & \cellcolor{codegreen} & \cellcolor{codegreen} & \cellcolor{codegreen} \\                  
    \hline
    \end{tabular}
    \caption{Bridge Pattern, Nested Generalization and Multiple Inheritance's tradeoffs.}
    \label{tabular:tradeoffs}
\end{table}

The meaning of the colums are the following
\begin{itemize}
    \item \textbf{Grow Gracefully} \newline
    For the example of the previous paragraph, the growth is due to 
    adding more geographies or power sources and in this case it is 
    linear.
    
    \item \textbf{Low Code Bulk} \newline
    

    \item \textbf{Fine Grained Control} \newline
    It takes into account the option of having a different algorithm 
    for any of the different $\textbf{NxM}$ possibilities or being stuck 
    with a single solution for all the concrete classes of vehicles.

    \item \textbf{Static Detect Bad Combos} \newline
    

    \item \textbf{Polymorphic on Both Sides} \newline
    

    \item \textbf{Share Common Code} \newline
    
\end{itemize}




\toclessnparagraph{Second Example for the guidlines}
This example is more symmetric than the previous one, resulting in 
being better managed in using a multiple-inheritance solution.
\newline
Suppose to start with two categories of vehicles, for example 
\highlighttt{LandVehicle} and \highlighttt{WaterVehicle}, and later 
an \highlighttt{AmphibiousVehicle} is needed as well.
\newline
Given this coditions
\begin{itemize}
    \item it is really needed to have an amphibious vehicle, instead of 
    using the other classes with a bit of changes;
    
    \item users need a \highlighttt{LandVehicle} reference that refers 
    to an \highlighttt{AmphibiousVehicle} and need to call methods on 
    the reference expecting the actual implementation is specific 
    to the referenced object;
    
    \item same as the previous point, with \highlighttt{WaterVehicle}.

\end{itemize}
in this case the \textit{multiple inheritance} would be the best choice. 
Answering the question seen in~\ref{parag:mi-guideline-example} could 
help being more confident with this choice.

\newpage
\toclessnparagraph{The Dreaded Diamond}
\noindent{}
\begin{wrapfigure}{L}{0.4\textwidth}% adapt widths of minipages to your needs
\centering{}
    \begin{tikzpicture}%[>=triangle 45,font=\sffamily]
        \node (X) at (0,1.5) [rectangle,draw] {\textbf{Base}};
        \node (Y) at (-1.5,0) [rectangle,draw] {\textbf{Der1}};
        \node (Z) at (1.5,0) [rectangle,draw] {\textbf{Der2}};
        \node (U) at (0,-1.5) [rectangle,draw] {\textbf{Join}};
        \draw [semithick,dashed] (X) -- (Y);
        \draw [semithick,dashed] (X) -- (Z);
        \draw [semithick,dashed] (Y) -- (U);
        \draw [semithick,dashed] (Z) -- (U);
    \end{tikzpicture}
    \caption{A Dreaded Diamond.}
    \label{fig:dreaded-diamond}
\end{wrapfigure}%

A \textit{dreaded diamond} describes a class structure in which a 
particular class appears more than once in an inheritance hierarcy 
and looks like the diagram in figure~\ref{fig:dreaded-diamond}.
\newline
Actually, C++ provides techniques to deal with the dreads of this 
triangle and this situation is not really dreaded but just something to 
be aware of.
\newline
The issue is due to \highlighttt{Base} being inherited 
twice with the consequence that any data member declared in 
\texttt{Base} appears twice within a \highlighttt{Join} object.
\newline
This duality creates ambiguities: when \highlighttt{Join} 
change a \highlighttt{Base}'s data member, it is not clear which derived 
class to go through for the change; likewise, a conversion 
\highlighttt{Join* $\rightarrow$ Base*} is ambigous.
\newline
C++ lets resolve the ambiguity with the \highlightit{scope operator}, 
there is a better solution though.
\newpage
\toclessnparagraph{\texttt{virtual} Inheritance}

\begin{wrapfigure}{L}{0.4\textwidth}% adapt widths of minipages to your needs
\begin{center}
    \begin{tikzpicture}%[>=triangle 45,font=\sffamily]
        \node (X) at (0,1.5) [rectangle,draw] {\textbf{Base}};
        \node (Y) at (-1.5,0) [rectangle,draw] {\textbf{Der1}};
        \node (Z) at (1.5,0) [rectangle,draw] {\textbf{Der2}};
        \node (U) at (0,-1.5) [rectangle,draw] {\textbf{Join}};

        \draw [semithick,dashed] (X) -- (Y);
        \draw [semithick,dashed] (X) -- (Z);
        \draw [semithick,dashed] (Y) -- (U);
        \draw [semithick,dashed] (Z) -- (U);
        
        \node at (2,0.5) {\highlighttt{virtual}};
        \node at (-2,0.5) {\highlighttt{virtual}};
=    \end{tikzpicture}
    \caption{A Nicer Diamond.}
    \label{fig:undreaded-diamond}
\end{center}
\end{wrapfigure}%
%~ \vspace{2cm}

To avoid duplicates in the \highlighttt{Join class}, every class that 
derives from \highlighttt{Base} needs to \highlighttt{virtual}-inherit. 
This way, an instance of \highlighttt{Join} will contain only one 
\highlighttt{Base} subobject and this solution is better than using 
fully qualified names.

\clearpage
\toclessnparagraph{delegate to a sister class}
\highlighttt{virtual} inheritance helps achieving delegation to a 
sister class.
\begin{lstlisting}[language=C++]
class Base
{
public:
    virtual void foo() = 0;
    virtual void bar() = 0;
};

class Der1 : public virtual Base
{
public:
    virtual void foo();
};
void Der1::foo()
{ bar(); }

class Der2 : public virtual Base
{
public:
    virtual void bar();
};

class Join : public Der1, public Der2
{
public:
    // ...
};
\end{lstlisting}
\highlighttt{Join} objects contain the \highlighttt{Der1}'s overridden 
\highlighttt{foo()} and the \highlighttt{Der2}'s overridden 
\highlighttt{bar()}. For this reason, when \highlighttt{Der1::foo()} 
calls \highlighttt{this->bar()} it ends up calling 
\highlighttt{Der2::bar()}.

\toclessnparagraph{Considerations needed when}
\vspace{-1cm}
\toclessnsubparagraph{\quad $\bullet$ using \texttt{virtual} inheritance}
Usually, \highlighttt{virtual} base classes and the ones that directly 
inherits from them are pure abstract classes, and contain very little 
if any data.

\toclessnsubparagraph{\quad $\bullet$ inheriting from a class that uses \texttt{virtual} inheritance}
The initialization list of the most-derived-class's constructor directly 
invokes the virtual base class's constructor.
\newline
C++ has a set of rules to make sure the \highlighttt{virtual} base 
class's constructor and destructor get called once per instance
\begin{itemize}
    \item \highlighttt{virtual} base classes are constructed befor all 
    non-highlighttt{virtual} base classes.
    \item constructors for \highlighttt{virtual} base classes anywhere 
    in you class's inheritance hierarhy are called by the most derived 
    class's constructor.
\end{itemize}
A consequence is the need to pass whatever parameters are required to call 
the \highlighttt{virtual} base class's constructor. In case of multiple 
\highlighttt{virtual} base classes, the most-derived class's constructor 
needs more parameters than other contexts require.
\newline
The practice of leaving the \highlighttt{virtual} base class without any 
data to initialize implies that the constructor probably takes no 
parameters.

\toclessnsubparagraph{\quad $\bullet$ using a class that uses \texttt{virtual} inheritance}
\highlighttt{dynamic\char`_cast} is preferable to C-style downcasts.
 
\toclessnparagraph{Order of constructors in a multiple and/or \texttt{virtual} inheritance}
First constructors to be executed are the virtual base classes' ones, 
in the order of appearance of base class names in the declaration of 
the derived class.
\newline
After the previous step, the construction order is from base class to 
derived class. Indeed, the compiler at first makes an hidden call to 
the non-virtual base classes' constructor in the derived class's 
constructor.
\newline
The previous rule is applied recursively: if \highlighttt{B1} inherits 
from \highlighttt{B1a} and \highlighttt{B1b}, and \highlighttt{B2}
inherits from \highlighttt{B2a} and \highlighttt{B2b}, the final order 
of initialization is: \newline
\highlighttt{B1a $\rightarrow$ B1b $\rightarrow$ B1 $\rightarrow$ B2a $\rightarrow$ B2b $\rightarrow$ B2 $\rightarrow$ D}.
\newline
The last order is determined by the order that the base classes appear 
in the declaration of the class.

\toclessnparagraph{Order of destructors in a multiple and/or \texttt{virtual} inheritance}
For what concerns the non-virtually inheritance, the destructor order 
is the opposite of the constructor order:
\newline
\highlighttt{D $\rightarrow$ B2 $\rightarrow$ B2b $\rightarrow$ B2a $\rightarrow$ B1 $\rightarrow$ B1b $\rightarrow$ B1a}
\newline
Afterwards, the destructor of the virtual base classes that appear in 
the hierarchy are called, taking into account a class only when it uniquely 
appear for the first time.
\newpage


\subsection{Beyond Classes}
\subsubsection{\href{https://isocpp.org/wiki/faq/templates}{Templates}}

A class template is a description of how to build a family of classes 
that all look basically the same, and a function template describes how 
to build a family of similar looking functions.
\newline
A class template is sometimes called \highlight{parameterized type} or 
\highlight{genericity}.
\newline
Class templates are often used to build type safe containers 
(although this only scratches the surface for how they can be used).

\begin{lstlisting}[language=C++]
// This would go into a header file such as "Array.h"
template<typename T>
class Array {
public:
  Array(int len=10)                
  : len_(len), data_(new T[len]) { }
 ~Array()                          { delete[] data_; }
  int len() const                  { return len_;     }
  const T& operator[](int i) const { return data_[check(i)]; }
  T&       operator[](int i)       { return data_[check(i)]; }
  Array(const Array<T>&);
  Array(Array<T>&&);
  Array<T>& operator= (const Array<T>&);
  Array<T>& operator= (Array<T>&&);
private:
  int len_;
  T*  data_;
  int check(int i) const {
    assert(i >= 0 && i < len_);
    return i;
  }
};
\end{lstlisting}
The template in the previous example removes the need to rewrite the 
same class for different data-types, using a general identifier called 
\texttt{T} in the example.
\newline
Like for a normal class, templates' methods can be defined outside the 
class.
\newline
Instantiations of class templates need to be explicit about the parameters 
over which they are instantiating, for example
\begin{lstlisting}[language=C++]
int main()
{
  Array<int>        ai;
  Array<Array<int>> aai;  // before C++11, 
                          // compiler would see a >> (right-shift)
}
\end{lstlisting}

The syntax for a function template is the following
\begin{lstlisting}[language=C++]
template<typename T>
void swap(T& x, T& y)
{
  T tmp = x;
  x = y;
  y = tmp;
}
\end{lstlisting}

Template functions usually do not need to be explicit about the parameters 
over which they are instantiating. The compiler can usually determine 
them automatically. Every time we used \texttt{swap()} with a given pair 
of types, the compiler will go to the above definition and will create 
yet another “template function” as an instantiation of the above.
\newline
Every once in a while it can be needed to explicitly select which version 
of a function template to use, because the compiler cannot deduce the type 
at all, or it would deduce the wrong type.
\newline
For example, the function template might have not any parameter of its 
template argument types 
%~ \begin{lstlisting}[language=C++]
%~ template<typename T>
%~ void f()
%~ {
  %~ // ...
%~ }
%~ \end{lstlisting}
and calls to this function need to explicitly select the template version.
\newline
It could be useful to force the compiler to select a function template 
that requires certain promotions on arguments. With a 
template like the following
\begin{lstlisting}[language=C++]
template<typename T>
void g(T x)
{
  // ...
}
\end{lstlisting}
\texttt{g("xyz")} would select \text{g<char*>(char*)} but to call the 
template for strings it is possible to use \texttt{g<std::string>("xyz")}. 
\newline
It might also happen that the function template has two parameters of the  
same type but the user calls it with two different types
\begin{lstlisting}[language=C++]
template<typename T>
void g(T x, T y);
int m = 0;
long n = 1;
g<int>(m, n);
\end{lstlisting}
In this case, the compiler can not deduce itself, and an explicit selection 
must be done.
\newline
It is possible to use \highlight{template specialization}, in case the 
function template’s observable behavior is consistent for all 
the T types with the differences limited to implementation details. 
For example, a function template that represent an object as a string 
could make use of \texttt{std::boolalpha} to represent a \texttt{bool} as 
\texttt{false} and \texttt{true}, or \texttt{std::precision} to represent a 
\texttt{double}, and so on.
\newline
In case the function template has common code along with a relatively 
small amount of \texttt{T}-specific code, 
\newline
Function templates participate in \highlight{name resolution for overloaded functions}, 
the rules are different though. For a template to be considered in overload 
resolution, the type has to match exactly. If the types do not match exactly, 
the conversions are not considered and the template is simply dropped 
from the set of viable functions. That’s what is known as \highlight{SFINAE} — 
Substitution Failure Is Not An Error.
\newline
The reasons why ... are 
\begin{itemize}
    \item A template is not a class or a function. A template is a “pattern” 
          that the compiler uses to generate a family of classes or functions.
    \item In order for the compiler to generate the code, it must see both 
          the template definition (not just declaration) and the specific 
          types/whatever used to “fill in” the template. For example, 
          if you’re trying to use a \texttt{Foo<int>}, the compiler must 
          see both the \texttt{Foo} template and the fact that you’re 
          trying to make a specific \texttt{Foo<int>}.
    \item Your compiler probably doesn’t remember the details of one \texttt{.cpp} 
          file while it is compiling another .cpp file. It could, but most 
          do not.
          \newline
          This is called the \highlight{separate compilation model}.
\end{itemize}

\subsubsection{\href{https://isocpp.org/wiki/faq/containers}{Container Classes}}
A contiguous array is the optimal construct for accessing a sequence of 
objects in memory, in terms of space and time.
\newline
In C++ there are better alternatives to plain C arrays, easier to write 
to read and less prone to errors but with same speed. Two of the main 
issues with C-style arrays are
\begin{itemize}
    \item C array doesn't know its own size
    \item the name of a C array converts to its first element almost always
\end{itemize}
The first issue implies that there can not be array assignment. Moreover, 
an array size must be determined at compile-time.
\newline
The second issue makes array interacts badly with inheritance. In the 
following code
\begin{lstlisting}[language=C++]
class Base { void fct(); /* ... */ };
class Derived : Base { /* ... */ };
void f(Base* p, int sz)
{
    for (int i=0; i<sz; ++i) p[i].fct();
}
Base ab[20];
Derived ad[20];
void g()
{
    f(ab,20);
    f(ad,20);    // disaster!
}
\end{lstlisting}
the array of \texttt{Derived} is treated as an array of \texttt{Base} 
and the different sizes of the two break the subscripting without the 
compiler clearly notifying it, as it would do with \texttt{vector}s.
\newline
Some of the benefits that come with the container classes are the following
\begin{itemize}
    \item it is possible to insert an element in almost any place without 
          the need 
    \item
\end{itemize}

Some of the things that can be done using arrays are
\begin{itemize}
    \item \texttt{std::map} can be used as a lookup table;
    \item \texttt{std::map} can be used as a sparse array or matrix;
    \item \texttt{std::array} can be used in place of plain arrays, since 
          it provides the user with boundary checks, memory management 
          in case of exceptions
    \item 
    \item 
\end{itemize}

The storage for a \texttt{std::vector<T>} or \texttt{std::array<T, N>} 
is guaranteed to be \highlight{contiguous}. This means that for any 
non-empty \texttt{std::vector<T>} or \texttt{std::array<T,N>} named 
\texttt{v} and an integer \texttt{n} in the range \texttt{0 \dots v.size()-1}, 
it is always sure that \texttt{\&v[0] + n == \&v[n]}.
\newline
On the other side, \texttt{v.begin()} is not guaranteed to be a \texttt{T*} 
therefore it is not guaranteed to have the same value as \texttt{\&v[0]}. 
Using \texttt{v.data()} is always safe though.
\newline
Containers provided by C++ are homogeneous, in the sense that they 
contain elements of the same type. In order to hold elements of several 
different types, it is possible to use a container of pointers to a 
polymorphyc type.
Homogeneous container are the easiest to use, give the best compile-time 
error messages and have no unnecessary run-time overhead.
\newline 
To have an heterogeneous container, it is possible to define a common 
interface for all the elements and make a container of pointers:
\begin{lstlisting}[language=C++]
class Io_obj { /* ... */ };    // the interface needed to take part in object I/O
vector<Io_obj*> vio;           // if you want to manage the pointers directly
vector<shared_ptr<Io_obj>> v2; // if you want a "smart pointer" to handle the objects
\end{lstlisting}
An alternative is to use libraries like 
\href{http://www.boost.org/doc/libs/1_55_0/doc/html/any.html}{\texttt{boost::any}}.
\newline
If a pointer is used, objects in the container can be accessed through 
the pointers themselves, levereaging the polymorphic behavior as well. 
To know the exact type of the object, it is possible to use 
\texttt{dynamic\char`_cast<>} or \texttt{typeid()}. To copy the container 
of different objects, the \highlight{virual costructor idiom} 
is probably needed (see~\ref{npar:virtual-ctor}).
\newline
In order to make things easier, it is possible to rely on a \texttt{shared\char`_pointer}.
\newline
It could be necessary to have a container of disjoint classes, that do 
not share a common base class. In this case, it is possible to create 
a handle class and make a container of handle objects. This approach 
comes with the benefit of encapsulating most of the memory management 
and object lifetime, but also with the downside of maintenance every time 
there are changes to the set of elements that can be contained.

\href{https://isocpp.org/wiki/faq/containers#slow-containers}{check and include}


\subsubsection{\href{https://isocpp.org/wiki/faq/class-libraries}{Class Libraries}}
The \highlighttt{STL} is a library that consists mainly of container 
classes, along with iterators and algorithms to work with the content 
of the containers. Techically speaking it is no longer meaningful since 
the classes it provides have been integrated into the standard library. 
Sometimes, it is described as a separate library though.
\newline
A good resource about STL is
\newline
\href{http://stdcxx.apache.org/doc/stdlibug/index.html}{Apache C++ Standard Library} 
\newline
The National-Institute-of-Health's-class-library, \highlight{NIHCL}, is 
a C++ translation of the Smalltalk class library. It can be retrieved 
\href{128.231.128.7/pub/NIHCL/nihcl-3.0.tar.Z}{here}.
\newline
For what concerns Numerical Recipes, following some resources
\newline
\href{http://www.siam.org/}{LAPACK: Linear Algebra Subroutine Library.}
\newline
\href{https://www.netlib.org/}{Netlib is a collection of mathematical software, papers, and databases.}
\newline
Other information about C++ class libraries can be found in the following 
resources
\newline
\href{http://www.trumphurst.com/cpplibs/cpplibs.php}{Available C++ Libraries FAQ maintained by Nikki Locke.}
\newline
\href{http://www.mathtools.net/}{Mathtools.net A Resource for the Technical Computing Community.}
\newline
\href{https://www.boost.org/}{boost C++ libraries}
\newline

\subsubsection{\href{https://isocpp.org/wiki/faq/exceptions}{Exceptions and Error Handlings}}

\subsubsection{\href{https://isocpp.org/wiki/faq/value-vs-ref-semantics}{Reference and Value Semantics}}

\highlighttt{virtual} data allows a derived class to change the exact 
class of a base class's member object. It is not directly supported by C++, 
it can be simulated though.
\newline
To \highlight{simulate a \texttt{virtual} data},
\begin{itemize}
    \item the base class must have a \highlight{pointer to the member object};
    \item the derived class must provide a \highlight{\texttt{new} object to be pointed 
    to by the base class's pointer}
\end{itemize}
The base class would also have one or more constructors that provide their own 
referent


\subsubsection{\href{https://isocpp.org/wiki/faq/inline-functions}{Inline Functions}}
Conceptually similar to what happens with a \texttt{\#define} macro, 
the compiler inline-expands a function call, inserting the function's 
code directly into the caller's code stream.
\newline
Because of the \highlight{procedurally integration} the optimizer can 
make on the called code, this expansion might improve the perfomances. 
Considering the following code
\begin{lstlisting}[language=C++]
void f()
{
  int x = /*...*/;
  int y = /*...*/;
  int z = /*...*/;
  // ...code that uses x, y and z...
  g(x, y, z);
  // ...more code that uses x, y and z...
}
\end{lstlisting}
in a typical C++ implementation with registers and a stack, registers 
and parameters get written on the stack just before the call to \texttt{g(x)}, 
then the parameters get read from the stack into \texttt{g(x)} and read 
again to restore the register after \texttt{g(x)} returns to \texttt{f(x)}.
\newline
If the compiler inline-expands the call to \texttt{g(x)}, there would not 
be a function call and the not needed writes and reads that come with it.
\newline
It is not automatic that an \texttt{inline} function can improve the 
perfomance, since it might depend on multiple factors.
\begin{itemize}
    \item 
    \item 
    \item 
\end{itemize}

\texttt{inline} functions should be used instead of \texttt{macro}s

https://isocpp.org/wiki/faq/misc-technical-issues\#macros-with-if \\
https://isocpp.org/wiki/faq/misc-technical-issues\#macros-with-multi-stmts \\
https://isocpp.org/wiki/faq/misc-technical-issues\#macros-with-token-pasting \\
https://isocpp.org/wiki/faq/big-picture\#use-evil-things-sometimes

In order to make both a non-member and function \texttt{inline}, the 
declaration stays the same but the definition must prepended with 
the keyword \texttt{inline} and written in a \highlight{header file}. 
In case the definition is placed in a \texttt{.cpp} file, unless the 
function is used only in that single file there will be an 
\texttt{unresolved external} error from the linker.
\newline
Another way to tell the compiler that a function is \texttt{inline} is 
to define the member function in the class body without the \texttt{inline} 
keyword in the definition. This way might be confusing 
for readers though, sinche it mixes the what (behavior) with the how 
(implementation). Therefore, in case of a highly reused class the 
definition outside the class should be preferred.
\newline
In case of an \texttt{inline} member function defined outside the class, 
the keyword must be put on the definition only, as this separation makes 
it easier for the user to read the header file to determine the 
\highlight{observable semantics} or \highlight{external behavior}.

\subsubsection{\href{https://isocpp.org/wiki/faq/pointers-to-members}{Pointers to Member Functions}}
Pointers to member functions are different than pointers to simple functions.
\newline
Considering the function
\begin{lstlisting}[language=C++]
int f(char a, float b);
\end{lstlisting}
its type changes depending on whether it is an ordinary function or a 
non-\texttt{static} member of an immaginary \texttt{Fred} class.
\begin{lstlisting}[language=C++]
int (*)(char,float);        // in case of an ordinary function
                            // or a static member function
int (Fred::*)(char,float)   // in case of a member function 
                            // of class Fred
\end{lstlisting}
Pointers to member functions require elaborate tecniques to be passed to 
signal handler, X event callback, system call that starts a thread/task 
and so on, since the interrupts are unable to provide the member function 
with a \texttt{this} pointer.
\newline 
One possible solution is to use a \texttt{static} member as interrupt.
\newline
When creating pointers to member a function for a class, for example
\begin{lstlisting}[language=C++]
class Fred 
{
public:
  int f(char x, float y);
  // ...
};
\end{lstlisting}
it useful to rely on \texttt{typedef} to create a pointer \texttt{FredMemFn} 
to any member of \texttt{Fred} class that takes a \texttt{(char, float)}
\begin{lstlisting}[language=C++]
typedef  int (Fred::*FredMemFn)(char x, float y);
\end{lstlisting}
Consequently, it is trivial to declare a member-function pointer
\begin{lstlisting}[language=C++]
int main()
{
  FredMemFn p = &Fred::f;
  // ...
}
\end{lstlisting}
or a function that receive a member-function pointer
\begin{lstlisting}[language=C++]
void userCode(FredMemFn p)
{ /*...*/ }
\end{lstlisting}
or a function that returns a member-function pointer
\begin{lstlisting}[language=C++]
void userCode(FredMemFn p)
{ /*...*/ }
\end{lstlisting}
C++17 provides the programmer with an 
\href{https://en.cppreference.com/w/cpp/utility/functional/invoke}{\texttt{std::invoke}} 
\begin{lstlisting}[language=C++]
void userCode(Fred& fred, FredMemFn p)
{
  int ans = std::invoke(p, fred, 'x', 3.14);
  // Would normally be: int ans = (fred.*p)('x', 3.14);
  // ...
}
\end{lstlisting}
In case \texttt{std::invoke} is not accessible, a macro like the following
\begin{lstlisting}[language=C++]
#define CALL_MEMBER_FN(object,ptrToMember)  ((object).*(ptrToMember))
\end{lstlisting}
could help reducing code complexity
\begin{lstlisting}[language=C++]
void userCode(Fred& fred, FredMemFn p)
{
  int ans = CALL_MEMBER_FN(fred,p)('x', 3.14);
  // Would normally be: int ans = (fred.*p)('x', 3.14);
  // ...
}
\end{lstlisting}
In case the member function is constant, the \texttt{typedef} should look 
like
\begin{lstlisting}[language=C++]
typedef int (Fred::*FredMemFn)(int) const;
\end{lstlisting}
 
\highlight{pointer-to-member access operators} are the last resource,
in case neither the \texttt{std::invoke} nor a macro are possible options
\begin{lstlisting}[language=C++]
class Fred { /*...*/ };
typedef  int (Fred::*FredMemFn)(int i, double d);
void sample(Fred x, Fred& y, Fred* z, FredMemFn func)
{
  x.*func(42, 3.14);
  y.*func(42, 3.14);
  z->*func(42, 3.14);
}    
\end{lstlisting}
Neither a pointer to a member function nor a pointer to a function can 
be converted to a \texttt{void*}
\newline
A \highlight{functionoid} is similar to a function pointer but with 
more flexibility and thread safety.

\subsubsection{\href{https://isocpp.org/wiki/faq/serialization}{Serialization and Unserialization}}


\subsection{Value Categories}
\label{subsubsec:value-categories}
\href{https://en.cppreference.com/w/cpp/language/value_category}{Cppreference documentation about \textbf{value categories}}. 
\newline    
Due to the expressiveness of the C++ language, the terms \highlightbf{lvalue} 
and \highlightbf{rvalue} references evolved in a more complex set.

\toclessnparagraph{glvalues}
\highlight{\textbf{G}eneralized} lvalue is used to refer to something that could be 
either a \highlightbf{lvalue} or \highlightbf{xvalue}.

\toclessnparagraph{rvalues}
This term is inherited from C, in which \highlightbf{rvalues} can be on the 
\highlight{\textbf{r}ight} side of an assignment. In C++ this term can refer to 
things that are either \highlightbf{xvalues} and \highlightbf{prvalues}.

\toclessnparagraph{lvalues}
This term is also inherited from C, in which \highlightbf{lvalues} can be 
on the \highlight{\textbf{l}eft} side of an assignment. Therefore, the simplest one 
is the name of a variable in a declaration.
\newline
Any expression resulting in an \highlightit{lvalue reference} is also an 
\highlightit{lvalue}.
\begin{lstlisting}[language=C+++]
int v;          // v is an lvalue of type int
A*  p1;         // *p1 is an lvalue
A&  r1 = v1;    // r1 is an lvalue
A&  f();        // f() is an lvalue
\end{lstlisting}

Accessing members of objects through pointers or references yields 
\highlightit{lvalues}. For example, with the following code
\begin{lstlisting}[language=C+++]
struct Foo
{
    Bar a;
    Bar b;
};

Foo& f2();
Foo* p2;
Foo  v2;
\end{lstlisting}
\texttt{f2().a}, \texttt{p2->b} and \texttt{v2.a} are all 
\highlightit{lvalues} of type \texttt{Bar}.
\newline
\highlightbf{String literals} are \highlightit{lvalues} of type array.
\newline
To conclude, a \highlightbf{named object} declared with an 
\highlightit{rvalue reference} is an \highlightit{lvalue}. The name rvalue 
reference indicates that it can bind to an \highlightit{rvalue}, and once 
you've declared a variable and given it a name it is an \highlightit{lvalue}.
\begin{lstlisting}[language=C+++]
void Foo(A&& bar)
{
    // within Foo, bar is an lvalue 
    // which will only bind to rvalues
}
\end{lstlisting}

\toclessnparagraph{xvalues}
This term indicates \highlight{e\textbf{X}piring} value: an unnamed object that 
will be destroyed soon. \highlightbf{xvalues} can be treated either as 
\highlightit{glvalues} or \highlightit{rvalues}, based on the context.
\newline
Usually, \highlightit{xvalues} only arise through \highlightit{explicit casts} and 
\highlightit{function calls}. If an expression is \highlight{cast to an \textit{rvalue} reference 
to some type \texttt{T}} then the result is an \highlight{\textit{xvalue} of type \texttt{T}} 
\begin{lstlisting}[language=C+++]
static_cast<A&&>(v1) // yields an xvalue of type A
\end{lstlisting}
If the return type of a function is an \highlightit{rvalue reference} to some 
type \highlightbf{T}, the result is an \highlightit{xvalue} of type \highlightbf{T}.
For example, the \texttt{std::move} is declared as
\begin{lstlisting}[language=C+++]
template <typename T>
constexpr remove_reference_t<T>&& move(T&& t) noexcept;
\end{lstlisting}
\texttt{std::move(v1)} is an \textbf{xvalue} of type \texttt{A}. 
The type deduction rules deduce \textbf{T} to be \texttt{A\&} 
because \texttt{v1} is an \textbf{lvalue}.

\toclessnparagraph{prvalues}
\highlight{\textbf{P}ure} \highlightit{rvalue} is an \highlightit{rvalue} 
which is not an \highlightit{xvalue}
\newline
Literals other than string literals are \highlightit{prvalues}. For example, 
$42$ is a \highlightit{prvalue} of type \texttt{int} and $3.141f$ is a 
\highlightit{prvalue} of type \texttt{float}
\newline
\highlightbf{Temporaries} are \highlightit{prvalues}. For example any temporaries 
created as a result of an implicit conversion is also a \highlightit{prvalue}. 
In the following code
\begin{lstlisting}[language=C+++]
int consume_string(std::string&& s);
int i = consume_string("hello");
\end{lstlisting}
the string literal in the function call will implicitly convert to a 
temporary of type 
\newline 
\texttt{std::string}, which can \highlight{bind to the rvalue 
reference} used for the function parameter since the temporary is a 
\highlightit{prvalue}

\subsection{Reference Binding}
\href{https://en.cppreference.com/w/cpp/language/reference}{cppreference documentation} 
about reference declaration.
\newline
The \highlightit{main difference} between \highlightit{lvalues} and \highlightit{rvalues} is 
the way they bind to references. The differences in type deduction rules 
can have a big impact too.

\subsection{lvalue references}
These references are declared with a single ampersand $\&$.
\newline
A \highlightbf{non-const lvalue reference will only bind to non-const 
lvalues} of the same type, or a class derived from the referenced type
\begin{lstlisting}[language=C+++]
class C : public A{};

int i = 42;
A a;
B b;
C c;
const A ca{};

A&   r1  = a;
A&   r2  = c;
A&   r3  = b;     // error, wrong type
int& r4  = i;
int& r5  = 42;    // error, cannot bind rvalue
A&   r6  = ca;    // error, cannot bind const object
                  // to non const ref
A&   r7  = r1;
A&   r8  = A();   // error, cannot bind rvalue
A&   r9  = B().a; // error, cannot bind rvalue
A&   r10 = C();   // error, cannot bind rvalue
\end{lstlisting}
A \highlightbf{const lvalue reference can also bind to rvalues}. The 
object bound to the reference must have the same type as the referenced 
type or a class derived from the referenced type. It is possible to 
bind both const and non-const values to a const lvalue reference.
\begin{lstlisting}[language=C+++]
const A&   cr1  = a;
const A&   cr2  = c;
const A&   cr3  = b;     // error, wrong type
const int& cr4  = i;
const int& cr5  = 42;    // rvalue can bind OK
const A&   cr6  = ca;    // OK, can bind const object to const ref
const A&   cr7  = cr1;
const A&   cr8  = A();   // OK, can bind rvalue
const A&   cr9  = B().a; // OK, can bind rvalue
const A&   cr10 = C();   // OK, can bind rvalue
const A&   cr11 = r1;
\end{lstlisting}
Binding a temporary object (i.e. a \highlightit{prvalue}) to a const lvalue 
reference \highlightit{extends the lifetime} of that temporary to the 
lifetime of the reference. In the previous example, it is possible 
to use \texttt{r8}, \texttt{r9} and \texttt{r10} later in the code.
\newline
Lifetime extension does not extend to references initialized from the 
first reference. If a function parameter is a const lvalue reference 
and gets bound to a temporary passed to the function call, the temporary 
is destroyed when the function returns. This happens also if the 
reference was stored in a longer-lived variable, such as a member of a 
newly constructed object.
\newline
\highlighttt{volatile} and \highlighttt{const volatile} lvalue references are much 
less used but behave as expected: volatile \texttt{T\&} bind to a volatile or 
non-volatile, non-const lvalue of type T or a class derived from T. 
However, volatile const lvalue references do not bind to rvalues.

\subsection{Reference Collapsing}
\label{subsubsec:reference-collapsing}
In templates and \texttt{typedef}/\texttt{using} aliases it is possible to form a 
``reference to reference.'' The \highlight{reference collapsing} rules then apply: 
an rvalue reference to an rvalue reference collapses to an rvalue reference; 
every other combination collapses to an lvalue reference.
\begin{lstlisting}[language=C++]
using lref = int&;
using rref = int&&;
int n;

lref&  r1 = n; // int&
lref&& r2 = n; // int&
rref&  r3 = n; // int&
rref&& r4 = 1; // int&&
\end{lstlisting}
This is the mechanism behind forwarding references (\texttt{T\&\&} in a deduced context) 
and \texttt{std::forward}. See also Section~\ref{subsubsec:value-categories}.

\newpage
\section{Contemporary C++: new language and library features}
\label{sec:contemporary-cpp}
This section summarizes language and library features by standard (C++11 onward). 
Entries are adapted from the open cheatsheet 
\href{https://github.com/AnthonyCalandra/modern-cpp-features}{AnthonyCalandra/modern-cpp-features}, 
with links back to the source for full detail. 
Where this book has a dedicated chapter (smart pointers, concurrency, containers), 
prefer that chapter for depth and use the entries here as a standard-by-standard index.

\subsection{C++11 new language features}
Following a summary of the new language features introduced with C++11.

%~ MOVE SEMANTICS
\subsubsection{\href{https://github.com/AnthonyCalandra/modern-cpp-features/tree/master\#move-semantics}
                    {move semantics}}
Move semantics is a new way of moving resources around in an optimal way, 
since it helps \highlightit{avoiding unnecessary copies} of temporary objects. 
It is based on rvalue references and plays an \highlightit{important role in 
exception safety} for C++.
\newline
Moving an object means to transfer ownership of some resource it manages 
to another object.

The first benefit of move semantics is performance optimization. When an 
object is about to reach the end of its lifetime, either because it's a 
temporary or by explicitly calling \texttt{std::move}, a move is often a cheaper 
way to transfer resources. For example, moving a \texttt{std::vector} is just 
copying some pointers and internal state over to the new vector -- copying 
would involve having to copy every single contained element in the vector, 
which is expensive and unnecessary if the old vector will soon be destroyed.

Moves also make it possible for non-copyable types such as \texttt{std::unique\char`_ptrs} 
(smart pointers) to guarantee at the language level that there is only 
ever one instance of a resource being managed at a time, while being able 
to transfer an instance between scopes.

%~ VARIADIC TEMPLATES
\subsubsection{\href{https://github.com/AnthonyCalandra/modern-cpp-features/tree/master\#variadic-templates}
                    {variadic templates}}
The \verb|...| syntax creates a parameter pack or expands one.
\newline
A template parameter pack is a template parameter that accepts zero or 
more template arguments (non-types, types, or templates).
\newline
A template with at least one parameter pack is called a variadic template.

\begin{lstlisting}[language=C++]
template <typename... T>
struct arity {
  constexpr static int value = sizeof...(T);
};
static_assert(arity<>::value == 0);
static_assert(arity<char, short, int>::value == 3);
\end{lstlisting}

An interesting use for this is creating an initializer list from a 
parameter pack in order to iterate over variadic function arguments.

\begin{lstlisting}[language=C++]
template <typename First, typename... Args>
auto sum(const First first, const Args... args) -> decltype(first) {
  const auto values = {first, args...};
  return std::accumulate(values.begin(), values.end(), First{0});
}

sum(1, 2, 3, 4, 5); // 15
sum(1, 2, 3);       // 6
sum(1.5, 2.0, 3.7); // 7.2
\end{lstlisting}

%~ RVALUE REFERENCES
\subsubsection{\href{https://github.com/AnthonyCalandra/modern-cpp-features/tree/master\#rvalue-references}
                    {rvalue references}}
An rvalue reference to T, which is a non-template type parameter 
(such as int, or a user-defined type), is created with the syntax \texttt{T\&\&}. 
Rvalue references only bind to rvalues.
An \highlightit{rvalue reference} can be used to extend the lifetime of 
temporary objects, making the referenced object modifiable through 
them. The reference must be const in order to bind to a const object, 
though const rvalue reference are much rarer.
\begin{lstlisting}[language=C+++]
const A make_const_A();
A&&       rr1  = a;                // error, cannot bind lvalue 
                                   // to rvalue reference
A&&       rr2  = A();
A&&       rr3  = B();              // error, wrong type
int&&     rr4  = i;                // error, cannot bind lvalue
int&&     rr5  = 42;
A&&       rr6  = make_const_A();   // error, cannot bind 
                                   //        const object 
                                   //        to non const ref
const A&& rr7  = A();
const A&& rr8  = make_const_A();
A&&       rr9  = B().a;
A&&       rr10 = C();
A&&       rr11 = std::move(a);     // std::move returns an rvalue
A&&       rr12 = rr11;             // error rvalue references 
                                   //       are lvalues
\end{lstlisting}
rvalue references extend the lifetime of temporary objects in the same 
way const lvalue references do, therefore temporaries associated with 
\texttt{rr2}, \texttt{rr5}, \texttt{rr7}, \texttt{rr8}, \texttt{rr9} 
and \texttt{rr10} remain valid until the corresponding references are 
destroyed. 

%~ FORWARDING REFERENCES
\subsubsection{\href{https://github.com/AnthonyCalandra/modern-cpp-features/tree/master\#forwarding-references}
                    {forwarding references}}
\href{https://en.cppreference.com/w/cpp/utility/forward}{Cppreference documentation about \textbf{forward} utility}. 
\newline
\href{https://en.cppreference.com/w/cpp/language/reference#Forwarding_references}{Cppreference about \textbf{forwarding references}}.
\newline
Perfect forwarding is an important C++0x feature built on top of rvalue 
references. It provides the programmer with the capability automatically 
apply the move semantics, even when there are intervening function calls 
to separate the source and the destination of a move.
\newline
Examples include \highlightit{constructors} and \highlightit{setter functions}, 
that can forward arguments they receive directly to the data member they 
are initializing or setting. Standard Library functions like 
\highlighttt{make\char`_shared} "perfect-forwards" its arguments to the class 
constructor of whatever object the to-be-created \texttt{shared\char`_ptr} 
is to point to.
\newline
The idiomatic expression of the perfect forwarding make use of the 
\texttt{std::forward} 
\begin{lstlisting}[language=C+++]
template <typename T>
void wrapper(T&& arg) 
{
    wrapped(std::forward<T1>(t1), std::forward<T2>(t2));
}
\end{lstlisting}

%~ INITIALIZER LIST
\subsubsection{\href{https://github.com/AnthonyCalandra/modern-cpp-features/tree/master\#initializer-lists}
                    {initializer lists}}
\href{https://en.cppreference.com/w/cpp/utility/initializer_list}{Cppreference documentation about Initializer list}
\newline
An object of type \highlighttt{std::initializer\char`_list<T>} is a 
lightweight proxy object that provides access to an array of objects of 
type \highlighttt{const T}. 

\subsubsection{Uniform initalization syntax and semantics}
C++98 provides the users with several ways of initializing an object. 
These ways depend on the type and the initialization context and have 
different ways of working.
\newline
Some examples can be the following code lines
\begin{lstlisting}[language=C++]
string a[] = { "foo", " bar" };        // ok: initialize array
vector<string> v = { "foo", " bar" };  // error: initializer list 
                                       // for non-aggregate vector
void f(string a[]);
f( { "foo", " bar" } );                // syntax error: 
                                       // block as argument
\end{lstlisting}
and
\begin{lstlisting}[language=C++]    
int a = 2;            // "assignment style"
int aa[] = { 2, 3 };  // assignment style with list
complex z(1,2);       // "functional style" initialization
x = Ptr(y);           // "functional style" for 
                      // conversion/cast/construction
        
int a(1);             // variable definition
int b();              // function declaration
int b(foo);           // variable definition or function declaration
\end{lstlisting}

C++11 introduces a \highlight{\texttt{\{\}}-initializer}, 
which provides a much simpler and consistent way to perform initialization, 
for all different initialization context as in the following example
\begin{lstlisting}[language=C++]
X x1 = X{1,2}; 
X x2 = {1,2};   // the = is optional
X x3{1,2}; 
X* p = new X{1,2}; 
struct D : X 
{
    D(int x, int y) :X{x,y} { /* ... */ };
};
struct S
{
    int a[3];
    S(int x, int y, int z) :a{x,y,z} { /* ... */ }; // solution to
                                                    // old problem
};
\end{lstlisting}

\highlight{\texttt{\{\}}-initializer} produces the same result in any 
context
\begin{lstlisting}[language=C++]
X x{a}; 
X* p = new X{a};
z = X{a};         // use as cast
f({a});           // function argument (of type X)
return {a};       // function return value (function returning X)
\end{lstlisting}

\highlight{\texttt{\{\}}-initializer} is also useful to solve the 
\highlight{most vexing parse} problem, explained in~\ref{subsubsec:most-vex-parse}.

%~ STATIC ASSERTIONS
\subsubsection{\href{https://github.com/AnthonyCalandra/modern-cpp-features/tree/master\#static-assertions}
{static assertions}}
Assertions that are evaluated at compile-time.
\begin{lstlisting}[language=C++]
constexpr int x = 0;
constexpr int y = 1;
static_assert(x == y, "x != y");
\end{lstlisting}

%~ AUTO
\subsubsection{\href{https://github.com/AnthonyCalandra/modern-cpp-features/tree/master\#auto}
                    {auto}}
\label{subsubsec:c++auto}
\href{https://en.cppreference.com/w/cpp/language/auto}{Cppreference documentation about \texttt{\textbf{auto}}}.
\newline
References in bibliography:~\cite{gotw-auto-variablesI}, \cite{gotw-auto-variablesII},
\cite{gotw-almost-always-auto}.
\newline
This placeholder type specifier has the following meanings:
\begin{itemize}
    \item for \highlightbf{variables}: \newline
    the variable that is being declared will be automatically deduced 
    from its initializer;
    \item for \highlightbf{functions}: \newline
    the return type will be deduced from its return statements;
    \item for \highlightbf{non-type template parameters}: \newline
    the type will be deduced from the argument.
\end{itemize}

Following a summary of \highlight{contextes in which \texttt{\textbf{auto}} may 
appear}. 
\newline
In the summary, the 
\href{https://en.cppreference.com/w/cpp/language/template_argument_deduction#Other_contexts}{rules for template argument deduction} 
are referenced.

\begin{itemize}
    \item if used \highlightbf{in the type specifier of a variable} the variable 
    type is deduced from the initializer list using the template 
    argument deduction:
    \begin{lstlisting}[language=C++]
auto x = expr;
    \end{lstlisting}
    In case \highlightit{no pointer or reference} is 
    attached to \texttt{auto}, both \texttt{const} and references are 
    ignored:
    \begin{lstlisting}[language=C++]
int y = 10;
int& r = y;
auto x = r;               // type is int (reference ignored)

const int y = 10;
auto x = y;               // type is int (const ignored)

int y = 10;
const int& r = y;
auto x = r;               // type is int
                          // (const and reference ignored)

const int a[10] = {};
auto x = a;               // type is int*
                          // (array to pointer conversion)
    \end{lstlisting}
    %~ \begin{lstlisting}[language=C++]
%~ auto (*p)() -> int;       // declares p as 
                          %~ // pointer to function returning int
%~ auto (*q)() -> auto = p;  // declares q as 
                          %~ // pointer to function returning T,
                          %~ // deduced from the type of p
    %~ \end{lstlisting}

\highlightit{Modifiers} which may accompany \texttt{auto} will partecipate in the 
type deduction. For example, in the code
    \begin{lstlisting}[language=C++]
const auto& i = expr;
template<class U> void f(constU& u);
    \end{lstlisting}
    the type deduced for the first code line is the same deduced for 
    the argument \texttt{u} of the second line if the function call 
    \texttt{f(expr)} was compiled.
    \newline
    Therefore \highlighttt{auto\char`&\char`&} may be deduced both as an 
    \highlightit{lvalue} reference and an \highlightit{rvalue} reference.
    \newline
    In case \highlighttt{auto} is used to declare multiple variables, the 
    deduced type must match.
    
    \item If used \highlightbf{in a function defined with the trailing return type 
    syntax}, it is just part of the syntax and automatic type deduction 
    is not performed
    \begin{lstlisting}[language=C++]
auto function_name(parameter_list) -> return_type
{
    // Function implementation
}  
    \end{lstlisting}

    \item If used \highlightbf{as return type of a function defined without
    the trailing return type}, or\newline \highlightbf{as a return type 
    of a lambda expression},    the keyword \texttt{auto} indicates 
    that the return type will be deduced from the return statement, 
    using the rules for template argument deduction.

    \item If used \highlight{\textbf{in combination with \texttt{decltype}}} and the declared 
    type of the variable is \texttt{decltype(auto)}, the keyword 
    \texttt{auto} is replaced with the expression (or expression list) 
    of its initializer and the actual type is deduced using the 
    rules for \texttt{decltype}.

    \item If used \highlight{\textbf{along with \texttt{decltype} in the return type of a 
    function}} and the return type of the function is 
    \texttt{decltype(auto)}, the keyword \texttt{auto} is replaced with 
    the operand of its return statement, and the actual return type is 
    deduced using the rules of \texttt{decltype}.

    \item If used in the \highlight{\textbf{type-id of a \texttt{new} expression}}, 
    the type is deduced from the initializer.
    
    \item If used \highlight{\textbf{in the parameter declaration of a non-type 
    template parameter.}}, for example \highlighttt{template<auto I> struct A;}, 
    its type is deduced from the corresponding argument.
    
    \item if used \highlightbf{as a nested scope like in  \texttt{auto::}} ???
    
    \item if used \highlightbf{in the parameter declaration of a lambda-expression}, 
    such lambda expression is a \highlightit{generic lambda}. This means 
    that the folowing generig lambda will be written by the compiler as 
    an instance of the templates which follows 
    \begin{lstlisting}[language=C++]
auto g_lambda = [] (auto a) { return a; } ;
class /* unnamed */
{
public:
    template<typename T>
    T operator () (T a) const { return a; }
};
    \end{lstlisting}

    %~ \item void f(auto) since C++20
    
    \item if used \highlight{\textbf{in a function parameter declaration}}, 
    like \highlighttt{void f(auto);}, the function declaration introduces 
    an abbrevited function template~\cite{cppreference-abbreviated-functemplate}
    
\end{itemize}                    

auto-typed variables are deduced by the compiler according to the type 
of their initializer.
\begin{lstlisting}[language=C++]
auto a = 3.14; // double
auto b = 1; // int
auto& c = b; // int&
auto d = { 0 }; // std::initializer_list<int>
auto&& e = 1; // int&&
auto&& f = b; // int&
auto g = new auto(123); // int*
const auto h = 1; // const int
auto i = 1, j = 2, k = 3; // int, int, int
auto l = 1, m = true, n = 1.61; // error -- `l` deduced to be int, `m` is bool
auto o; // error -- `o` requires initializer
\end{lstlisting}

Extremely useful for readability, especially for complicated types:
\begin{lstlisting}[language=C++]
std::vector<int> v = ...;
std::vector<int>::const_iterator cit = v.cbegin();
// vs.
auto cit = v.cbegin();
\end{lstlisting}
Functions can also deduce the return type using auto. In C++11, a return 
type must be specified either explicitly, or using decltype like so:
\begin{lstlisting}[language=C++]
template <typename X, typename Y>
auto add(X x, Y y) -> decltype(x + y) {
  return x + y;
}
add(1, 2); // == 3
add(1, 2.0); // == 3.0
add(1.5, 1.5); // == 3.0
\end{lstlisting}

The trailing return type in the above example is the declared type 
(see section on \texttt{decltype}) of the expression \texttt{x + y}. 
For example, if \texttt{x} is an integer and \texttt{y} is a double, 
\texttt{decltype(x + y)} is a double. Therefore, the above function will 
deduce the type depending on what type the expression \texttt{x + y} yields. 
Notice that the trailing return type has access to its parameters, 
and this when appropriate.

\newpage
%~ LAMBDA EXXPRESSIONS
\subsubsection{\href{https://github.com/AnthonyCalandra/modern-cpp-features/tree/master\#lambda-expressions}
                    {lambda expressions}}
\vspace{-0.3cm}
\href{https://en.cppreference.com/w/cpp/language/lambda}{Cppreference about \textbf{lambda expressions}}
\newline
A \highlight{lambda expression} produces a \highlight{closure}, i.e.\ an unnamed 
function object of a unique compiler-generated class type that can capture 
variables from the enclosing scope.

\vspace{-0.3cm}
\begin{table}[h!t]
    \begin{tabularx}{\textwidth}{l|X}
    \multirow{11}{*}{captures} & a comma-separated list of zero or more captures, optionally beginning \\
                              & with a capture-default \\
                              & A lambda expression can use a variable without capturing it if the variable \\
                              & \, $\bullet$ is a non-local variable or has static or thread local storage duration \\
                              & \, \, \, (in which case the variable cannot be captured), or \\
                              & \, $\bullet$ is a reference that has been initialized with a constant expression.\\
                              & A lambda expression can read the value of a variable without capturing it \\
                              & if the variable \\
                              & \, $\bullet$ has const non-volatile integral or enumeration type and has been initialized \\ 
                              & \, \, \, with a constant expression, or \\
                              & \, $\bullet$ is constexpr and has no mutable members. \\
    \hline
    \multirow{2}{*}{tparams} & a non-empty comma-separated list of template parameters \\
                             & used to provide names to the template parameters of a generic lambda\\
    \hline
    t-requires & adds constraints to tparams \\
    \hline
    front-attr & since C++23 \textbf{study} \\
    \hline
    params & the parameter list of \texttt{operator()} of the closure type \\
    \hline
    \multirow{6}{*}{specs} & A list of the following specifiers, each specifier is 
                             allowed at most once\\
                           & in each sequence. If not provided, the objects captured by copy are \highlighttt{const} \\
                           &\, $\bullet$ \highlighttt{mutable} \, allows body to modify the objects captured by copy\\
                           &\, $\bullet$ \highlighttt{constexpr} \, since C++17\\
                           &\, $\bullet$ \highlighttt{consteval} \, since C++20\\
                           &\, $\bullet$ \highlighttt{static} \, since C++23\\
    \hline
    \multirow{2}{*}{exception} & provides the dynamic exception specification \\ 
              & or (until C++20) the noexcept specifier for operator() of the closure type \\
    \hline
    \multirow{2}{*}{back-attr} & an attribute specifier sequence applies to the type of operator() of the closure type\\
                               & (and thus the \texttt{[[noreturn]]} attribute cannot be used) \\
    \hline
    trailing-type & \texttt{-> ret} where \texttt{ret} specifies the return type\\
    \hline
    requires & adds constraints to \texttt{operator()} of the closure type\\
    \hline
    body & function body\\
    \hline
    
    %~ \multirow{}{}{} & \\
    \end{tabularx}
\end{table}

\newpage
%~ DECLTYPE
\subsubsection{\href{https://github.com/AnthonyCalandra/modern-cpp-features/tree/master\#decltype}
                    {decltype}}
\highlighttt{decltype} inspects the declared type of an entity or the 
type and value category of an expression.

Parentheses matter:
\begin{itemize}
    \item \texttt{decltype(entity)} --- if the operand is an unparenthesized 
          id-expression or class member access naming an entity, the result is 
          the \emph{declared} type of that entity (including references and 
          cv-qualifiers as declared).
    \item \texttt{decltype((expression))} --- if the operand is any other 
          expression (including a parenthesized id-expression), the result 
          follows the expression's value category:
          \begin{itemize}
              \item lvalue $\rightarrow$ \texttt{T\&}
              \item xvalue $\rightarrow$ \texttt{T\&\&}
              \item prvalue $\rightarrow$ \texttt{T}
          \end{itemize}
\end{itemize}

\begin{lstlisting}[language=C++]
int a = 1;
decltype(a) b = a;       // int  (declared type of a)
decltype((a)) c = a;     // int& (a is an lvalue expression)
const int& r = a;
decltype(r) d = a;       // const int&
decltype((r)) e = a;     // const int&  (still an lvalue)
\end{lstlisting}

See~\cite{cppreference-decltype}


%~ TYPE ALIASES
\subsubsection{\href{https://github.com/AnthonyCalandra/modern-cpp-features/tree/master\#type-aliases}
                    {type aliases}}
Semantically similar to using a typedef however, type aliases with using 
are easier to read and are compatible with templates.

\begin{lstlisting}[language=C++]
template <typename T>
using Vec = std::vector<T>;
Vec<int> v; // std::vector<int>

using String = std::string;
String s {"foo"};
\end{lstlisting}

%~ NULLPTR
\subsubsection{\href{https://github.com/AnthonyCalandra/modern-cpp-features/tree/master\#nullptr}
                    {nullptr}}
\href{https://en.cppreference.com/w/cpp/language/nullptr}{Cppreference documentation about \textbf{nullptr}}
\newline
The keyword \highlighttt{nullptr} denotes the pointer literal. It is a 
\highlighttt{prvalue} of type \highlighttt{std::nullptr\char`_t}. 
There exist implicit conversions from \texttt{nullptr} to null pointer 
value of any pointer type and any pointer to member type. 
Similar conversions exist for any null pointer constant, 
which includes values of type \highlighttt{std::nullptr\char`_t} as well 
as the macro NULL.

%~ STRONGLY TYPED ENUMS
\subsubsection{\href{https://github.com/AnthonyCalandra/modern-cpp-features/tree/master\#strongly-typed-enums}
                    {strongly-typed enums}}

Type-safe enums that solve a variety of problems with C-style enums including: 
implicit conversions, inability to specify the underlying type, scope pollution.
\begin{lstlisting}[language=C++]
// Specifying underlying type as `unsigned int`
enum class Color : unsigned int { Red = 0xff0000, Green = 0xff00, Blue = 0xff };
// `Red`/`Green` in `Alert` don't conflict with `Color`
enum class Alert : bool { Red, Green };
Color c = Color::Red;
\end{lstlisting}

%~ ATTRIBUTES
\subsubsection{\href{https://github.com/AnthonyCalandra/modern-cpp-features/tree/master\#attributes}
                    {attributes}}
Attributes provide a universal syntax over \texttt{\char`_\char`_attribute\char`_\char`_(...)}, 
\texttt{\char`_\char`_declspec}, etc.
\begin{lstlisting}[language=C++]
// `noreturn` attribute indicates `f` doesn't return.
[[ noreturn ]] void f() {
  throw "error";
}
\end{lstlisting}

%~ CONSTEXPR
\subsubsection{\href{https://github.com/AnthonyCalandra/modern-cpp-features/tree/master\#constexpr}
                    {constexpr}}
\href{https://en.cppreference.com/w/cpp/language/constexpr}{Cppreference about \texttt{constexpr}}
\newline
The constexpr specifier declares that it is possible to evaluate 
the value of the function or variable at compile time. Such variables 
and functions can then be used where only compile time constant 
expressions are allowed (provided that appropriate function arguments 
are given). 


%~ DELEGATING CONSTRUCTORS
\subsubsection{\href{https://github.com/AnthonyCalandra/modern-cpp-features/tree/master\#delegating-constructors}
                    {delegating constructors}}
Constructors can now call other constructors in the same class using an initializer list.
\begin{lstlisting}[language=C++]
struct Foo {
  int foo;
  Foo(int foo) : foo{foo} {}
  Foo() : Foo(0) {}
};

Foo foo;
foo.foo; // == 0
\end{lstlisting}

%~ USER-DEFINED LITERALS
\subsubsection{\href{https://github.com/AnthonyCalandra/modern-cpp-features/tree/master\#user-defined-literals}
                    {user-defined literals}}
User-defined literals allow you to extend the language and add your own 
syntax. To create a literal, define a \texttt{T operator "" X(\dots) \{ \dots \}} 
function that returns a type \texttt{T}, with a name \texttt{X}. Note that 
the name of this function defines the name of the literal. 
Any literal names not starting with an underscore are reserved and won't 
be invoked. There are rules on what parameters a user-defined literal 
function should accept, according to what type the literal is called on.
\newline
\highlight{Converting Celsius to Fahrenheit}
\begin{lstlisting}[language=C++]
// `unsigned long long` parameter required for integer literal.
long long operator "" _celsius(unsigned long long tempCelsius) {
  return std::llround(tempCelsius * 1.8 + 32);
}
24_celsius; // == 75
\end{lstlisting}
\highlight{String to integer conversion}
\begin{lstlisting}[language=C++]
// `const char*` and `std::size_t` required as parameters.
int operator "" _int(const char* str, std::size_t) {
  return std::stoi(str);
}

"123"_int; // == 123, with type `int`
\end{lstlisting}

%~ EXPLICIT VIRTUAL OVERRIDES
\subsubsection{\href{https://github.com/AnthonyCalandra/modern-cpp-features/tree/master\#explicit-virtual-overrides}
                    {explicit virtual overrides}}
Specifies that a virtual function overrides another virtual function. 
If the virtual function does not override a parent's virtual function, 
throws a compiler error.

\begin{lstlisting}[language=C++]
struct A {
  virtual void foo();
  void bar();
};

struct B : A {
  void foo() override; // correct -- B::foo overrides A::foo
  void bar() override; // error -- A::bar is not virtual
  void baz() override; // error -- B::baz does not override A::baz
};
\end{lstlisting}

%~ FINAL SPECIFIER
\subsubsection{\href{https://github.com/AnthonyCalandra/modern-cpp-features/tree/master\#final-specifier}
                    {final specifier}}
Specifies that a virtual function cannot be overridden in a derived class 
or that a class cannot be inherited from.
\begin{lstlisting}[language=C++]
struct A {
  virtual void foo();
};

struct B : A {
  virtual void foo() final;
};

struct C : B {
  virtual void foo(); // error -- declaration of 'foo' overrides a 'final' function
};
\end{lstlisting}
Class cannot be inherited from.
\begin{lstlisting}[language=C++]
struct A final {};
struct B : A {}; // error -- base 'A' is marked 'final'
\end{lstlisting}

%~ DEFAULT FUNCTIONS
\subsubsection{\href{https://github.com/AnthonyCalandra/modern-cpp-features/tree/master\#default-functions}
                    {default functions}}
A more elegant, efficient way to provide a default implementation of a function, such as a constructor.

\begin{lstlisting}[language=C++]
struct A {
  A() = default;
  A(int x) : x{x} {}
  int x {1};
};
A a; // a.x == 1
A a2 {123}; // a.x == 123
\end{lstlisting}
With inheritance:
\begin{lstlisting}[language=C++]
struct B {
  B() : x{1} {}
  int x;
};

struct C : B {
  // Calls B::B
  C() = default;
};

C c; // c.x == 1
\end{lstlisting}

%~ DELETED FUNCTIONS
\subsubsection{\href{https://github.com/AnthonyCalandra/modern-cpp-features/tree/master\#deleted-functions}
                    {deleted functions}}
A more elegant, efficient way to provide a deleted implementation of a function. 
Useful for preventing copies on objects.
\begin{lstlisting}[language=C++]
class A {
  int x;

public:
  A(int x) : x{x} {};
  A(const A&) = delete;
  A& operator=(const A&) = delete;
};

A x {123};
A y = x; // error -- call to deleted copy constructor
y = x; // error -- operator= deleted
\end{lstlisting}

%~ RANGE-BASED FOR LOOPS
\subsubsection{\href{https://github.com/AnthonyCalandra/modern-cpp-features/tree/master\#range-based-for-loops}
                    {range-based for loops}}
A \highlight{\texttt{range-for} statement} applies to all standard 
containers and string as well, providing a less-typing way to 
\texttt{for}-loop over a \textit{range} of elements.
\newline
An example of usage in the following code
\begin{lstlisting}[language=C++]
std::array<int, 5> a {1, 2, 3, 4, 5};
for (int& x : a) x *= 2;
// a == { 2, 4, 6, 8, 10 }

std::array<int, 5> a {1, 2, 3, 4, 5};
for (int x : a) x *= 2;
// a == { 1, 2, 3, 4, 5 }
\end{lstlisting}
In the previous loops, note the difference when using \texttt{int} and \texttt{int\&}.

%~ SPECIAL MEMBER FUNCTIONS FOR MOVE SEMANTICS
\subsubsection{\href{https://github.com/AnthonyCalandra/modern-cpp-features/tree/master\#special-member-functions-for-move-semantics}
                    {special member functions for move semantics}}
The copy constructor and copy assignment operator are called when copies 
are made, and with C++11's introduction of move semantics, there is now 
a move constructor and move assignment operator for moves.

\begin{lstlisting}[language=C++]
struct A {
  std::string s;
  A() : s{"test"} {}
  A(const A& o) : s{o.s} {}
  A(A&& o) : s{std::move(o.s)} {}
  A& operator=(A&& o) {
   s = std::move(o.s);
   return *this;
  }
};

A f(A a) {
  return a;
}

A a1 = f(A{}); // move-constructed from rvalue temporary
A a2 = std::move(a1); // move-constructed using std::move
A a3 = A{};
a2 = std::move(a3); // move-assignment using std::move
a1 = f(A{}); // move-assignment from rvalue temporary
\end{lstlisting}

%~ CONVERTING CONSTRUCTORS
\subsubsection{\href{https://github.com/AnthonyCalandra/modern-cpp-features/tree/master\#converting-constructors}
                    {converting constructors}}
Converting constructors will convert values of braced list syntax into 
constructor arguments.

\begin{lstlisting}[language=C++]
struct A {
  A(int) {}
  A(int, int) {}
  A(int, int, int) {}
};

A a {0, 0}; // calls A::A(int, int)
A b(0, 0); // calls A::A(int, int)
A c = {0, 0}; // calls A::A(int, int)
A d {0, 0, 0}; // calls A::A(int, int, int)
\end{lstlisting}

Note that the braced list syntax does not allow narrowing:

\begin{lstlisting}[language=C++]
struct A {
  A(int) {}
};

A a(1.1); // OK
A b {1.1}; // Error narrowing conversion from double to int
\end{lstlisting}

Note that if a constructor accepts a \texttt{std::initializer\char`_list}, it will be 
called instead:
\begin{lstlisting}[language=C++]
struct A {
  A(int) {}
  A(int, int) {}
  A(int, int, int) {}
  A(std::initializer_list<int>) {}
};

A a {0, 0}; // calls A::A(std::initializer_list<int>)
A b(0, 0); // calls A::A(int, int)
A c = {0, 0}; // calls A::A(std::initializer_list<int>)
A d {0, 0, 0}; // calls A::A(std::initializer_list<int>)
\end{lstlisting}

%~ EXPLICIT CONVERSION FUNCTIONS
\subsubsection{\href{https://github.com/AnthonyCalandra/modern-cpp-features/tree/master\#explicit-conversion-functions}
                    {explicit conversion functions}}
Conversion functions can now be made explicit using the explicit specifier.

\begin{lstlisting}[language=C++]
struct A {
  operator bool() const { return true; }
};

struct B {
  explicit operator bool() const { return true; }
};

A a;
if (a); // OK calls A::operator bool()
bool ba = a; // OK copy-initialization selects A::operator bool()

B b;
if (b); // OK calls B::operator bool()
bool bb = b; // error copy-initialization does not consider B::operator bool()
\end{lstlisting}

%~ INLINE NAMESPACES
\subsubsection{\href{https://github.com/AnthonyCalandra/modern-cpp-features/tree/master\#inline-namespaces}
                    {inline-namespaces}}
All members of an inline namespace are treated as if they were part of its parent namespace, allowing specialization of functions and easing the process of versioning. This is a transitive property, if A contains B, which in turn contains C and both B and C are inline namespaces, C's members can be used as if they were on A.
\begin{lstlisting}[language=C++]
namespace Program {
  namespace Version1 {
    int getVersion() { return 1; }
    bool isFirstVersion() { return true; }
  }
  inline namespace Version2 {
    int getVersion() { return 2; }
  }
}

int version {Program::getVersion()};              // Uses getVersion() from Version2
int oldVersion {Program::Version1::getVersion()}; // Uses getVersion() from Version1
bool firstVersion {Program::isFirstVersion()};    // Does not compile when Version2 is added
\end{lstlisting}
%~ NON-STATIC DATA MEMBER INITIALIZERS
\subsubsection{\href{https://github.com/AnthonyCalandra/modern-cpp-features/tree/master\#non-static-data-member-initializers}
                    {non-static data member initializers}}
Allows non-static data members to be initialized where they are declared, potentially cleaning up constructors of default initializations.

\begin{lstlisting}[language=C++]
// Default initialization prior to C++11
class Human {
    Human() : age{0} {}
  private:
    unsigned age;
};
// Default initialization on C++11
class Human {
  private:
    unsigned age {0};
};
\end{lstlisting}

\subsubsection{\href{https://github.com/AnthonyCalandra/modern-cpp-features/tree/master\#right-angle-brackets}
                    {right angle brackets}}
C++11 is now able to infer when a series of right angle brackets is used 
as an operator or as a closing statement of typedef, without having to add whitespace.
\begin{lstlisting}[language=C++]
typedef std::map<int, std::map <int, std::map <int, int> > > cpp98LongTypedef;
typedef std::map<int, std::map <int, std::map <int, int>>>   cpp11LongTypedef;

\end{lstlisting}

%~ REF-QUALIFIED MEMBER FUNCTIONS
\subsubsection{\href{https://github.com/AnthonyCalandra/modern-cpp-features/tree/master\#ref-qualified-member-functions}
                    {ref-qualified member functions}}
Member functions can now be qualified depending on whether *this is an lvalue or rvalue reference.

\begin{lstlisting}[language=C++]
struct Bar {
  // ...
};

struct Foo {
  Bar getBar() & { return bar; }
  Bar getBar() const& { return bar; }
  Bar getBar() && { return std::move(bar); }
private:
  Bar bar;
};

Foo foo{};
Bar bar = foo.getBar(); // calls `Bar getBar() &`

const Foo foo2{};
Bar bar2 = foo2.getBar(); // calls `Bar Foo::getBar() const&`

Foo{}.getBar(); // calls `Bar Foo::getBar() &&`
std::move(foo).getBar(); // calls `Bar Foo::getBar() &&`

std::move(foo2).getBar(); // calls `Bar Foo::getBar() const&&`
\end{lstlisting}

%~ TRAILING RETURN TYPE
\subsubsection{\href{https://github.com/AnthonyCalandra/modern-cpp-features/tree/master\#trailing-return-types}
                    {trailing return types}}
C++11 allows functions and lambdas an alternative syntax for specifying their return types.
\begin{lstlisting}[language=C++]
int f() {
  return 123;
}
// vs.
auto f() -> int {
  return 123;
}

auto g = []() -> int {
  return 123;
};
\end{lstlisting}
This feature is especially useful when certain return types cannot be resolved:

\begin{lstlisting}[language=C++]
// NOTE: This does not compile!
template <typename T, typename U>
decltype(a + b) add(T a, U b) {
    return a + b;
}

// Trailing return types allows this:
template <typename T, typename U>
auto add(T a, U b) -> decltype(a + b) {
    return a + b;
}
\end{lstlisting}

In C++14, decltype(auto) (C++14) can be used instead.


%~ NOEXCEPT SPECIFIER
\subsubsection{\href{https://github.com/AnthonyCalandra/modern-cpp-features/tree/master\#noexcept-specifier}
                    {noexcept specifier}}
\href{https://en.cppreference.com/w/cpp/language/noexcept}{Cppreference about \texttt{noexcept}}
The noexcept specifier specifies whether a function could throw exceptions. It is an improved version of throw().
\begin{lstlisting}[language=C++]
void func1() noexcept;        // does not throw
void func2() noexcept(true);  // does not throw
void func3() throw();         // does not throw

void func4() noexcept(false); // may throw
\end{lstlisting}
Non-throwing functions are permitted to call potentially-throwing functions. Whenever an exception is thrown and the search for a handler encounters the outermost block of a non-throwing function, the function std::terminate is called.
\begin{lstlisting}[language=C++]
extern void f();  // potentially-throwing
void g() noexcept {
    f();          // valid, even if f throws
    throw 42;     // valid, effectively a call to std::terminate
}
\end{lstlisting}

%~ char32_t, char16_t
\subsubsection{\href{https://github.com/AnthonyCalandra/modern-cpp-features/tree/master\#char32_t-and-char16_t}
                    {char32\char`_t and char16\char`_t}}
Provides standard types for representing UTF-8 strings.
\begin{lstlisting}[language=C++]
char32_t utf8_str[] = U"\u0123";
char16_t utf8_str[] = u"\u0123";
\end{lstlisting}

%~ RAW STRING LITERALS
\subsubsection{\href{https://github.com/AnthonyCalandra/modern-cpp-features/tree/master\#raw-string-literals}
                    {raw string literals}}
C++11 introduces a new way to declare string literals as "raw string literals".
Characters issued from an escape sequence (tabs, line feeds, single backslashes, etc.) 
can be inputted raw while preserving formatting. This is useful, for example, 
to write literary text, which might contain a lot of quotes or special 
formatting. This can make your string literals easier to read and maintain.
\newline
A raw string literal is declared using the following syntax:
\begin{lstlisting}[language=C++]
R"delimiter(raw_characters)delimiter"
\end{lstlisting}
where:
\begin{itemize}
    \item delimiter is an optional sequence of characters made of any 
          source character except parentheses, backslashes and spaces.
    \item \texttt{raw\char`_characters} is any raw character sequence; must not contain 
          the closing sequence ")delimiter".
\end{itemize}

Example:
\begin{lstlisting}[language=C++]
// msg1 and msg2 are equivalent.
const char* msg1 = "\nHello,\n\tworld!\n";
const char* msg2 = R"(
Hello,
	world!
)";
\end{lstlisting}

\newpage
\subsection{C++11 new library features}

%~ MOVE
\subsubsection{\href{https://github.com/AnthonyCalandra/modern-cpp-features/tree/master\#stdmove}
                    {std::move}}
\texttt{std::move} indicates that the object passed to it may have its 
resources transferred. Using objects that have been moved from should 
be used with care, as they can be left in an unspecified state 
(see: \href{What can I do with a moved-from object?}{http://stackoverflow.com/questions/7027523/what-can-i-do-with-a-moved-from-object}).

A definition of \texttt{std::move} (performing a move is nothing more 
than casting to an rvalue reference):
\begin{lstlisting}[language=C++]
template <typename T>
typename remove_reference<T>::type&& move(T&& arg) {
  return static_cast<typename remove_reference<T>::type&&>(arg);
}
\end{lstlisting}
Transferring \texttt{std::unique\char`_ptrs}:
\begin{lstlisting}[language=C++]
std::unique_ptr<int> p1 {new int{0}};  // in practice, use std::make_unique
std::unique_ptr<int> p2 = p1; // error -- cannot copy unique pointers
std::unique_ptr<int> p3 = std::move(p1); // move `p1` into `p3`
                                         // now unsafe to dereference object held by `p1`
\end{lstlisting}

%~ FORWARD
\subsubsection{\href{https://github.com/AnthonyCalandra/modern-cpp-features/tree/master\#stdforward}
                    {std::forward}}
Returns the arguments passed to it while maintaining their value category 
and cv-qualifiers. Useful for generic code and factories. 
Used in conjunction with forwarding references.

A definition of \texttt{std::forward}:
\begin{lstlisting}[language=C++]
template <typename T>
T&& forward(typename remove_reference<T>::type& arg) {
  return static_cast<T&&>(arg);
}
\end{lstlisting}

An example of a function wrapper which just forwards other A objects to 
a new A object's copy or move constructor:
\begin{lstlisting}[language=C++]
struct A {
  A() = default;
  A(const A& o) { std::cout << "copied" << std::endl; }
  A(A&& o) { std::cout << "moved" << std::endl; }
};

template <typename T>
A wrapper(T&& arg) {
  return A{std::forward<T>(arg)};
}

wrapper(A{}); // moved
A a;
wrapper(a); // copied
wrapper(std::move(a)); // moved
\end{lstlisting}

See also: forwarding references, rvalue references.

%~ THREAD
\subsubsection{\href{https://github.com/AnthonyCalandra/modern-cpp-features/tree/master\#stdthread}
                    {std::thread}}
The \texttt{std::thread} library provides a standard way to control threads, 
such as spawning and killing them. In the example below, multiple threads 
are spawned to do different calculations and then the program waits 
for all of them to finish.
\begin{lstlisting}[language=C++]
void foo(bool clause) { /* do something... */ }

std::vector<std::thread> threadsVector;
threadsVector.emplace_back([]() {
  // Lambda function that will be invoked
});
threadsVector.emplace_back(foo, true);  // thread will run foo(true)
for (auto& thread : threadsVector) {
  thread.join(); // Wait for threads to finish
}
\end{lstlisting}

\subsubsection{\href{https://github.com/AnthonyCalandra/modern-cpp-features/tree/master\#stdto_string}
                    {std::to\char`_string}}
Converts a numeric argument to a std::string.
\begin{lstlisting}[language=C++]
std::to_string(1.2); // == "1.2"
std::to_string(123); // == "123"
\end{lstlisting}

%~ TYPE TRAITS
\subsubsection{\href{https://github.com/AnthonyCalandra/modern-cpp-features/tree/master\#type-traits}
                    {type traits}}
Type traits defines a compile-time template-based interface to query or modify the properties of types.
\begin{lstlisting}[language=C++]
static_assert(std::is_integral<int>::value);
static_assert(std::is_same<int, int>::value);
static_assert(std::is_same<std::conditional<true, int, double>::type, int>::value);
\end{lstlisting}

%~ SMART POINTERS
\subsubsection{\href{https://github.com/AnthonyCalandra/modern-cpp-features/tree/master\#smart-pointers}
                    {smart pointers}}
C++11 introduces new smart pointers: \texttt{std::unique\char`_ptr}, 
\texttt{std::shared\char`_ptr}, \texttt{std::weak\char`_ptr}. \texttt{std::auto\char`_ptr} 
now becomes deprecated and then eventually removed in C++17.

\texttt{std::unique\char`_ptr} is a non-copyable, movable pointer that 
manages its own heap-allocated memory. Note: Prefer using the \texttt{std::make\char`_X} 
helper functions as opposed to using constructors. 
See~\S\ref{sec:std-make-unique} (\texttt{std::make\char`_unique}) and 
\S\ref{sec:std-make-shared} (\texttt{std::make\char`_shared}).
\begin{lstlisting}[language=C++]
std::unique_ptr<Foo> p1 { new Foo{} };  // `p1` owns `Foo`
if (p1) {
  p1->bar();
}

{
  std::unique_ptr<Foo> p2 {std::move(p1)};  // Now `p2` owns `Foo`
  f(*p2);

  p1 = std::move(p2);  // Ownership returns to `p1` -- `p2` gets destroyed
}

if (p1) {
  p1->bar();
}
// `Foo` instance is destroyed when `p1` goes out of scope
\end{lstlisting}

A \texttt{std::shared\char`_ptr} is a smart pointer that manages a resource 
that is shared across multiple owners. A shared pointer holds a control 
block which has a few components such as the managed object and a 
reference counter. All control block access is thread-safe, however, 
manipulating the managed object itself is not thread-safe.
\begin{lstlisting}[language=C++]
void foo(std::shared_ptr<T> t) {
  // Do something with `t`...
}

void bar(std::shared_ptr<T> t) {
  // Do something with `t`...
}

void baz(std::shared_ptr<T> t) {
  // Do something with `t`...
}

std::shared_ptr<T> p1 {new T{}};
// Perhaps these take place in another threads?
foo(p1);
bar(p1);
baz(p1);
\end{lstlisting}

%~ CHRONO
\subsubsection{\href{https://github.com/AnthonyCalandra/modern-cpp-features/tree/master\#stdchrono}
                    {std::chrono}}
The chrono library contains a set of utility functions and types that 
deal with durations, clocks, and time points. One use case of this library 
is benchmarking code:
\begin{lstlisting}[language=C++]
std::chrono::time_point<std::chrono::steady_clock> start, end;
start = std::chrono::steady_clock::now();
// Some computations...
end = std::chrono::steady_clock::now();

std::chrono::duration<double> elapsed_seconds = end - start;
double t = elapsed_seconds.count(); // t number of seconds, represented as a `double`
\end{lstlisting}

%~ TUPLES
\subsubsection{\href{https://github.com/AnthonyCalandra/modern-cpp-features/tree/master\#tuples}
                    {tuples}}
Tuples are a fixed-size collection of heterogeneous values. 
Access the elements of a \texttt{std::tuple} by unpacking using \texttt{std::tie}, 
or using \texttt{std::get}.
\begin{lstlisting}[language=C++]
// `playerProfile` has type `std::tuple<int, const char*, const char*>`.
auto playerProfile = std::make_tuple(51, "Frans Nielsen", "NYI");
std::get<0>(playerProfile); // 51
std::get<1>(playerProfile); // "Frans Nielsen"
std::get<2>(playerProfile); // "NYI"
\end{lstlisting}

%~ TIE
\subsubsection{\href{https://github.com/AnthonyCalandra/modern-cpp-features/tree/master\#stdtie}
                    {std::tie}}
Creates a tuple of lvalue references. Useful for unpacking std::pair and std::tuple objects. Use std::ignore as a placeholder for ignored values. In C++17, structured bindings should be used instead.
\begin{lstlisting}[language=C++]
// With tuples...
std::string playerName;
std::tie(std::ignore, playerName, std::ignore) = std::make_tuple(91, "John Tavares", "NYI");

// With pairs...
std::string yes, no;
std::tie(yes, no) = std::make_pair("yes", "no");
\end{lstlisting}

%~ ARRAY
\subsubsection{\href{https://github.com/AnthonyCalandra/modern-cpp-features/tree/master\#stdarray}
                    {std::array}}
\texttt{std::array} is a container built on top of a C-style array. 
Supports common container operations such as sorting.
\begin{lstlisting}[language=C++]
std::array<int, 3> a = {2, 1, 3};
std::sort(a.begin(), a.end()); // a == { 1, 2, 3 }
for (int& x : a) x *= 2; // a == { 2, 4, 6 }
\end{lstlisting}

%~ UNORDERED CONTAINERS
\subsubsection{\href{https://github.com/AnthonyCalandra/modern-cpp-features/tree/master\#unordered-containers}
                    {unordered containers}}
These containers maintain average constant-time complexity for search, insert, and remove operations. In order to achieve constant-time complexity, sacrifices order for speed by hashing elements into buckets. There are four unordered containers:
\begin{itemize}
    \item \texttt{unordered\char`_set}
    \item \texttt{unordered\char`_multiset}
    \item \texttt{unordered\char`_map}
    \item \texttt{unordered\char`_multimap}
\end{itemize}


%~ MAKE SHARED
\subsubsection{\href{https://github.com/AnthonyCalandra/modern-cpp-features/tree/master\#stdmake_shared}
                    {std::make\char`_shared}}
\label{sec:std-make-shared}
\texttt{std::make\char`_shared} is the recommended way to create instances 
of \texttt{std::shared\char`_ptrs} due to the following reasons:
\begin{itemize}
    \item Avoid having to use the new operator.
    \item Prevents code repetition when specifying the underlying type the pointer shall hold.
    \item It provides exception-safety. Suppose we were calling a function foo like so:
\begin{lstlisting}[language=C++]
foo(std::shared_ptr<T>{new T{}}, function_that_throws(), std::shared_ptr<T>{new T{}});
\end{lstlisting}
The compiler is free to call \texttt{new T\{\}}, then 
\texttt{function\char`_that\char`_throws()}, and so on\dots \\
Since we have allocated data on the heap in the first construction of a T, 
we have introduced a leak here. With \texttt{std::make\char`_shared}, 
we are given exception-safety:
\begin{lstlisting}[language=C++]
foo(std::make_shared<T>(), function_that_throws(), std::make_shared<T>());
\end{lstlisting}

    \item Prevents having to do two allocations. When calling 
          \texttt{std::shared\char`_ptr\{ new T\{\} \}}, we have to allocate 
          memory for \texttt{T}, then in the shared pointer we have to 
          allocate memory for the control block within the pointer.
\end{itemize}

See the section on smart pointers for more information on 
\texttt{std::unique\char`_ptr} and \\ \texttt{std::shared\char`_ptr}.


%~ REF
\subsubsection{\href{https://github.com/AnthonyCalandra/modern-cpp-features/tree/master\#stdref}
                    {std::ref}}
\texttt{std::ref(val)} is used to create object of type 
\texttt{std::reference\char`_wrapper} that holds reference of val. 
Used in cases when usual reference passing using \texttt{\&} does not 
compile or \texttt{\&} is dropped due to type deduction. \texttt{std::cref} 
is similar but created reference wrapper holds a const reference to val.
\begin{lstlisting}[language=C++]
// create a container to store reference of objects.
auto val = 99;
auto _ref = std::ref(val);
_ref++;
auto _cref = std::cref(val);
//_cref++; does not compile
std::vector<std::reference_wrapper<int>>vec; // vector<int&>vec does not compile
vec.push_back(_ref); // vec.push_back(&i) does not compile
cout << val << endl; // prints 100
cout << vec[0] << endl; // prints 100
cout << _cref; // prints 100
\end{lstlisting}

%~ MEMORY MODEL
\subsubsection{\href{https://github.com/AnthonyCalandra/modern-cpp-features/tree/master\#memory-model}
                    {memory model}}
C++11 introduces a memory model for C++, which means library support for 
threading and atomic operations. Some of these operations include 
(but aren't limited to) atomic loads/stores, compare-and-swap, atomic flags, 
promises, futures, locks, and condition variables.

%~ ASYNC
\subsubsection{\href{https://github.com/AnthonyCalandra/modern-cpp-features/tree/master\#stdasync}
                    {std::async}}
\texttt{std::async} runs the given function either asynchronously or lazily-evaluated, 
then returns a \texttt{std::future} which holds the result of that function call.

The first parameter is the policy which can be:
\begin{itemize}
    \item \texttt{std::launch::async | std::launch::deferred} It is up to the implementation whether to perform asynchronous execution or lazy evaluation.
    \item \texttt{std::launch::async} Run the callable object on a new thread.
    \item \texttt{std::launch::deferred} Perform lazy evaluation on the current thread.
\end{itemize}
\begin{lstlisting}[language=C++]
int foo() {
  /* Do something here, then return the result. */
  return 1000;
}

auto handle = std::async(std::launch::async, foo);  // create an async task
auto result = handle.get();  // wait for the result
\end{lstlisting}

%~ BEGIN/END
\subsubsection{\href{https://github.com/AnthonyCalandra/modern-cpp-features/tree/master\#stdbeginend}
                    {std::begin/end}}
\texttt{std::begin} and \texttt{std::end} free functions were added to return begin and 
end iterators of a container generically. These functions also work with 
raw arrays which do not have begin and end member functions.
\begin{lstlisting}[language=C++]
template <typename T>
int CountTwos(const T& container) {
  return std::count_if(std::begin(container), std::end(container), [](int item) {
    return item == 2;
  });
}

std::vector<int> vec = {2, 2, 43, 435, 4543, 534};
int arr[8] = {2, 43, 45, 435, 32, 32, 32, 32};
auto a = CountTwos(vec); // 2
auto b = CountTwos(arr);  // 1
\end{lstlisting}

\newpage
\subsection{C++14 new language features}
Following a summary of the new language features introduced with C++14.
\subsubsection{\href{https://github.com/AnthonyCalandra/modern-cpp-features/tree/master\#binary-literals}
                    {binary literals}}
Binary literals provide a convenient way to represent a base-2 number.
It is possible to separate digits with '.
\begin{lstlisting}[language=C++]
0b110 // == 6
0b1111'1111 // == 255
\end{lstlisting}

\subsubsection{\href{https://github.com/AnthonyCalandra/modern-cpp-features/tree/master\#generic-lambda-expressions}
                    {generic lambda expressions}}
C++14 now allows the auto type-specifier in the parameter list, 
enabling polymorphic lambdas.
\begin{lstlisting}[language=C++]
auto identity = [](auto x) { return x; };
int three = identity(3); // == 3
std::string foo = identity("foo"); // == "foo"
\end{lstlisting}

\subsubsection{\href{https://github.com/AnthonyCalandra/modern-cpp-features/tree/master\#lambda-capture-initializers}
                    {lambda capture initializers}}
This allows creating lambda captures initialized with arbitrary expressions.
The name given to the captured value does not need to be related 
to any variables in the enclosing scopes and introduces a new name inside the lambda body. 
The initializing expression is evaluated when the lambda is created (not when it is invoked).
\begin{lstlisting}[language=C++]
int factory(int i) { return i * 10; }
auto f = [x = factory(2)] { return x; }; // returns 20

auto generator = [x = 0] () mutable {
  // this would not compile without 'mutable' as we are modifying x on each call
  return x++;
};
auto a = generator(); // == 0
auto b = generator(); // == 1
auto c = generator(); // == 2
\end{lstlisting}

Because it is now possible to move (or forward) values into a lambda that 
could previously be only captured by copy or reference we can now capture 
move-only types in a lambda by value. Note that in the below example 
the \texttt{p} in the capture-list of \texttt{task2} on the left-hand-side of 
\texttt{=} is a new variable private to the lambda body and does not refer to the original p.
\begin{lstlisting}[language=C++]
auto p = std::make_unique<int>(1);

auto task1 = [=] { *p = 5; }; // ERROR: std::unique_ptr cannot be copied
// vs.
auto task2 = [p = std::move(p)] { *p = 5; }; // OK: p is move-constructed into the closure object
// the original p is empty after task2 is created
\end{lstlisting}

Using this reference-captures can have different names than the referenced variable.
\begin{lstlisting}[language=C++]
auto x = 1;
auto f = [&r = x, x = x * 10] {
  ++r;
  return r + x;
};
f(); // sets x to 2 and returns 12
\end{lstlisting}

\subsubsection{\href{https://github.com/AnthonyCalandra/modern-cpp-features/tree/master\#return-type-deduction}
                    {return type deduction}}
Using an auto return type in C++14, the compiler will attempt to deduce 
the type for you. With lambdas, you can now deduce its return type using 
auto, which makes returning a deduced reference or rvalue reference possible.
\begin{lstlisting}[language=C++]
// Deduce return type as `int`.
auto f(int i) {
 return i;
}

template <typename T>
auto& f(T& t) {
  return t;
}

// Returns a reference to a deduced type.
auto g = [](auto& x) -> auto& { return f(x); };
int y = 123;
int& z = g(y); // reference to `y`
\end{lstlisting}

\subsubsection{\href{https://github.com/AnthonyCalandra/modern-cpp-features/tree/master\#decltypeauto}
                    {decltype(auto)}}
The \texttt{decltype(auto)} type-specifier also deduces a type like auto does. 
However, it deduces return types while keeping their references and 
cv-qualifiers, while auto will not.
\begin{lstlisting}[language=C++]
const int x = 0;
auto x1 = x; // int
decltype(auto) x2 = x; // const int
int y = 0;
int& y1 = y;
auto y2 = y1; // int
decltype(auto) y3 = y1; // int&
int&& z = 0;
auto z1 = std::move(z); // int
decltype(auto) z2 = std::move(z); // int&&

// Note: Especially useful for generic code!

// Return type is `int`.
auto f(const int& i) {
 return i;
}

// Return type is `const int&`.
decltype(auto) g(const int& i) {
 return i;
}

int x = 123;
static_assert(std::is_same<const int&, decltype(f(x))>::value == 0);
static_assert(std::is_same<int, decltype(f(x))>::value == 1);
static_assert(std::is_same<const int&, decltype(g(x))>::value == 1);
\end{lstlisting}


\subsubsection{\href{https://github.com/AnthonyCalandra/modern-cpp-features/tree/master\#relaxing-constraints-on-constexpr-functions}
                    {relaxing constraints on constexpr functions}}
In C++11, constexpr function bodies could only contain a very limited set of syntaxes, including (but not limited to): typedefs, usings, and a single return statement. In C++14, the set of allowable syntaxes expands greatly to include the most common syntax such as if statements, multiple returns, loops, etc.
\begin{lstlisting}[language=C++]
constexpr int factorial(int n) {
  if (n <= 1) {
    return 1;
  } else {
    return n * factorial(n - 1);
  }
}
factorial(5); // == 120
\end{lstlisting}

\subsubsection{\href{https://github.com/AnthonyCalandra/modern-cpp-features/tree/master\#variable-templates}
                    {variable templates}}
C++14 allows variables to be templated:
\begin{lstlisting}[language=C++]
template<class T>
constexpr T pi = T(3.1415926535897932385);
template<class T>
constexpr T e  = T(2.7182818284590452353);
\end{lstlisting}


\subsubsection{\href{https://github.com/AnthonyCalandra/modern-cpp-features/tree/master\#deprecated-attribute}
                    {\texttt{[[deprecated]]} attribute}}
C++14 introduces the \texttt{[[deprecated]]} attribute to indicate that a unit (function, class, etc.) is discouraged and likely yield compilation warnings. If a reason is provided, it will be included in the warnings.
\begin{lstlisting}[language=C++]
[[deprecated]]
void old_method();
[[deprecated("Use new_method instead")]]
void legacy_method();
\end{lstlisting}

\subsection{C++14 new library features}
Following a summary of the new library features introduced with C++14.
\subsubsection{\href{https://github.com/AnthonyCalandra/modern-cpp-features/tree/master\#user-defined-literals-for-standard-library-types}
                    {user-defined literals for standard library types}}
New user-defined literals for standard library types, including new built-in 
literals for chrono and \texttt{basic\char`_string}. These can be constexpr meaning 
they can be used at compile-time. Some uses for these literals include 
compile-time integer parsing, binary literals, and imaginary number literals.
\begin{lstlisting}[language=C++]
using namespace std::chrono_literals;
auto day = 24h;
day.count(); // == 24
std::chrono::duration_cast<std::chrono::minutes>(day).count(); // == 1440
\end{lstlisting}

\subsubsection{\href{https://github.com/AnthonyCalandra/modern-cpp-features/tree/master\#compile-time-integer-sequences}
                    {compile-time integer sequences}}
The class template \texttt{std::integer\char`_sequence} represents a 
compile-time sequence of integers. There are a few helpers built on top:

\begin{itemize}
    \item \texttt{std::make\char`_integer\char`_sequence<T, N>} - creates a sequence of 0, ..., N - 1 with type T.
    \item \texttt{std::index\char`_sequence\char`_for<T...>} - converts a template parameter pack into an integer sequence.
\end{itemize}

Convert an array into a tuple:
\begin{lstlisting}[language=C++]
template<typename Array, std::size_t... I>
decltype(auto) a2t_impl(const Array& a, std::integer_sequence<std::size_t, I...>) {
  return std::make_tuple(a[I]...);
}

template<typename T, std::size_t N, typename Indices = std::make_index_sequence<N>>
decltype(auto) a2t(const std::array<T, N>& a) {
  return a2t_impl(a, Indices());
}
\end{lstlisting}

\subsubsection{\href{https://github.com/AnthonyCalandra/modern-cpp-features/tree/master\#stdmake_unique}
                    {std::make\char`_unique}}
\label{sec:std-make-unique}
\texttt{std::make\char`_unique} is the recommended way to create instances 
of \texttt{std::unique\char`_ptrs} due to the following reasons:
\begin{itemize}
    \item Avoid having to use the new operator.
    \item Prevents code repetition when specifying the underlying type 
          the pointer shall hold.
    \item Most importantly, it provides exception-safety. Suppose we were 
          calling a function foo like so:
\begin{lstlisting}[language=C++]
foo(std::unique_ptr<T>{new T{}}}, function_that_throws(), 
    std::unique_ptr<T>{new T{}})};
\end{lstlisting}
\end{itemize}

The compiler is free to call \highlighttt{new T\{\}}, then \highlighttt{function\char`_that\char`_throws()}, 
and so on\dots{}
\newline
Since we have allocated data on the heap in the first 
construction of a T, we have introduced a leak here. 
With \highlighttt{std::make\char`_unique}, we are given exception-safety:

\begin{lstlisting}[language=C++]
foo(std::make_unique<T>(), function_that_throws(),
    std::make_unique<T>());
\end{lstlisting}
See the section on smart pointers (C++11) for more information on 
\texttt{std::unique\char`_ptr} and \texttt{std::shared\char`_ptr}.

\subsection{C++17 new language features}
Following a summary of the new language features introduced with C++17.
\subsubsection{\href{https://github.com/AnthonyCalandra/modern-cpp-features/tree/master\#template-argument-deduction-for-class-templates}
                    {template argument deduction for class templates}}
Automatic template argument deduction much like how it's done for functions, 
but now including class constructors.
\begin{lstlisting}[language=C++]
template <typename T = float>
struct MyContainer {
  T val;
  MyContainer() : val{} {}
  MyContainer(T val) : val{val} {}
  // ...
};
MyContainer c1 {1}; // OK MyContainer<int>
MyContainer c2; // OK MyContainer<float>
\end{lstlisting}

\subsubsection{\href{https://github.com/AnthonyCalandra/modern-cpp-features/tree/master\#declaring-non-type-template-parameters-with-auto}
                    {declaring non-type template parameters with auto}}
Following the deduction rules of auto, while respecting the non-type 
template parameter list of allowable types[*], template arguments can be 
deduced from the types of its arguments:
\begin{lstlisting}[language=C++]
template <auto... seq>
struct my_integer_sequence {
  // Implementation here ...
};

// Explicitly pass type `int` as template argument.
auto seq = std::integer_sequence<int, 0, 1, 2>();
// Type is deduced to be `int`.
auto seq2 = my_integer_sequence<0, 1, 2>();
\end{lstlisting}

* - For example, you cannot use a double as a template parameter type, which also makes this an invalid deduction using auto.
\subsubsection{\href{https://github.com/AnthonyCalandra/modern-cpp-features/tree/master\#folding-expressions}
                    {folding expressions}}
A fold expression performs a fold of a template parameter pack over a binary operator.
\begin{itemize}
    \item An expression of the form \texttt{(... op e)} or \texttt{(e op ...)}, 
          where op is a fold-operator and e is an unexpanded parameter 
          pack, are called unary folds.
    \item An expression of the form \texttt{(e1 op ... op e2)}, where op 
          are fold-operators, is called a binary fold. Either \texttt{e1} 
          or \texttt{e2} is an unexpanded parameter pack, but not both.
\end{itemize}
\begin{lstlisting}[language=C++]
template <typename... Args>
bool logicalAnd(Args... args) {
    // Binary folding.
    return (true && ... && args);
}
bool b = true;
bool& b2 = b;
logicalAnd(b, b2, true); // == true

template <typename... Args>
auto sum(Args... args) {
    // Unary folding.
    return (... + args);
}
sum(1.0, 2.0f, 3); // == 6.0
\end{lstlisting}

\subsubsection{\href{https://github.com/AnthonyCalandra/modern-cpp-features/tree/master\#new-rules-for-auto-deduction-from-braced-init-list}
                    {new rules for auto deduction from braced-init-list}}
Changes to auto deduction when used with the uniform initialization syntax. 
Previously, \texttt{auto x \{3\}}; deduces a \texttt{std::initializer\char`_list<int>}, 
which now deduces to int.
\begin{lstlisting}[language=C++]

auto x1 {1, 2, 3}; // error: not a single element
auto x2 = {1, 2, 3}; // x2 is std::initializer_list<int>
auto x3 {3}; // x3 is int
auto x4 {3.0}; // x4 is double
\end{lstlisting}

\subsubsection{\href{https://github.com/AnthonyCalandra/modern-cpp-features/tree/master\#constexpr-lambda}
                    {constexpr lambda}}
Compile-time lambdas using constexpr.

\begin{lstlisting}[language=C++]
auto identity = [](int n) constexpr { return n; };
static_assert(identity(123) == 123);
\end{lstlisting}

\begin{lstlisting}[language=C++]
constexpr auto add = [](int x, int y) {
  auto L = [=] { return x; };
  auto R = [=] { return y; };
  return [=] { return L() + R(); };
};

static_assert(add(1, 2)() == 3);
\begin{lstlisting}[language=C++]

constexpr int addOne(int n) {
  return [n] { return n + 1; }();
}

static_assert(addOne(1) == 2);
\end{lstlisting}

\subsubsection{\href{https://github.com/AnthonyCalandra/modern-cpp-features/tree/master\#lambda-capture-this-by-value}
                    {lambda capture this by value}}
Capturing this in a lambda's environment was previously reference-only. 
An example of where this is problematic is asynchronous code using callbacks 
that require an object to be available, potentially past its lifetime. 
\texttt{*this} (C++17) will now make a copy of the current object, while 
\texttt{this} (C++11) continues to capture by reference.

\begin{lstlisting}[language=C++]
struct MyObj {
  int value {123};
  auto getValueCopy() {
    return [*this] { return value; };
  }
  auto getValueRef() {
    return [this] { return value; };
  }
};
MyObj mo;
auto valueCopy = mo.getValueCopy();
auto valueRef = mo.getValueRef();
mo.value = 321;
valueCopy(); // 123
valueRef(); // 321
\end{lstlisting}

\subsubsection{\href{https://github.com/AnthonyCalandra/modern-cpp-features/tree/master\#inline-variables}
                    {inline variables}}
The inline specifier can be applied to variables as well as to functions. 
A variable declared inline has the same semantics as a function declared inline.

\begin{lstlisting}[language=C++]
// Disassembly example using compiler explorer.
struct S { int x; };
inline S x1 = S{321}; // mov esi, dword ptr [x1]
                      // x1: .long 321

S x2 = S{123};        // mov eax, dword ptr [.L_ZZ4mainE2x2]
                      // mov dword ptr [rbp - 8], eax
                      // .L_ZZ4mainE2x2: .long 123
\end{lstlisting}

It can also be used to declare and define a static member variable, 
such that it does not need to be initialized in the source file.

\begin{lstlisting}[language=C++]
struct S {
  S() : id{count++} {}
  ~S() { count--; }
  int id;
  static inline int count{0}; // declare and initialize count to 0 within the class
};
\end{lstlisting}

\subsubsection{\href{https://github.com/AnthonyCalandra/modern-cpp-features/tree/master\#nested-namespaces}
                    {nested namespaces}}
Using the namespace resolution operator to create nested namespace definitions.

\begin{lstlisting}[language=C++]
namespace A {
  namespace B {
    namespace C {
      int i;
    }
  }
}
\end{lstlisting}

The code above can be written like this:

\begin{lstlisting}[language=C++]
namespace A::B::C {
  int i;
}
\end{lstlisting}

\subsubsection{\href{https://github.com/AnthonyCalandra/modern-cpp-features/tree/master\#structured-bindings}
                    {structured bindings}}
A proposal for de-structuring initialization, that would allow writing 
\texttt{auto [ x, y, z ] = expr;} where the type of expr was a tuple-like 
object, whose elements would be bound to the variables x, y, and z (which 
this construct declares). Tuple-like objects include \texttt{std::tuple}, 
\texttt{std::pair}, \texttt{std::array}, and aggregate structures.

\begin{lstlisting}[language=C++]
using Coordinate = std::pair<int, int>;
Coordinate origin() {
  return Coordinate{0, 0};
}

const auto [ x, y ] = origin();
x; // == 0
y; // == 0
\end{lstlisting}

\begin{lstlisting}[language=C++]
std::unordered_map<std::string, int> mapping {
  {"a", 1},
  {"b", 2},
  {"c", 3}
};

// Destructure by reference.
for (const auto& [key, value] : mapping) {
  // Do something with key and value
}
\end{lstlisting}

\subsubsection{\href{https://github.com/AnthonyCalandra/modern-cpp-features/tree/master\#selection-statements-with-initializer}
                    {selection statements with initializer}}
New versions of the if and switch statements which simplify common code 
patterns and help users keep scopes tight.

\begin{lstlisting}[language=C++]
{
  std::lock_guard<std::mutex> lk(mx);
  if (v.empty()) v.push_back(val);
}
// vs.
if (std::lock_guard<std::mutex> lk(mx); v.empty()) {
  v.push_back(val);
}
\end{lstlisting}

\begin{lstlisting}[language=C++]
Foo gadget(args);
switch (auto s = gadget.status()) {
  case OK: gadget.zip(); break;
  case Bad: throw BadFoo(s.message());
}
// vs.
switch (Foo gadget(args); auto s = gadget.status()) {
  case OK: gadget.zip(); break;
  case Bad: throw BadFoo(s.message());
}
\end{lstlisting}

\subsubsection{\href{https://github.com/AnthonyCalandra/modern-cpp-features/tree/master\#constexpr-if}
                    {constexpr if}}
Write code that is instantiated depending on a compile-time condition.

\begin{lstlisting}[language=C++]
template <typename T>
constexpr bool isIntegral() {
  if constexpr (std::is_integral<T>::value) {
    return true;
  } else {
    return false;
  }
}
static_assert(isIntegral<int>() == true);
static_assert(isIntegral<char>() == true);
static_assert(isIntegral<double>() == false);
struct S {};
static_assert(isIntegral<S>() == false);
\end{lstlisting}

\subsubsection{\href{https://github.com/AnthonyCalandra/modern-cpp-features/tree/master\#utf-8-character-literals}
                    {utf-8 character literals}}
A character literal that begins with u8 is a character literal of type char. 
The value of a UTF-8 character literal is equal to its ISO 10646 code point value.

\begin{lstlisting}[language=C++]
char x = u8'x';
\end{lstlisting}

\subsubsection{\href{https://github.com/AnthonyCalandra/modern-cpp-features/tree/master\#direct-list-initialization-of-enums}
                    {direct-list-initialization of enums}}
Enums can now be initialized using braced syntax.
\begin{lstlisting}[language=C++]

enum byte : unsigned char {};
byte b {0}; // OK
byte c {-1}; // ERROR
byte d = byte{1}; // OK
byte e = byte{256}; // ERROR
\end{lstlisting}

\subsubsection{\href{https://github.com/AnthonyCalandra/modern-cpp-features/tree/master\#fallthrough-nodiscard-maybe_unused-attributes}
                    {[[fallthrough]], [[nodiscard]], [[maybe\char`_unused]] attributes}}
C++17 introduces three new attributes:
\newline
\texttt{[[fallthrough]]}, \texttt{[[nodiscard]]} and \texttt{[[maybe\char`_unused]]}.
\begin{itemize}
    \item \texttt{[[fallthrough]]} indicates to the compiler that falling through in a 
                          switch statement is intended behavior. This attribute 
                          may only be used in a switch statement, and 
                          must be placed before the next case/default label.

\begin{lstlisting}[language=C++]
switch (n) {
  case 1: 
    // ...
    [[fallthrough]];
  case 2:
    // ...
    break;
  case 3:
    // ...
    [[fallthrough]];
  default:
    // ...
}
\end{lstlisting}

    \item \texttt{[[nodiscard]]} issues a warning when either a function or class 
                        has this attribute and its return value is discarded.

\begin{lstlisting}[language=C++]
[[nodiscard]] bool do_something() {
  return is_success; // true for success, false for failure
}

do_something(); // warning: ignoring return value of 'bool do_something()',
                // declared with attribute 'nodiscard'
\end{lstlisting}

\begin{lstlisting}[language=C++]
// Only issues a warning when `error_info` is returned by value.
struct [[nodiscard]] error_info {
  // ...
};

error_info do_something() {
  error_info ei;
  // ...
  return ei;
}

do_something(); // warning: ignoring returned value of type 'error_info',
                // declared with attribute 'nodiscard'
\end{lstlisting}

    \item \texttt{[[maybe\char`_unused]]} indicates to the compiler that a variable 
                                 or parameter might be unused and is intended.

\begin{lstlisting}[language=C++]
void my_callback(std::string msg, [[maybe_unused]] bool error) {
  // Don't care if `msg` is an error message, just log it.
  log(msg);
}
\end{lstlisting}
\end{itemize}
\subsubsection{\href{https://github.com/AnthonyCalandra/modern-cpp-features/tree/master\#__has_include}
                    {\char`_\char`_has\char`_include}}
\char`_\char`_has\char`_include (operand) operator may be used in 
\#if and \#elif expressions to check whether a header or source file 
(operand) is available for inclusion or not.

One use case of this would be using two libraries that work the same way, 
using the backup/experimental one if the preferred one is not found on the system.

\begin{lstlisting}[language=C++]
#ifdef __has_include
#  if __has_include(<optional>)
#    include <optional>
#    define have_optional 1
#  elif __has_include(<experimental/optional>)
#    include <experimental/optional>
#    define have_optional 1
#    define experimental_optional
#  else
#    define have_optional 0
#  endif
#endif
\end{lstlisting}

It can also be used to include headers existing under different names or 
locations on various platforms, without knowing which platform the program 
is running on, OpenGL headers are a good example for this which are located 
in OpenGL\ directory on macOS and GL\ on other platforms.

\begin{lstlisting}[language=C++]
#ifdef __has_include
#  if __has_include(<OpenGL/gl.h>)
#    include <OpenGL/gl.h>
#    include <OpenGL/glu.h>
#  elif __has_include(<GL/gl.h>)
#    include <GL/gl.h>
#    include <GL/glu.h>
#  else
#    error No suitable OpenGL headers found.
# endif
#endif
\end{lstlisting}

\subsubsection{\href{https://github.com/AnthonyCalandra/modern-cpp-features/tree/master\#class-template-argument-deduction}
                    {class template argument deduction}}
Class template argument deduction (CTAD) allows the compiler to deduce 
template arguments from constructor arguments.

\begin{lstlisting}[language=C++]
std::vector v{ 1, 2, 3 }; // deduces std::vector<int>

std::mutex mtx;
auto lck = std::lock_guard{ mtx }; // deduces to std::lock_guard<std::mutex>

auto p = new std::pair{ 1.0, 2.0 }; // deduces to std::pair<double, double>
\end{lstlisting}

For user-defined types, deduction guides can be used to guide the compiler 
how to deduce template arguments if applicable:

\begin{lstlisting}[language=C++]
template <typename T>
struct container {
  container(T t) {}

  template <typename Iter>
  container(Iter beg, Iter end);
};

// deduction guide
template <typename Iter>
container(Iter b, Iter e) -> container<typename std::iterator_traits<Iter>::value_type>;

container a{ 7 }; // OK: deduces container<int>

std::vector<double> v{ 1.0, 2.0, 3.0 };
auto b = container{ v.begin(), v.end() }; // OK: deduces container<double>

container c{ 5, 6 }; // ERROR: std::iterator_traits<int>::value_type is not a type
\end{lstlisting}


\subsection{C++17 new library features:}
Following a summary of the new library features introduced with C++17.
\subsubsection{\href{https://github.com/AnthonyCalandra/modern-cpp-features/tree/master\#stdvariant}
                    {std::variant}}
The class template std::variant represents a type-safe union. An instance of std::variant at any given time holds a value of one of its alternative types (it's also possible for it to be valueless).

\begin{lstlisting}[language=C++]
std::variant<int, double> v{ 12 };
std::get<int>(v); // == 12
std::get<0>(v); // == 12
v = 12.0;
std::get<double>(v); // == 12.0
std::get<1>(v); // == 12.0
\end{lstlisting}

\subsubsection{\href{https://github.com/AnthonyCalandra/modern-cpp-features/tree/master\#stdoptional}
                    {std::optional}}
The class template std::optional manages an optional contained value, i.e. a value that may or may not be present. A common use case for optional is the return value of a function that may fail.

\begin{lstlisting}[language=C++]
std::optional<std::string> create(bool b) {
  if (b) {
    return "Godzilla";
  } else {
    return {};
  }
}

create(false).value_or("empty"); // == "empty"
create(true).value(); // == "Godzilla"
// optional-returning factory functions are usable as conditions of while and if
if (auto str = create(true)) {
  // ...
}
\end{lstlisting}

\subsubsection{\href{https://github.com/AnthonyCalandra/modern-cpp-features/tree/master\#stdany}
                    {std::any}}
A type-safe container for single values of any type.

\begin{lstlisting}[language=C++]
std::any x {5};
x.has_value() // == true
std::any_cast<int>(x) // == 5
std::any_cast<int&>(x) = 10;
std::any_cast<int>(x) // == 10
\end{lstlisting}

\subsubsection{\href{https://github.com/AnthonyCalandra/modern-cpp-features/tree/master\#stdstring_view}
                    {std::string\char`_view}}
A non-owning reference to a string. Useful for providing an abstraction on top of strings (e.g. for parsing).

\begin{lstlisting}[language=C++]
// Regular strings.
std::string_view cppstr {"foo"};
// Wide strings.
std::wstring_view wcstr_v {L"baz"};
// Character arrays.
char array[3] = {'b', 'a', 'r'};
std::string_view array_v(array, std::size(array));
\end{lstlisting}

\begin{lstlisting}[language=C++]
std::string str {"   trim me"};
std::string_view v {str};
v.remove_prefix(std::min(v.find_first_not_of(" "), v.size()));
str; //  == "   trim me"
v; // == "trim me"
\end{lstlisting}

\subsubsection{\href{https://github.com/AnthonyCalandra/modern-cpp-features/tree/master\#stdinvoke}
                    {std::invoke}}
Invoke a Callable object with parameters. Examples of callable objects are std::function or lambdas; objects that can be called similarly to a regular function.

\begin{lstlisting}[language=C++]
template <typename Callable>
class Proxy {
  Callable c_;

public:
  Proxy(Callable c) : c_{ std::move(c) } {}

  template <typename... Args>
  decltype(auto) operator()(Args&&... args) {
    // ...
    return std::invoke(c_, std::forward<Args>(args)...);
  }
};

const auto add = [](int x, int y) { return x + y; };
Proxy p{ add };
p(1, 2); // == 3
\end{lstlisting}

\subsubsection{\href{https://github.com/AnthonyCalandra/modern-cpp-features/tree/master\#stdapply}
                    {std::apply}}
Invoke a Callable object with a tuple of arguments.

\begin{lstlisting}[language=C++]
  auto add = [](int x, int y) {
  return x + y;
};
std::apply(add, std::make_tuple(1, 2)); // == 3
\end{lstlisting}

\subsubsection{\href{https://github.com/AnthonyCalandra/modern-cpp-features/tree/master\#stdfilesystem}
                    {std::filesystem}}
The new std::filesystem library provides a standard way to manipulate files, directories, and paths in a filesystem.

Here, a big file is copied to a temporary path if there is available space:

\begin{lstlisting}[language=C++]
const auto bigFilePath {"bigFileToCopy"};
if (std::filesystem::exists(bigFilePath)) {
  const auto bigFileSize {std::filesystem::file_size(bigFilePath)};
  std::filesystem::path tmpPath {"/tmp"};
  if (std::filesystem::space(tmpPath).available > bigFileSize) {
    std::filesystem::create_directory(tmpPath.append("example"));
    std::filesystem::copy_file(bigFilePath, tmpPath.append("newFile"));
  }
}
\end{lstlisting}

\subsubsection{\href{https://github.com/AnthonyCalandra/modern-cpp-features/tree/master\#stdbyte}
                    {std::byte}}
The new std::byte type provides a standard way of representing data as a byte. Benefits of using std::byte over char or unsigned char is that it is not a character type, and is also not an arithmetic type; while the only operator overloads available are bitwise operations.

s\begin{lstlisting}[language=C++]
td::byte a {0};
std::byte b {0xFF};
int i = std::to_integer<int>(b); // 0xFF
std::byte c = a & b;
int j = std::to_integer<int>(c); // 0
\end{lstlisting}

Note that std::byte is simply an enum, and braced initialization of enums become possible thanks to direct-list-initialization of enums.

\subsubsection{\href{https://github.com/AnthonyCalandra/modern-cpp-features/tree/master\#splicing-for-maps-and-sets}
                    {splicing for maps and sets}}
Moving nodes and merging containers without the overhead of expensive copies, moves, or heap allocations/deallocations.

Moving elements from one map to another:

\begin{lstlisting}[language=C++]
std::map<int, string> src {{1, "one"}, {2, "two"}, {3, "buckle my shoe"}};
std::map<int, string> dst {{3, "three"}};
dst.insert(src.extract(src.find(1))); // Cheap remove and insert of { 1, "one" } from `src` to `dst`.
dst.insert(src.extract(2)); // Cheap remove and insert of { 2, "two" } from `src` to `dst`.
// dst == { { 1, "one" }, { 2, "two" }, { 3, "three" } };
\end{lstlisting}

Inserting an entire set:

\begin{lstlisting}[language=C++]
std::set<int> src {1, 3, 5};
std::set<int> dst {2, 4, 5};
dst.merge(src);
// src == { 5 }
// dst == { 1, 2, 3, 4, 5 }
\end{lstlisting}

Inserting elements which outlive the container:

\begin{lstlisting}[language=C++]
auto elementFactory() {
  std::set<...> s;
  s.emplace(...);
  return s.extract(s.begin());
}
s2.insert(elementFactory());
\end{lstlisting}

Changing the key of a map element:

\begin{lstlisting}[language=C++]
std::map<int, string> m {{1, "one"}, {2, "two"}, {3, "three"}};
auto e = m.extract(2);
e.key() = 4;
m.insert(std::move(e));
// m == { { 1, "one" }, { 3, "three" }, { 4, "two" } }
\end{lstlisting}

\subsubsection{\href{https://github.com/AnthonyCalandra/modern-cpp-features/tree/master\#parallel-algorithms}
                    {parallel algorithms}}
Many of the STL algorithms, such as the copy, find and sort methods, 
started to support the parallel execution policies: \texttt{seq}, \texttt{par} and 
\texttt{par\char`_unseq} which translate to "\texttt{sequentially}", 
"\texttt{parallel}" and "\texttt{parallel unsequenced}".

\begin{lstlisting}[language=C++]
std::vector<int> longVector;
// Find element using parallel execution policy
auto result1 = std::find(std::execution::par, std::begin(longVector), std::end(longVector), 2);
// Sort elements using sequential execution policy
auto result2 = std::sort(std::execution::seq, std::begin(longVector), std::end(longVector));
\end{lstlisting}

\subsubsection{\href{https://github.com/AnthonyCalandra/modern-cpp-features/tree/master\#stdsample}
                    {std::sample}}
Samples n elements in the given sequence (without replacement) where every element has an equal chance of being selected.

\begin{lstlisting}[language=C++]
const std::string ALLOWED_CHARS = "abcdefghijklmnopqrstuvwxyzABCDEFGHIJKLMNOPQRSTUVWXYZ0123456789";
std::string guid;
// Sample 5 characters from ALLOWED_CHARS.
std::sample(ALLOWED_CHARS.begin(), ALLOWED_CHARS.end(), std::back_inserter(guid),
  5, std::mt19937{ std::random_device{}() });

std::cout << guid; // e.g. G1fW2
\end{lstlisting}

\subsubsection{\href{https://github.com/AnthonyCalandra/modern-cpp-features/tree/master\#stdclamp}
                    {std::clamp}}
Clamp given value between a lower and upper bound.

\begin{lstlisting}[language=C++]
std::clamp(42, -1, 1); // == 1
std::clamp(-42, -1, 1); // == -1
std::clamp(0, -1, 1); // == 0

// `std::clamp` also accepts a custom comparator:
std::clamp(0, -1, 1, std::less<>{}); // == 0
\end{lstlisting}

\subsubsection{\href{https://github.com/AnthonyCalandra/modern-cpp-features/tree/master\#stdreduce}
                    {std::reduce}}
Fold over a given range of elements. Conceptually similar to std::accumulate, but std::reduce will perform the fold in parallel. Due to the fold being done in parallel, if you specify a binary operation, it is required to be associative and commutative. A given binary operation also should not change any element or invalidate any iterators within the given range.

The default binary operation is std::plus with an initial value of 0.

\begin{lstlisting}[language=C++]
const std::array<int, 3> a{ 1, 2, 3 };
std::reduce(std::cbegin(a), std::cend(a)); // == 6
// Using a custom binary op:
std::reduce(std::cbegin(a), std::cend(a), 1, std::multiplies<>{}); // == 6
\end{lstlisting}

Additionally you can specify transformations for reducers:

\begin{lstlisting}[language=C++]
std::transform_reduce(std::cbegin(a), std::cend(a), 0, std::plus<>{}, times_ten); // == 60

const std::array<int, 3> b{ 1, 2, 3 };
const auto product_times_ten = [](const auto a, const auto b) { return a * b * 10; };

std::transform_reduce(std::cbegin(a), std::cend(a), std::cbegin(b), 0, std::plus<>{}, product_times_ten); // == 140
\end{lstlisting}

\newpage
\subsubsection{\href{https://github.com/AnthonyCalandra/modern-cpp-features/tree/master\#prefix-sum-algorithms}
                    {prefix sum algorithms}}
Support for prefix sums (both inclusive and exclusive scans) along with transformations.

\begin{lstlisting}[language=C++]
const std::array<int, 3> a{ 1, 2, 3 };

std::inclusive_scan(std::cbegin(a), std::cend(a),
    std::ostream_iterator<int>{ std::cout, " " }, std::plus<>{}); // 1 3 6

std::exclusive_scan(std::cbegin(a), std::cend(a),
    std::ostream_iterator<int>{ std::cout, " " }, 0, std::plus<>{}); // 0 1 3

const auto times_ten = [](const auto n) { return n * 10; };

std::transform_inclusive_scan(std::cbegin(a), std::cend(a),
    std::ostream_iterator<int>{ std::cout, " " }, std::plus<>{}, times_ten); // 10 30 60

std::transform_exclusive_scan(std::cbegin(a), std::cend(a),
    std::ostream_iterator<int>{ std::cout, " " }, 0, std::plus<>{}, times_ten); // 0 10 30
\end{lstlisting}


\subsubsection{\href{https://github.com/AnthonyCalandra/modern-cpp-features/tree/master\#gcd-and-lcm}
                    {gcd and lcm}}
Greatest common divisor (GCD) and least common multiple (LCM).

\begin{lstlisting}[language=C++]
const int p = 9;
const int q = 3;
std::gcd(p, q); // == 3
std::lcm(p, q); // == 9
\end{lstlisting}

\subsubsection{\href{https://github.com/AnthonyCalandra/modern-cpp-features/tree/master\#stdnot_fn}
                    {std::not\char`_fn}}
Utility function that returns the negation of the result of the given function.

\begin{lstlisting}[language=C++]
const std::ostream_iterator<int> ostream_it{ std::cout, " " };
const auto is_even = [](const auto n) { return n % 2 == 0; };
std::vector<int> v{ 0, 1, 2, 3, 4 };

// Print all even numbers.
std::copy_if(std::cbegin(v), std::cend(v), ostream_it, is_even); // 0 2 4
// Print all odd (not even) numbers.
std::copy_if(std::cbegin(v), std::cend(v), ostream_it, std::not_fn(is_even)); // 1 3
\end{lstlisting}

\subsubsection{\href{https://github.com/AnthonyCalandra/modern-cpp-features/tree/master\#string-conversion-tofrom-numbers}
                    {string conversion to/from numbers}}
Convert integrals and floats to a string or vice-versa. Conversions are non-throwing, do not allocate, and are more secure than the equivalents from the C standard library.

Users are responsible for allocating enough storage required for \texttt{std::to\char`_chars}, or the function will fail by setting the error code object in its return value.

These functions allow you to optionally pass a base (defaults to base-10) or a format specifier for floating type input.

    \texttt{std::to\char`_chars} returns a (non-const) char pointer which is one-past-the-end of the string that the function wrote to inside the given buffer, and an error code object.
    \texttt{std::from\char`_chars} returns a const char pointer which on success is equal to the end pointer passed to the function, and an error code object.

Both error code objects returned from these functions are equal to the default-initialized error code object on success.

Convert the number 123 to a std::string:

\begin{lstlisting}[language=C++]
const int n = 123;

// Can use any container, string, array, etc.
std::string str;
str.resize(3); // hold enough storage for each digit of `n`

const auto [ ptr, ec ] = std::to_chars(str.data(), str.data() + str.size(), n);

if (ec == std::errc{}) { std::cout << str << std::endl; } // 123
else { /* handle failure */ }
\end{lstlisting}

Convert from a std::string with value "123" to an integer:

\begin{lstlisting}[language=C++]
const std::string str{ "123" };
int n;

const auto [ ptr, ec ] = std::from_chars(str.data(), str.data() + str.size(), n);

if (ec == std::errc{}) { std::cout << n << std::endl; } // 123
else { /* handle failure */ }
\end{lstlisting}

\subsubsection{\href{https://github.com/AnthonyCalandra/modern-cpp-features/tree/master\#rounding-functions-for-chrono-durations-and-timepoints}
                    {rounding functions for chrono durations and timepoints}}
Provides \texttt{abs}, \texttt{round}, \texttt{ceil}, and \texttt{floor} helper 
functions for \texttt{std::chrono::duration} and \texttt{std::chrono::time\char`_point}.
\begin{lstlisting}[language=C++]
std::chrono::milliseconds a{ -5500 };
std::chrono::milliseconds d = std::chrono::abs(a); // == 5500ms
std::chrono::round<seconds>(d); // == 6s
std::chrono::ceil<seconds>(d); // == 6s
std::chrono::floor<seconds>(d); // == 5s
\end{lstlisting}

\newpage
\subsection{C++20 new language features}
Following a summary of the new language features introduced with C++20.
\subsubsection{\href{https://github.com/AnthonyCalandra/modern-cpp-features/tree/master\#coroutines}
                    {coroutines}}
Coroutines are special functions that can have their execution suspended 
and resumed. To define a coroutine, the \texttt{co\char`_return}, \texttt{co\char`_await}, 
or \texttt{co\char`_yield} keywords must be present in the function's body. 
C++20's coroutines are stackless; unless optimized out by the compiler, 
their state is allocated on the heap.

An example of a coroutine is a generator function, which yields (i.e. generates) 
a value at each invocation:
\begin{lstlisting}[language=C++]
generator<int> range(int start, int end) {
  while (start < end) {
    co_yield start;
    start++;
  }

  // Implicit co_return at the end of this function:
  // co_return;
}

for (int n : range(0, 10)) {
  std::cout << n << std::endl;
}
\end{lstlisting}

The above range generator function generates values starting at start 
until end (exclusive), with each iteration step yielding the current 
value stored in start. The generator maintains its state across each 
invocation of range (in this case, the invocation is for each iteration 
in the for loop). \texttt{co\char`_yield} takes the given expression, yields (i.e. returns) 
its value, and suspends the coroutine at that point. Upon resuming, 
execution continues after the \texttt{co\char`_yield}.

Another example of a coroutine is a task, which is an asynchronous computation 
that is executed when the task is awaited:
\begin{lstlisting}[language=C++]
task<void> echo(socket s) {
  for (;;) {
    auto data = co_await s.async_read();
    co_await async_write(s, data);
  }

  // Implicit co_return at the end of this function:
  // co_return;
}
\end{lstlisting}
In this example, the \texttt{co\char`_await} keyword is introduced. 
This keyword takes an expression and suspends execution if the thing 
you're awaiting on (in this case, the read or write) is not ready, 
otherwise you continue execution. (Note that under the hood, 
\texttt{co\char`_yield} uses \texttt{co\char`_await}.)

Using a task to lazily evaluate a value:
\begin{lstlisting}[language=C++]
task<int> calculate_meaning_of_life() {
  co_return 42;
}

auto meaning_of_life = calculate_meaning_of_life();
// ...
co_await meaning_of_life; // == 42
\end{lstlisting}

Note: While these examples illustrate how to use coroutines at a basic level, 
there is lots more going on when the code is compiled. These examples are not
meant to be complete coverage of C++20's coroutines. Since the generator and
task classes are not provided by the standard library yet, I used the 
cppcoro library to compile these examples.

\subsubsection{\href{https://github.com/AnthonyCalandra/modern-cpp-features/tree/master\#concepts}
                    {concepts}}
\label{sec:cpp20-concepts}
Concepts are named compile-time predicates which constrain types. They take the following form:

\begin{lstlisting}[language=C++]
template < template-parameter-list >
concept concept-name = constraint-expression;
\end{lstlisting}

where constraint-expression evaluates to a constexpr Boolean. Constraints should model semantic requirements, such as whether a type is a numeric or hashable. A compiler error results if a given type does not satisfy the concept it's bound by (i.e. constraint-expression returns false). Because constraints are evaluated at compile-time, they can provide more meaningful error messages and runtime safety.

\begin{lstlisting}[language=C++]
// `T` is not limited by any constraints.
template <typename T>
concept always_satisfied = true;
// Limit `T` to integrals.
template <typename T>
concept integral = std::is_integral_v<T>;
// Limit `T` to both the `integral` constraint and signedness.
template <typename T>
concept signed_integral = integral<T> && std::is_signed_v<T>;
// Limit `T` to both the `integral` constraint and the negation of the `signed_integral` constraint.
template <typename T>
concept unsigned_integral = integral<T> && !signed_integral<T>;
\end{lstlisting}

There are a variety of syntactic forms for enforcing concepts:

\begin{lstlisting}[language=C++]
// Forms for function parameters:
// `T` is a constrained type template parameter.
template <my_concept T>
void f(T v);

// `T` is a constrained type template parameter.
template <typename T>
  requires my_concept<T>
void f(T v);

// `T` is a constrained type template parameter.
template <typename T>
void f(T v) requires my_concept<T>;

// `v` is a constrained deduced parameter.
void f(my_concept auto v);

// `v` is a constrained non-type template parameter.
template <my_concept auto v>
void g();

// Forms for auto-deduced variables:
// `foo` is a constrained auto-deduced value.
my_concept auto foo = ...;

// Forms for lambdas:
// `T` is a constrained type template parameter.
auto f = []<my_concept T> (T v) {
  // ...
};
// `T` is a constrained type template parameter.
auto f = []<typename T> requires my_concept<T> (T v) {
  // ...
};
// `T` is a constrained type template parameter.
auto f = []<typename T> (T v) requires my_concept<T> {
  // ...
};
// `v` is a constrained deduced parameter.
auto f = [](my_concept auto v) {
  // ...
};
// `v` is a constrained non-type template parameter.
auto g = []<my_concept auto v> () {
  // ...
};
\end{lstlisting}

The requires keyword is used either to start a requires clause or a requires expression:

\begin{lstlisting}[language=C++]
template <typename T>
  requires my_concept<T> // `requires` clause.
void f(T);

template <typename T>
concept callable = requires (T f) { f(); }; // `requires` expression.

template <typename T>
  requires requires (T x) { x + x; } // `requires` clause and expression on same line.
T add(T a, T b) {
  return a + b;
}
\end{lstlisting}

Note that the parameter list in a requires expression is optional. Each requirement in a requires expression is one of the following:
\begin{itemize}
    \item Simple requirements --- asserts that the given expression is valid.
\begin{lstlisting}[language=C++]
template <typename T>
concept callable = requires (T f) { f(); };
\end{lstlisting}

    \item Type requirements --- denoted by the \texttt{typename} keyword followed by a type name, asserts that the given type name is valid.
\begin{lstlisting}[language=C++]
struct foo {
  int foo;
};

struct bar {
  using value = int;
  value data;
};

struct baz {
  using value = int;
  value data;
};

// Using SFINAE, enable if `T` is a `baz`.
template <typename T, typename = std::enable_if_t<std::is_same_v<T, baz>>>
struct S {};

template <typename T>
using Ref = T&;

template <typename T>
concept C = requires {
                     // Requirements on type `T`:
  typename T::value; // A) has an inner member named `value`
  typename S<T>;     // B) must have a valid class template specialization for `S`
  typename Ref<T>;   // C) must be a valid alias template substitution
};

template <C T>
void g(T a);

g(foo{}); // ERROR: Fails requirement A.
g(bar{}); // ERROR: Fails requirement B.
g(baz{}); // PASS.
\end{lstlisting}

    \item Compound requirements --- an expression in braces followed by a trailing return type or type constraint.
\begin{lstlisting}[language=C++]
template <typename T>
concept C = requires(T x) {
  {*x} -> std::convertible_to<typename T::inner>; // the type of the expression `*x` is convertible to `T::inner`
  {x + 1} -> std::same_as<int>; // the expression `x + 1` satisfies `std::same_as<decltype((x + 1))>`
  {x * 1} -> std::convertible_to<T>; // the type of the expression `x * 1` is convertible to `T`
};
\end{lstlisting}

    \item Nested requirements --- denoted by the \texttt{requires} keyword, specify additional constraints (such as those on local parameter arguments).
\begin{lstlisting}[language=C++]
template <typename T>
concept C = requires(T x) {
  requires std::same_as<sizeof(x), size_t>;
};
\end{lstlisting}
\end{itemize}

See also~\S\ref{sec:cpp20-concepts-library} (concepts library).

%~ THREE-WAY COMPARISON
\subsubsection{\href{https://github.com/AnthonyCalandra/modern-cpp-features/tree/master\#three-way-comparison}
                    {three-way comparison}}
C++20 introduces the spaceship operator (\texttt{<=>}) as a way to write comparison 
functions that reduce boilerplate and help define clearer comparison semantics. 
Defining a three-way comparison operator can autogenerate the other comparison 
operators (\texttt{==}, \texttt{!=}, \texttt{<}, etc.).

Three orderings are introduced:
\begin{itemize}
    \item \texttt{std::strong\char`_ordering}: distinguishes items being equal 
          (identical and interchangeable). Provides \texttt{less}, \texttt{greater}, 
          \texttt{equivalent}, and \texttt{equal}. Examples: integers, case-sensitive strings.
    \item \texttt{std::weak\char`_ordering}: distinguishes items being equivalent 
          (not identical, but interchangeable for comparison). Provides \texttt{less}, 
          \texttt{greater}, and \texttt{equivalent}. Examples: case-insensitive strings.
    \item \texttt{std::partial\char`_ordering}: like weak ordering but includes the 
          case when an ordering isn't possible. Provides \texttt{less}, \texttt{greater}, 
          \texttt{equivalent}, and \texttt{unordered}. Examples: floating-point values (e.g.\ \texttt{NaN}).
\end{itemize}

A defaulted three-way comparison operator does a member-wise comparison:
\begin{lstlisting}[language=C++]
struct foo {
  int a;
  bool b;
  char c;

  // Compare `a` first, then `b`, then `c` ...
  friend auto operator<=>(const foo&) const = default;
};

foo f1{0, false, 'a'}, f2{0, true, 'b'};
f1 < f2; // == true
f1 == f2; // == false
f1 >= f2; // == false
\end{lstlisting}

You can also define your own comparisons:
\begin{lstlisting}[language=C++]
struct foo {
  int x;
  bool b;
  char c;

  friend std::strong_ordering operator<=>(const foo& other) const {
      return x <=> other.x;
  }
};

foo f1{0, false, 'a'}, f2{0, true, 'b'};
f1 < f2; // == false
f1 == f2; // == true
f1 >= f2; // == true
\end{lstlisting}

\subsubsection{\href{https://github.com/AnthonyCalandra/modern-cpp-features/tree/master\#designated-initializers}
                    {designated initializers}}
C-style designated initializer syntax. Any member fields that are not explicitly listed in the designated initializer list are default-initialized.

\begin{lstlisting}[language=C++]
struct A {
  int x;
  int y;
  int z = 123;
};

A a {.x = 1, .z = 2}; // a.x == 1, a.y == 0, a.z == 2
\end{lstlisting}

\subsubsection{\href{https://github.com/AnthonyCalandra/modern-cpp-features/tree/master\#template-syntax-for-lambdas}
                    {template syntax for lambdas}}
Use familiar template syntax in lambda expressions.

\begin{lstlisting}[language=C++]
auto f = []<typename T>(std::vector<T> v) {
  // ...
};
\end{lstlisting}

\subsubsection{\href{https://github.com/AnthonyCalandra/modern-cpp-features/tree/master\#range-based-for-loop-with-initializer}
                    {range-based for loop with initializer}}
This feature simplifies common code patterns, helps keep scopes tight, and offers an elegant solution to a common lifetime problem.

\begin{lstlisting}[language=C++]
for (auto v = std::vector{1, 2, 3}; auto& e : v) {
  std::cout << e;
}
// prints "123"
\end{lstlisting}

\subsubsection{\href{https://github.com/AnthonyCalandra/modern-cpp-features/tree/master\#likely-and-unlikely-attributes}
                    {[[likely]] and [[unlikely]] attributes}}
Provides a hint to the optimizer that the labelled statement has a high probability of being executed.

\begin{lstlisting}[language=C++]
switch (n) {
case 1:
  // ...
  break;

[[likely]] case 2:  // n == 2 is considered to be arbitrarily more
  // ...            // likely than any other value of n
  break;
}
\end{lstlisting}

If one of the likely/unlikely attributes appears after the right parenthesis of an if-statement, it indicates that the branch is likely/unlikely to have its substatement (body) executed.

\begin{lstlisting}[language=C++]
int random = get_random_number_between_x_and_y(0, 3);
if (random > 0) [[likely]] {
  // body of if statement
  // ...
}
\end{lstlisting}

It can also be applied to the substatement (body) of an iteration statement.

\begin{lstlisting}[language=C++]
while (unlikely_truthy_condition) [[unlikely]] {
  // body of while statement
  // ...
}
\end{lstlisting}

\subsubsection{\href{https://github.com/AnthonyCalandra/modern-cpp-features/tree/master\#deprecate-implicit-capture-of-this}
                    {deprecate implicit capture of this}}
Implicitly capturing this in a lambda capture using [=] is now deprecated; prefer capturing explicitly using [=, this] or [=, *this].

\begin{lstlisting}[language=C++]
struct int_value {
  int n = 0;
  auto getter_fn() {
    // BAD:
    // return [=]() { return n; };

    // GOOD:
    return [=, *this]() { return n; };
  }
};
\end{lstlisting}

\subsubsection{\href{https://github.com/AnthonyCalandra/modern-cpp-features/tree/master\#class-types-in-non-type-template-parameters}
                    {class types in non-type template parameters}}
Classes can now be used in non-type template parameters. Objects passed in as template arguments have the type const T, where T is the type of the object, and has static storage duration.

\begin{lstlisting}[language=C++]
struct foo {
  foo() = default;
  constexpr foo(int) {}
};

template <foo f>
auto get_foo() {
  return f;
}

get_foo(); // uses implicit constructor
get_foo<foo{123}>();
\end{lstlisting}

\subsubsection{\href{https://github.com/AnthonyCalandra/modern-cpp-features/tree/master\#constexpr-virtual-functions}
                    {constexpr virtual functions}}
Virtual functions can now be constexpr and evaluated at compile-time. constexpr virtual functions can override non-constexpr virtual functions and vice-versa.

\begin{lstlisting}[language=C++]
struct X1 {
  virtual int f() const = 0;
};

struct X2: public X1 {
  constexpr virtual int f() const { return 2; }
};

struct X3: public X2 {
  virtual int f() const { return 3; }
};

struct X4: public X3 {
  constexpr virtual int f() const { return 4; }
};

constexpr X4 x4;
x4.f(); // == 4
\end{lstlisting}

\subsubsection{\href{https://github.com/AnthonyCalandra/modern-cpp-features/tree/master\#explicitbool}
                    {explicit(bool)}}
Conditionally select at compile-time whether a constructor is made explicit or not. explicit(true) is the same as specifying explicit.

\begin{lstlisting}[language=C++]
struct foo {
  // Specify non-integral types (strings, floats, etc.) require explicit construction.
  template <typename T>
  explicit(!std::is_integral_v<T>) foo(T) {}
};

foo a = 123; // OK
foo b = "123"; // ERROR: explicit constructor is not a candidate (explicit specifier evaluates to true)
foo c {"123"}; // OK
\end{lstlisting}

\subsubsection{\href{https://github.com/AnthonyCalandra/modern-cpp-features/tree/master\#immediate-functions}
                    {immediate functions}}
Similar to constexpr functions, but functions with a consteval specifier must produce a constant. These are called immediate functions.

\begin{lstlisting}[language=C++]
consteval int sqr(int n) {
  return n * n;
}

constexpr int r = sqr(100); // OK
int x = 100;
int r2 = sqr(x); // ERROR: the value of 'x' is not usable in a constant expression
                 // OK if `sqr` were a `constexpr` function
\end{lstlisting}

\subsubsection{\href{https://github.com/AnthonyCalandra/modern-cpp-features/tree/master\#using-enum}
                    {using enum}}
Bring an enum's members into scope to improve readability. Before:

\begin{lstlisting}[language=C++]
enum class rgba_color_channel { red, green, blue, alpha };

std::string_view to_string(rgba_color_channel channel) {
  switch (channel) {
    case rgba_color_channel::red:   return "red";
    case rgba_color_channel::green: return "green";
    case rgba_color_channel::blue:  return "blue";
    case rgba_color_channel::alpha: return "alpha";
  }
}
\end{lstlisting}

After:

\begin{lstlisting}[language=C++]
enum class rgba_color_channel { red, green, blue, alpha };

std::string_view to_string(rgba_color_channel my_channel) {
  switch (my_channel) {
    using enum rgba_color_channel;
    case red:   return "red";
    case green: return "green";
    case blue:  return "blue";
    case alpha: return "alpha";
  }
}
\end{lstlisting}

\subsubsection{\href{https://github.com/AnthonyCalandra/modern-cpp-features/tree/master\#lambda-capture-of-parameter-pack}
                    {lambda capture of parameter pack}}
Capture parameter packs by value:

\begin{lstlisting}[language=C++]
template <typename... Args>
auto f(Args&&... args){
    // BY VALUE:
    return [...args = std::forward<Args>(args)] {
        // ...
    };
}
\end{lstlisting}

Capture parameter packs by reference:

\begin{lstlisting}[language=C++]
template <typename... Args>
auto f(Args&&... args){
    // BY REFERENCE:
    return [&...args = std::forward<Args>(args)] {
        // ...
    };
}
\end{lstlisting}

\subsubsection{\href{https://github.com/AnthonyCalandra/modern-cpp-features/tree/master\#char8_t}
                    {char8\char`_t}}
Provides a standard type for representing UTF-8 strings.

\begin{lstlisting}[language=C++]
char8_t utf8_str[] = u8"\u0123";
\end{lstlisting}

\subsubsection{\href{https://github.com/AnthonyCalandra/modern-cpp-features/tree/master\#constinit}
                    {constinit}}
The constinit specifier requires that a variable must be initialized at compile-time.

\begin{lstlisting}[language=C++]
const char* g() { return "dynamic initialization"; }
constexpr const char* f(bool p) { return p ? "constant initializer" : g(); }

constinit const char* c = f(true); // OK
constinit const char* d = f(false); // ERROR: `g` is not constexpr, so `d` cannot be evaluated at compile-time.
\end{lstlisting}

\subsubsection{\href{https://github.com/AnthonyCalandra/modern-cpp-features/tree/master\#__va_opt__}
                    {\char`_\char`_VA\char`_OPT\char`_\char`_}}
Helps support variadic macros by evaluating to the given argument if the 
variadic macro is non-empty.
\begin{lstlisting}[language=C++]
#define F(...) f(0 __VA_OPT__(,) __VA_ARGS__)
F(a, b, c) // replaced by f(0, a, b, c)
F()        // replaced by f(0)
\end{lstlisting}

\subsection{C++20  new library features}
Following a summary of the new library features introduced with C++20.
\subsubsection{\href{https://github.com/AnthonyCalandra/modern-cpp-features/tree/master\#concepts-library}
                    {concepts library}}
\label{sec:cpp20-concepts-library}
Concepts are also provided by the standard library for building more complicated concepts. Some of these include:

Core language concepts:
  \begin{itemize}
    \item \texttt{same\char`_as} --- specifies two types are the same.
    \item \texttt{derived\char`_from} --- specifies that a type is derived from another type.
    \item \texttt{convertible\char`_to} --- specifies that a type is implicitly convertible to another type.
    \item \texttt{common\char`_with} --- specifies that two types share a common type.
    \item \texttt{integral} --- specifies that a type is an integral type.
    \item \texttt{default\char`_constructible} --- specifies that an object of a type can be default-constructed.
  \end{itemize}
Comparison concepts:
  \begin{itemize}
    \item \texttt{boolean} --- specifies that a type can be used in Boolean contexts.
    \item \texttt{equality\char`_comparable} --- specifies that \texttt{operator==} is an equivalence relation.
  \end{itemize}
Object concepts:
  \begin{itemize}
    \item \texttt{movable} --- specifies that an object of a type can be moved and swapped.
    \item \texttt{copyable} --- specifies that an object of a type can be copied, moved, and swapped.
    \item \texttt{semiregular} --- specifies that an object of a type can be copied, moved, swapped, and default constructed.
    \item \texttt{regular} --- specifies that a type is regular, that is, it is both \texttt{semiregular} and \texttt{equality\char`_comparable}.
  \end{itemize}
Callable concepts:
  \begin{itemize}
    \item \texttt{invocable} --- specifies that a callable type can be invoked with a given set of argument types.
    \item \texttt{predicate} --- specifies that a callable type is a Boolean predicate.
  \end{itemize}
See also~\S\ref{sec:cpp20-concepts} (concepts).

\subsubsection{\href{https://github.com/AnthonyCalandra/modern-cpp-features/tree/master\#synchronized-buffered-outputstream}
                    {synchronized buffered outputstream}}
Buffers output operations for the wrapped output stream ensuring synchronization (i.e. no interleaving of output).

\begin{lstlisting}[language=C++]
std::osyncstream{std::cout} << "The value of x is:" << x << std::endl;
\end{lstlisting}


\subsubsection{\href{https://github.com/AnthonyCalandra/modern-cpp-features/tree/master\#stdspan}
                    {std::span}}
A span is a view (i.e. non-owning) of a container providing bounds-checked access to a contiguous group of elements. Since views do not own their elements they are cheap to construct and copy -- a simplified way to think about views is they are holding references to their data. As opposed to maintaining a pointer/iterator and length field, a span wraps both of those up in a single object.

Spans can be dynamically-sized or fixed-sized (known as their extent). Fixed-sized spans benefit from bounds-checking.

Span doesn't propogate const so to construct a read-only span use std::span<const T>.

Example: using a dynamically-sized span to print integers from various containers.

\begin{lstlisting}[language=C++]
void print_ints(std::span<const int> ints) {
    for (const auto n : ints) {
        std::cout << n << std::endl;
    }
}

print_ints(std::vector{ 1, 2, 3 });
print_ints(std::array<int, 5>{ 1, 2, 3, 4, 5 });

int a[10] = { 0 };
print_ints(a);
// etc.
\end{lstlisting}

Example: a statically-sized span will fail to compile for containers that don't match the extent of the span.

\begin{lstlisting}[language=C++]
void print_three_ints(std::span<const int, 3> ints) {
    for (const auto n : ints) {
        std::cout << n << std::endl;
    }
}

print_three_ints(std::vector{ 1, 2, 3 }); // ERROR
print_three_ints(std::array<int, 5>{ 1, 2, 3, 4, 5 }); // ERROR
int a[10] = { 0 };
print_three_ints(a); // ERROR

std::array<int, 3> b = { 1, 2, 3 };
print_three_ints(b); // OK

// You can construct a span manually if required:
std::vector c{ 1, 2, 3 };
print_three_ints(std::span<const int, 3>{ c.data(), 3 }); // OK: set pointer and length field.
print_three_ints(std::span<const int, 3>{ c.cbegin(), c.cend() }); // OK: use iterator pairs.
\end{lstlisting}

\subsubsection{\href{https://github.com/AnthonyCalandra/modern-cpp-features/tree/master\#bit-operations}
                    {bit operations}}
C++20 provides a new <bit> header which provides some bit operations including popcount.

\begin{lstlisting}[language=C++]
std::popcount(0u); // 0
std::popcount(1u); // 1
std::popcount(0b1111'0000u); // 4
\end{lstlisting}

\subsubsection{\href{https://github.com/AnthonyCalandra/modern-cpp-features/tree/master\#math-constants}
                    {math constants}}
Mathematical constants including PI, Euler's number, etc. defined in the <numbers> header.

\begin{lstlisting}[language=C++]
std::numbers::pi; // 3.14159...
std::numbers::e; // 2.71828...

\subsubsection{\href{https://github.com/AnthonyCalandra/modern-cpp-features/tree/master\#stdis_constant_evaluated}
                    {std::is\char`_constant\char`_evaluated}}
Predicate function which is truthy when it is called in a compile-time context.

\begin{lstlisting}[language=C++]
constexpr bool is_compile_time() {
    return std::is_constant_evaluated();
}

constexpr bool a = is_compile_time(); // true
bool b = is_compile_time(); // false
\end{lstlisting}

\subsubsection{\href{https://github.com/AnthonyCalandra/modern-cpp-features/tree/master\#stdmake_shared-supports-arrays}
                    {std::make\char`_shared supports arrays}}


\begin{lstlisting}[language=C++]
auto p = std::make_shared<int[]>(5); // pointer to `int[5]`
// OR
auto p = std::make_shared<int[5]>(); // pointer to `int[5]`
\end{lstlisting}

\subsubsection{\href{https://github.com/AnthonyCalandra/modern-cpp-features/tree/master\#starts_with-and-ends_with-on-strings}
                    {starts\char`_with and ends\char`_with on strings}}
Strings (and string views) now have the starts\char`_with and ends\char`_with member functions to check if a string starts or ends with the given string.

\begin{lstlisting}[language=C++]
std::string str = "foobar";
str.starts_with("foo"); // true
str.ends_with("baz"); // false
\end{lstlisting}

\subsubsection{\href{https://github.com/AnthonyCalandra/modern-cpp-features/tree/master\#check-if-associative-container-has-element}
                    {check if associative container has element}}
Associative containers such as sets and maps have a contains member function, which can be used instead of the "find and check end of iterator" idiom.

\begin{lstlisting}[language=C++]
std::map<int, char> map {{1, 'a'}, {2, 'b'}};
map.contains(2); // true
map.contains(123); // false

std::set<int> set {1, 2, 3};
set.contains(2); // true
\end{lstlisting}

\subsubsection{\href{https://github.com/AnthonyCalandra/modern-cpp-features/tree/master\#stdbit_cast}
                    {std::bit\char`_cast}}
A safer way to reinterpret an object from one type to another.

\begin{lstlisting}[language=C++]
float f = 123.0;
int i = std::bit_cast<int>(f);
\end{lstlisting}

\subsubsection{\href{https://github.com/AnthonyCalandra/modern-cpp-features/tree/master\#stdmidpoint}
                    {std::midpoint}}
Calculate the midpoint of two integers safely (without overflow).

\begin{lstlisting}[language=C++]
std::midpoint(1, 3); // == 2
\end{lstlisting}

\subsubsection{\href{https://github.com/AnthonyCalandra/modern-cpp-features/tree/master\#stdto_array}
                    {std::to\char`_array}}
Converts the given array/"array-like" object to a std::array.

\begin{lstlisting}[language=C++]
std::to_array("foo"); // returns `std::array<char, 4>`
std::to_array<int>({1, 2, 3}); // returns `std::array<int, 3>`

int a[] = {1, 2, 3};
std::to_array(a); // returns `std::array<int, 3>`
\end{lstlisting}

\subsubsection{\href{https://github.com/AnthonyCalandra/modern-cpp-features/tree/master\#text-formatting}
                    {text formatting}}
Provides a compile-time checked string formatting library via \texttt{std::format}. 
Runtime formatting uses \texttt{std::vformat} and related helpers.

\begin{lstlisting}[language=C++]
std::format("{}", 123); // OK -- returns "123"
std::format("{} {}", 123); // ERROR -- not enough arguments
std::format("{} {}", "Here's a number:", 123); // OK
\end{lstlisting}

Formatting a string based on a formatter created at runtime:
\begin{lstlisting}[language=C++]
std::string fmt = "{} {}";
fmt += "{}{}";
std::vformat(fmt, std::make_format_args("Here's a number:", 1, 2, 3));
// OK -- returns "Here's a number: 123"
\end{lstlisting}

When formatting fails (such as an invalid format string), \texttt{std::vformat} 
throws \texttt{std::format\char`_error}.

To format custom types:
\begin{lstlisting}[language=C++]
struct fraction {
  int numerator;
  int denominator;
};

template <>
struct std::formatter<fraction> {
  constexpr auto parse(std::format_parse_context& ctx) {
    return ctx.begin();
  }

  auto format(const fraction& f, std::format_context& ctx) const {
    return std::format_to(ctx.out(), "{0:d}/{1:d}", f.numerator, f.denominator);
  }
};

fraction f{1, 2};
std::format("{}", f); // == "1/2"
\end{lstlisting}

\subsubsection{\href{https://github.com/AnthonyCalandra/modern-cpp-features/tree/master\#stdbind_front}
                    {std::bind\char`_front}}
Binds the first $N$ arguments to a given free function, lambda, or member function.
\begin{lstlisting}[language=C++]
const auto f = [](int a, int b, int c) { return a + b + c; };
const auto g = std::bind_front(f, 1, 1);
g(1); // == 3
\end{lstlisting}

\subsubsection{\href{https://github.com/AnthonyCalandra/modern-cpp-features/tree/master\#uniform-container-erasure}
                    {uniform container erasure}}
Provides \texttt{std::erase} and/or \texttt{std::erase\char`_if} for STL containers 
such as string, list, vector, map, etc. Both return the number of erased elements.
\begin{lstlisting}[language=C++]
std::vector v{0, 1, 0, 2, 0, 3};
std::erase(v, 0); // v == {1, 2, 3}
std::erase_if(v, [](int n) { return n == 0; }); // v == {1, 2, 3}
\end{lstlisting}

\subsubsection{\href{https://github.com/AnthonyCalandra/modern-cpp-features/tree/master\#three-way-comparison-helpers}
                    {three-way comparison helpers}}
Helper functions for naming comparison results:
\begin{lstlisting}[language=C++]
std::is_eq(0 <=> 0); // == true
std::is_lteq(0 <=> 1); // == true
std::is_gt(0 <=> 1); // == false
\end{lstlisting}

\subsubsection{\href{https://github.com/AnthonyCalandra/modern-cpp-features/tree/master\#stdlexicographical_compare_three_way}
                    {std::lexicographical\char`_compare\char`_three\char`_way}}
Lexicographically compares two ranges using three-way comparison and produces 
a result of the strongest applicable comparison category type.
\begin{lstlisting}[language=C++]
std::vector a{0, 0, 0}, b{0, 0, 0}, c{1, 1, 1};

auto cmp_ab = std::lexicographical_compare_three_way(
    a.begin(), a.end(), b.begin(), b.end());
std::is_eq(cmp_ab); // == true

auto cmp_ac = std::lexicographical_compare_three_way(
    a.begin(), a.end(), c.begin(), c.end());
std::is_lt(cmp_ac); // == true
\end{lstlisting}

\subsubsection{\href{https://github.com/AnthonyCalandra/modern-cpp-features/tree/master\#stdjthread}
                    {std::jthread}}
A thread of execution (like \texttt{std::thread}) that joins on destruction and 
can be signaled to stop. Unlike \texttt{std::thread}, you need not check 
\texttt{joinable()} before joining --- the destructor joins automatically.
Request a stop via \texttt{request\char`_stop} or the thread's \texttt{stop\char`_source}:
\begin{lstlisting}[language=C++]
std::jthread t{
    [](std::stop_token stoken) {
        while (!stoken.stop_requested()) {
            std::this_thread::sleep_for(1s);
        }
    }
};

t.request_stop();
// OR, through the stop source:
std::stop_source stopSource = t.get_stop_source();
stopSource.request_stop();
\end{lstlisting}

\subsubsection{\href{https://github.com/AnthonyCalandra/modern-cpp-features/tree/master\#safe-integral-comparisons}
                    {safe integral comparisons}}
Compare integers, including those of distinct types, without the dangers of 
integer conversion.
\begin{lstlisting}[language=C++]
-1 > 0U; // == true
std::cmp_greater(-1, 0U); // == false

std::cmp_equal(0U, 0); // == true
std::cmp_less_equal(-1, 1U); // == true

std::in_range<unsigned>(-1); // == false
std::in_range<char>(999999); // == false
\end{lstlisting}

\subsection{C++23 (to expand)}
\label{subsec:cpp23-planned}
C++23 material will be added selectively later (e.g.\ \texttt{std::expected},
\texttt{std::print}, deducing \texttt{this}, flat containers / \texttt{mdspan} as relevant). 
Until then, use the upstream cheatsheet and cppreference. C++26 waits until the standard is stable.

