\chapter{Low Latency Engineering}
\label{chap:lowlatency}

Earlier chapters give the building blocks: pinning and scheduling 
(Chapter~\ref{chap:linuxprocesses}), caches and false sharing 
(Chapter~\ref{chap:cacheperformance}), NUMA (Chapter~\ref{chap:numa}), 
non-blocking I/O and \texttt{epoll} (Chapters~\ref{chap:linuxnetworking} 
and~\ref{chap:eventdrivenio}), and kernel bypass when the stack itself is the limit 
(Chapter~\ref{chap:kernelbypass}). 
This chapter is the \highlight{ops checklist}: isolate cores, poll instead of sleep, 
keep the hot path cache-friendly and allocation-free, and optimize for latency even when 
that hurts bulk throughput.

\section{CPU Isolation}
\label{sec:cpu-isolation}
Assumes per-thread pinning from Section~\ref{subsec:sched-setaffinity}. 
This section covers kernel boot parameters that isolate cores from general scheduler noise.
\newline
Pinning alone is not enough if the kernel still steals your core for timers, RCU callbacks, 
or unrelated threads. Isolation makes a core \highlight{mostly yours}.

\subsection{\texttt{isolcpus}}
\label{subsec:isolcpus}
Boot parameter that removes listed CPUs from the general scheduler's regular balancing pool 
(exact semantics depend on kernel version and flags)~\cite{kernel-isolcpus}. 
You then \texttt{sched\char`_setaffinity} your hot threads onto those CPUs.
\newline
Typical pattern: housekeeping on CPUs 0--$N$, strategy/feed on isolated CPUs $N{+}1$ onward. 
Without also pinning IRQs and other noise away, isolation is incomplete 
(Chapter~\ref{chap:linuxprocesses}).

\subsection{\texttt{nohz\_full}}
\label{subsec:nohz-full}
\texttt{nohz\char`_full=}~\cite{kernel-nohz} reduces scheduler timer ticks on listed CPUs when 
only one runnable task is present --- less periodic interruption on a quiet isolated core. 
Needs a kernel built with full dynticks support; validate on your distro. 
Pair with \texttt{isolcpus} and a single pinned busy-loop or poll thread per core.

\subsection{\texttt{rcu\_nocbs}}
\label{subsec:rcu-nocbs}
\texttt{rcu\char`_nocbs=}~\cite{kernel-rcu} (and related \texttt{rcu\char`_nocb\char`_poll}) moves 
RCU callback processing off the listed CPUs onto housekeeping cores. 
Otherwise RCU softirqs can inject jitter on your ``isolated'' CPU. 
Treat isolation as a \highlight{set}: \texttt{isolcpus} + \texttt{nohz\char`_full} + 
\texttt{rcu\char`_nocbs} + IRQ affinity + pinned threads.

\section{Busy Polling}
\label{sec:busy-polling}
Builds on non-blocking sockets (Section~\ref{sec:blocking-vs-nonblocking-io}) and \texttt{epoll} 
(Chapter~\ref{chap:eventdrivenio}). Trades CPU usage for lower wake-up latency.
\newline
Blocking \texttt{epoll\char`_wait} / \texttt{recv} sleeps the thread; wake-up adds 
unpredictable delay. Alternatives on the critical path:
\begin{itemize}
    \item \texttt{epoll\char`_wait} with timeout \texttt{0} in a tight loop (spin until ready).
    \item Socket \highlight{busy poll} options (\texttt{SO\char`_BUSY\char`_POLL} and related 
          sysctls) so \texttt{recv}/\texttt{poll} spin briefly in the kernel networking stack 
          before sleeping~\cite{man-busy-poll}.
    \item User-space poll loops on bypass rings (Chapter~\ref{chap:kernelbypass}).
\end{itemize}
Cost: a core at $\sim$100\% for as long as you spin. Only do this on isolated CPUs dedicated 
to that job.
\par\noindent\textbf{Interview tip:} sleep = jitter; busy poll = latency for watts and opportunity cost.

\section{Cache Optimization}
\label{sec:lowlatency-cache-optimization}
See Chapter~\ref{chap:cacheperformance} for cache hierarchy, false sharing, and 
\texttt{perf}-based profiling. This section covers operational cache tuning only.
\begin{itemize}
    \item Keep hot structures within few cache lines; align and pad to avoid false sharing.
    \item Prefer sequential access and preallocated contiguous buffers (rings, arenas).
    \item Pin thread and memory on the same NUMA node as the NIC (Chapter~\ref{chap:numa}).
    \item Warm critical code and data at startup (touch pages, run a dry path) to avoid 
          cold-cache and major-fault surprises (Chapter~\ref{chap:memorymapping}).
    \item Measure with \texttt{perf} --- do not guess which miss hurts the tick-to-trade path.
\end{itemize}

\section{Avoiding Allocations}
\label{sec:avoiding-allocations}
See Section~\ref{sec:spsc-ring-buffer} for pre-allocated SPSC ring buffers. 
This section covers general allocation avoidance on hot paths.
\newline
\texttt{new}/\texttt{malloc}, \texttt{std::string} growth, and hidden temporaries cause:
\begin{itemize}
    \item Unbounded latency (allocator locks, syscalls, page faults),
    \item Cache/TLB disruption,
    \item Jitter under load that never shows in a quiet unit test.
\end{itemize}
Practices:
\begin{itemize}
    \item Preallocate pools and rings at startup; recycle objects 
          (Section~\ref{subsec:memory-pools}).
    \item Fixed-size messages; no unbounded buffering on the hot thread.
    \item Defer logging, formatting, and allocations to a colder thread via a lock-free queue.
    \item Prefer stack/static storage or arenas with LIFO free for short-lived data.
\end{itemize}

\section{Latency vs Throughput}
\label{sec:latency-vs-throughput}
\begin{center}
\begin{tabular}{|l|p{5.8cm}|p{5.8cm}|}
\hline
\textbf{Goal} & \textbf{Optimize for} & \textbf{Often sacrifices} \\
\hline
Latency & Median and tail response time; 
  predictable wake-ups; short paths & CPU utilization, batching efficiency, 
  multi-tenant density \\
\hline
Throughput & Messages/sec or bandwidth; 
  batching, coalescing, deep queues & Per-message delay; tail latency under load \\
\hline
\end{tabular}
\end{center}
HFT market-data and order paths are usually \highlight{latency-first}: busy poll, 
\texttt{TCP\char`_NODELAY}, shallow queues, one connection per hot thread, isolation. 
A bulk file transfer or analytics job is throughput-first: large buffers, interrupts 
coalesced, sleep when idle.
\par\noindent\textbf{Interview tip:} know which axis you are on. Techniques that maximize packets/sec 
(Nagle, interrupt coalescing, shared noisy cores) often \highlight{hurt} tick-to-trade 
latency --- and the reverse wastes capacity if you only care about bulk bandwidth.
