\chapter{Linux Processes and Threads}
\label{chap:linuxprocesses}

From a C++ program, concurrency starts with \texttt{std::thread} (Chapter~\ref{chap:concurrency}). 
Underneath, the operating system still sees \highlight{processes} and \highlight{threads}: separate 
schedulable entities with PIDs, address spaces, and CPU time. 
This chapter connects the C++ view to the Linux view, which matters for latency tuning 
(affinity, scheduling classes, syscall cost).

\section{Processes vs Threads}
\label{sec:processes-vs-threads}

\subsection{Processes}
\label{subsec:processes}
A \highlight{program} is an executable file on disk. A \highlight{process} is that program 
\highlightit{loaded into memory and running}: the OS gives it a \highlightbf{PID} 
(\emph{process identifier}), private resources, and its own virtual address space 
(Section~\ref{sec:address-space}).
\newline
Creating a new process (e.g.\ via \texttt{fork(2)}~\cite{man-fork}) duplicates the address 
space: the child gets a new PID and a copy-on-write view of memory. 
Multiprocess designs (separate feed handler and strategy engine) communicate via IPC 
(pipes, shared memory --- Chapter~\ref{chap:memorymapping}), not by sharing heap pointers.
\newline
Every C++ program starts as \highlight{one process}. \texttt{main()} runs on the process's 
\highlight{initial thread} (also called the \highlight{main thread}).

\subsection{Threads}
\label{subsec:threads}
See Chapter~\ref{chap:concurrency} for the C++ thread API (\texttt{std::thread}, \texttt{std::jthread}, 
and related types). Here we focus on what the \highlight{kernel} sees.
\newline
A \highlight{thread} is a unit of execution \highlightit{inside} a process. 
Threads in the same process \highlight{share} the virtual address space (heap, globals, code); 
each thread has its own stack, registers, and program counter. 
The kernel schedules threads (Linux treats each thread as a \highlight{task} with its own TID).
\newline
On Linux, \texttt{std::thread} is typically implemented with \highlight{pthreads}. 
Pthread creation ultimately uses the \texttt{clone(2)}~\cite{man-clone} system call with 
flags that share memory with the parent --- a new thread, not a new process.
\newline
You do \highlight{not} need to learn the full pthread API (\texttt{pthread\_create}, \texttt{void*}, 
attributes). Use \texttt{std::thread} for structure and lifetime; use \texttt{native\_handle()} 
only for Linux-specific tuning (affinity, real-time priority) in Section~\ref{sec:thread-affinity}.

\begin{lstlisting}[language=C++]
#include <iostream>
#include <thread>
#include <unistd.h>  // getpid()

int main()
{
    const pid_t pid = getpid();
    std::cout << "Process PID: " << pid << "\n";
    std::cout << "Main std::thread::id: " << std::this_thread::get_id() << "\n";

    std::thread worker([pid] {
        std::cout << "Worker std::thread::id: "
                  << std::this_thread::get_id() << "\n";
        std::cout << "Worker sees same PID: " << getpid()
                  << " (expected " << pid << ")\n";
    });
    worker.join();
}
\end{lstlisting}

\noindent
\texttt{std::thread::get\_id()} is a C++ identifier; \texttt{getpid()}~\cite{man-getpid} returns 
the OS process ID. All threads print the \highlight{same PID} because they belong to one process.

\section{Address Space}
\label{sec:address-space}
Each process has its own \highlight{virtual address space}: a range of virtual addresses mapped 
by the MMU to physical RAM (or swap). The layout usually includes:
\begin{itemize}
    \item \highlightbf{Code} (\texttt{.text}) --- machine instructions, often read-only.
    \item \highlightbf{Data} --- globals and statics; \highlightbf{heap} (dynamic allocation); 
          \highlightbf{stack} per thread (function frames, local variables).
    \item \highlightbf{Kernel space} --- not directly accessible from user code; entered on syscalls.
\end{itemize}
Memory is managed in \highlight{pages} (commonly 4~KiB on x86-64). 
\highlight{All threads of a process share the same virtual mapping} for heap and globals; 
each thread's stack grows in the same address space but on separate stacks.
\newline
For translation cost (TLB), see Section~\ref{sec:tlb}. 
For demand paging and page faults, see Section~\ref{subsec:page-faults}.

\section{User Mode vs Kernel Mode}
\label{sec:user-kernel-mode}
The CPU runs in \highlight{user mode} when executing your C++ code. 
Sensitive operations (I/O, scheduling, memory mapping, locking in the kernel) require a 
\highlight{system call}: the CPU switches to \highlight{kernel mode}, runs the kernel handler, 
and returns to user mode.
\newline
Examples that trap into the kernel from a typical trading stack:
\begin{itemize}
    \item \texttt{read}/\texttt{recv} on a socket (Chapter~\ref{chap:linuxnetworking}).
    \item \texttt{mmap} for shared memory (Chapter~\ref{chap:memorymapping}).
    \item \texttt{pthread\_setaffinity\_np} / \texttt{sched\_setaffinity} to pin threads (Section~\ref{subsec:sched-setaffinity}).
    \item Blocking mutex or I/O wait --- the thread may be descheduled inside the kernel.
\end{itemize}
Each user/kernel transition has \highlight{latency cost} (flush pipeline, TLB effects, cache coldness). 
Low-latency paths minimize syscalls and blocking on hot paths; kernel bypass (Chapter~\ref{chap:kernelbypass}) 
exists partly to avoid this stack for networking.

\section{Linux Scheduler}
\label{sec:linux-scheduler}
The \highlight{scheduler} decides \highlight{which thread runs on which CPU core} and for how long. 
You do not pick this in portable C++; the kernel does, unless you influence it with affinity, 
scheduling policy, or isolation (Sections~\ref{sec:thread-affinity} and~\ref{sec:cpu-isolation}).

\subsection{Context Switch}
\label{subsec:context-switch}
A \highlight{context switch} occurs when the CPU stops running one thread and runs another:
\begin{itemize}
    \item Save the current thread's registers and program counter into its \highlight{task struct}.
    \item Restore another thread's saved state.
    \item Optionally migrate execution to a different core ( \highlight{cache cold start} --- Chapter~\ref{chap:cacheperformance}).
\end{itemize}
Context switches are \highlight{expensive} compared to a function call: microseconds, not nanoseconds, 
and they add \highlight{jitter} --- a problem for HFT latency budgets. 
Pinning threads (Section~\ref{sec:thread-affinity}) reduces \highlight{migration} between cores.

\subsection{Scheduling Classes}
\label{subsec:scheduling-classes}
Linux exposes several scheduling policies~\cite{man-sched}:
\begin{itemize}
    \item \highlightbf{SCHED\_OTHER} (default) --- CFS (\emph{Completely Fair Scheduler}). 
          General-purpose, time-sharing; good throughput, \highlight{not} hard latency guarantees.
    \item \highlightbf{SCHED\_BATCH}, \highlightbf{SCHED\_IDLE} --- lower priority background work.
    \item \highlightbf{SCHED\_FIFO}, \highlightbf{SCHED\_RR} --- real-time classes (Section~\ref{subsec:realtime-scheduling}).
\end{itemize}
Most C++ threads run under CFS unless you explicitly request a real-time policy with 
\texttt{sched\_setscheduler(2)}~\cite{man-sched-setscheduler}.

\subsection{Real-Time Scheduling}
\label{subsec:realtime-scheduling}
\highlight{SCHED\_FIFO} and \highlight{SCHED\_RR} run at higher priority than normal threads. 
\highlight{SCHED\_FIFO} runs until it blocks or yields; \highlight{SCHED\_RR} adds a time quantum 
among equal-priority RT threads.
\newline
\highlight{Use with care}: a busy RT thread can \highlight{starve} the rest of the system (including 
housekeeping and logging). Typical HFT setups combine \highlight{RT policy on dedicated cores} with 
\highlight{isolcpus} (Section~\ref{sec:cpu-isolation}) rather than marking every thread RT.
\newline
To raise priority on an existing \texttt{std::thread}, use \texttt{pthread\_setschedparam}~\cite{man-pthread-setschedparam} on 
\texttt{native\_handle()} (requires appropriate privileges):
\begin{lstlisting}[language=C++]
#include <pthread.h>
#include <sched.h>

void set_fifo_priority(std::thread& t, int priority)
{
    sched_param param{};
    param.sched_priority = priority;  // 1..99 for SCHED_FIFO on Linux
    pthread_setschedparam(t.native_handle(), SCHED_FIFO, &param);
}
\end{lstlisting}

\section{Thread Affinity}
\label{sec:thread-affinity}

\subsection{CPU Affinity and Thread Pinning}
\label{subsec:cpu-affinity-thread-pinning}
\highlight{CPU affinity} restricts \highlight{which cores} a thread may run on. 
\highlight{Thread pinning} means setting affinity to a \highlight{single core} (or a small fixed set) 
so the thread stays cache-hot and avoids scheduler migration.
\newline
By default, Linux may \highlight{migrate} threads across cores to balance load. That causes:
\begin{itemize}
    \item \highlight{Cache loss} --- data warmed in L1/L2 on the old core is not reused on the new one 
          (Chapter~\ref{chap:cacheperformance}).
    \item \highlight{Extra context switches} --- bad for microsecond-scale budgets 
          (Section~\ref{subsec:context-switch}).
\end{itemize}
Why it matters for HFT:
\begin{itemize}
    \item \highlight{Less jitter} --- fewer context switches and core hops.
    \item \highlight{Warmer caches} --- L1/L2 data stays valid on one core.
    \item \highlight{Predictable layout} --- dedicate cores to feed, strategy, and gateway roles.
\end{itemize}

\highlightbf{Avoid pinning hot paths to core 0} on typical Linux servers: core 0 often handles many 
hardware interrupts (IRQs), timers, and kernel housekeeping, which adds \highlight{non-deterministic jitter}. 
Prefer cores reserved for the application (Section~\ref{sec:cpu-isolation}).
\newline
\highlightbf{Hyper-threading (HT):} two logical CPUs may share one physical core (siblings). 
For maximum isolation, pin latency-critical threads to \highlight{one logical CPU per physical core} 
(or disable HT in BIOS on dedicated trading boxes). Inspect topology with \texttt{lscpu -e} 
(Section~\ref{subsec:cpu-topology}).
\newline
Pinning APIs are in the next subsection. 
For NUMA node placement and \texttt{libnuma}, see Section~\ref{subsec:numa-thread-cpu-binding}. 
For OS-level core isolation (\texttt{isolcpus}), see Section~\ref{sec:cpu-isolation}. 
For moving NIC interrupts off trading cores, see Section~\ref{subsec:irq-affinity-overview}.

\subsection[Topology]{Machine Topology and Core Selection}
\label{subsec:cpu-topology}
Before pinning, inspect the machine. Useful commands:
\begin{itemize}
    \item \texttt{lscpu -e} --- table of logical CPUs with \texttt{NODE}, \texttt{SOCKET}, \texttt{CORE}.
    \item \texttt{lstopo} (hwloc) --- visual map of cores, caches, and NUMA nodes.
    \item \texttt{lspci | grep -i ether} --- locate the \highlight{NIC} 
          (\emph{Network Interface Card}, the Ethernet hardware).
\end{itemize}
On a single-socket laptop, all CPUs often share \texttt{NODE 0}; multi-socket servers need the thread 
and its memory on the \highlight{same NUMA node} as the NIC (Chapter~\ref{chap:numa}).
\newline
Example workflow for a PCI Ethernet device at \texttt{05:00.0}:
\begin{enumerate}
    \item Read NUMA node: \texttt{cat /sys/bus/pci/devices/0000:05:00.0/numa\_node}.
    \item If the value is \texttt{-1}, the kernel reports no specific node (common on consumer hardware); 
          treat as node \texttt{0} on single-node systems.
    \item List CPUs on that node: \texttt{cat /sys/devices/system/node/node0/cpulist} 
          (may show ranges such as \texttt{0-15}).
    \item Pin trading threads to cores in that list, usually \highlight{avoiding core 0} and HT siblings.
\end{enumerate}
Typical role split on a tuned server: one core for \highlight{network IRQ / RX}, separate cores for 
\highlight{strategy logic}, with generic OS work kept off isolated cores.

\subsection[IRQ Overview]{Network Interrupts}
\label{subsec:irq-affinity-overview}
When a packet arrives, the NIC raises an \highlight{interrupt (IRQ)}. The kernel runs an interrupt 
handler, often on a core that may \highlight{preempt} your trading thread --- another source of jitter.
\newline
Inspect distribution with \texttt{cat /proc/interrupts} (look for your driver name, e.g.\ \texttt{eth0}). 
Columns are per-CPU interrupt counts. If core 0 handles most NIC interrupts, consider moving IRQ affinity 
(\texttt{/proc/irq/<N>/smp\_affinity\_list}) to a \highlight{dedicated non-trading core}. 
This requires root and careful validation; combine with \texttt{isolcpus} (Section~\ref{sec:cpu-isolation}) 
and busy polling (Section~\ref{sec:busy-polling}) where appropriate.

\subsection[Pinning APIs]{Pinning with \texttt{pthread\_setaffinity\_np} and \texttt{sched\_setaffinity}}
\label{subsec:sched-setaffinity}
There are two common Linux APIs for the same idea:
\begin{itemize}
    \item \highlightbf{\texttt{pthread\_setaffinity\_np(3)}}~\cite{man-pthread-setaffinity} --- 
          pthread layer; natural fit with \texttt{std::thread::native\_handle()}.
    \item \highlightbf{\texttt{sched\_setaffinity(2)}}~\cite{man-sched-setaffinity} --- 
          scheduler syscall; takes a kernel \highlight{TID} (\emph{thread ID}). 
          On Linux, \texttt{pthread\_setaffinity\_np} typically wraps this call.
\end{itemize}
Both use \texttt{cpu\_set\_t} and macros \texttt{CPU\_ZERO}, \texttt{CPU\_SET}, etc. 
Compile with \texttt{\#define \_GNU\_SOURCE} before includes for the pthread extension.
\newline
\texttt{cpu\_set\_t} is a \highlight{bitmask}: bit \texttt{n} set means the thread \highlight{may} run on 
logical CPU \texttt{n}. \texttt{CPU\_ZERO} clears all bits; each \texttt{CPU\_SET(core, \&cpuset)} 
\highlight{sets} one bit (calling it in a loop \highlight{accumulates} allowed cores --- it does not replace 
the whole set). \texttt{pthread\_setaffinity\_np} tells the kernel the allowed mask for that thread.

\begin{lstlisting}[language=C++]
#define _GNU_SOURCE
#include <pthread.h>
#include <sched.h>

#include <vector>

bool pin_current_thread_to_cpus(const std::vector<int>& cores)
{
    cpu_set_t cpuset;
    CPU_ZERO(&cpuset);
    for (int core : cores) {
        CPU_SET(core, &cpuset);
    }
    return pthread_setaffinity_np(pthread_self(), sizeof(cpu_set_t), &cpuset) == 0;
}

// Example: allow cores 4-7 only (avoid core 0 on a 16-logical-CPU box)
// pin_current_thread_to_cpus({4, 5, 6, 7});
\end{lstlisting}

\begin{lstlisting}[language=C++]
#define _GNU_SOURCE
#include <pthread.h>
#include <sched.h>

#include <cerrno>
#include <cstring>
#include <iostream>
#include <thread>

// Pin the calling pthread to a single CPU (0-based index).
bool pin_current_pthread_to_cpu(int cpu)
{
    cpu_set_t cpuset;
    CPU_ZERO(&cpuset);
    CPU_SET(cpu, &cpuset);

    const int rc = pthread_setaffinity_np(
        pthread_self(), sizeof(cpu_set_t), &cpuset);
    if (rc != 0) {
        std::cerr << "pthread_setaffinity_np: " << std::strerror(rc) << "\n";
        return false;
    }
    return true;
}

// Pin an existing std::thread (call after thread is running).
bool pin_std_thread_to_cpu(std::thread& t, int cpu)
{
    cpu_set_t cpuset;
    CPU_ZERO(&cpuset);
    CPU_SET(cpu, &cpuset);

    const int rc = pthread_setaffinity_np(
        t.native_handle(), sizeof(cpu_set_t), &cpuset);
    if (rc != 0) {
        std::cerr << "pthread_setaffinity_np: " << std::strerror(rc) << "\n";
        return false;
    }
    return true;
}

int main()
{
    std::thread worker([] {
        if (!pin_current_pthread_to_cpu(2)) {
            return;
        }
        // Hot loop: strategy or feed handler work on core 2...
    });

    worker.join();
}
\end{lstlisting}

\noindent
The same pinning via the syscall directly (useful when you have a TID or no pthread handle):

\begin{lstlisting}[language=C++]
#include <sched.h>
#include <sys/syscall.h>
#include <unistd.h>

#include <cstdint>

pid_t linux_tid()
{
    return static_cast<pid_t>(syscall(SYS_gettid));
}

bool pin_current_thread_via_sched(int cpu)
{
    cpu_set_t cpuset;
    CPU_ZERO(&cpuset);
    CPU_SET(cpu, &cpuset);
    return sched_setaffinity(linux_tid(), sizeof(cpu_set_t), &cpuset) == 0;
}
\end{lstlisting}

\noindent
Verify affinity with \texttt{sched\_getaffinity(2)}~\cite{man-sched-getaffinity} or 
by reading \texttt{/proc/self/task/<tid>/status} (\texttt{Cpus\_allowed\_list}). 
In production, pinning is combined with \highlight{dedicated cores} (\texttt{isolcpus}, 
Section~\ref{sec:cpu-isolation}) and \highlight{NUMA-local} memory (Chapter~\ref{chap:numa}) 
so the pinned thread's data stays on the same socket.
