\chapter{Software Testing}
\label{chap:testing}

\section{Seven Tesing Principles}

\section{SDLC vs. STLC}

\section{Python Unit Testing}

\subsection{\texttt{unittest}}
\href{https://docs.python.org/3/library/unittest.html}{\texttt{unittest} — Unit testing framework}

It supports test automation, sharing of setup and shutdown code for tests, 
aggregation of tests into collections, and independence of the tests from 
the reporting framework.

To achieve this, unittest supports some important concepts in an 
object-oriented way:

\begin{itemize}
    \item \highlightbf{test fixture}
    A test fixture represents the \highlightit{preparation needed} to perform one or 
    more tests, and any associated \highlightit{cleanup actions}. This may involve, 
    for example, creating temporary or proxy databases, directories, 
    or starting a server process.

    \item \highlightbf{test case}
    A test case is the \highlightit{individual unit of testing}. It checks for a 
    specific response to a particular set of inputs. \highlighttt{unittest} provides a 
    base class, \highlighttt{TestCase}, which may be used to create new test cases.

    \item \highlightbf{test suite}
    A test suite is a \highlightit{collection of test cases, test suites, or both}. I1t 
    is used to aggregate tests that should be executed together.

    \item \highlightbf{test runner}
    A test runner is a \highlightit{component which orchestrates} the execution of 
    tests and provides the outcome to the user. The runner may use a 
    graphical interface, a textual interface, or return a special value 
    to indicate the results of executing the tests.
\end{itemize}


A testcase is created by subclassing 
\href{https://docs.python.org/3/library/unittest.html#unittest.TestCase}{\texttt{unittest.TestCase}} 
or using \href{https://docs.python.org/3/library/unittest.html#unittest.FunctionTestCase}{FunctionTestCase} for legacy compatibility\\
Child class methods' names start by \texttt{test\char`_} as a naming convention 
to inform the test runner. 

\href{https://docs.python.org/3/library/unittest.html#unittest.main}{\texttt{unittest.main}} 
provides a command line interface to execute tests.

In the following tables~\ref{tab:assertmethod},~\ref{tab:exceptionassertmethods},~\ref{tab:morespecificassertmethods} 
and~\ref{tab:typespecificassertmethods} a \highlight{list of assert methods provided 
by \texttt{TestCase}}

\begin{table}[h!t]
    \begin{tabularx}{\textwidth}{|X|l|l|}
    \hline
    Method & Checks that & New in \\
    \hline
    \href{https://docs.python.org/3/library/unittest.html\#unittest.TestCase.assertEqual}
         {\texttt{assertEqual(a, b)}} & \highlighttt{a $==$ b} & \\
    \hline
    \href{https://docs.python.org/3/library/unittest.html\#unittest.TestCase.assertNotEqual}
         {\texttt{assertNotEqual(a, b)}} & \highlighttt{a $!=$ b} & \\
    \hline
    \href{https://docs.python.org/3/library/unittest.html\#unittest.TestCase.assertTrue}
         {\texttt{assertTrue(x) bool(x)}} & \highlighttt{is True} & \\
    \hline
    \href{https://docs.python.org/3/library/unittest.html\#unittest.TestCase.assertFalse}
         {\texttt{assertFalse(x) bool(x)}} & \highlighttt{is False} & \\
    \hline
    \href{https://docs.python.org/3/library/unittest.html\#unittest.TestCase.assertIs}
         {\texttt{assertIs(a, b)}} & \highlighttt{a is b} & 3.1 \\
    \hline
    \href{https://docs.python.org/3/library/unittest.html\#unittest.TestCase.assertIsNot}
         {\texttt{assertIsNot(a, b)}} & \highlighttt{a is not b} & 3.1 \\
    \hline
    \href{https://docs.python.org/3/library/unittest.html\#unittest.TestCase.assertIsNone}
         {\texttt{assertIsNone(x)}} & x \highlighttt{is None} & 3.1 \\
    \hline
    \href{https://docs.python.org/3/library/unittest.html\#unittest.TestCase.assertIsNotNone}
         {\texttt{assertIsNotNone(x)}} & x \highlighttt{is not None} & 3.1 \\
    \hline
    \href{https://docs.python.org/3/library/unittest.html\#unittest.TestCase.assertIn}
         {\texttt{assertIn(a, b)}} & \highlighttt{a in b} & 3.1 \\
    \hline
    \href{https://docs.python.org/3/library/unittest.html\#unittest.TestCase.assertNotIn}
         {\texttt{assertNotIn(a, b)}} & \highlighttt{a not in b} & 3.1 \\
    \hline
    \href{https://docs.python.org/3/library/unittest.html\#unittest.TestCase.assertIsInstance}
         {\texttt{assertIsInstance(a, b)}} & \highlighttt{isinstance(a, b)} & 3.2 \\
    \hline
    \href{https://docs.python.org/3/library/unittest.html\#unittest.TestCase.assertNotIsInstance}
         {\texttt{assertNotIsInstance(a, b)}} & \highlighttt{not isinstance(a, b)} & 3.2 \\
    \hline
    \end{tabularx}
    \caption{TestCase assert methods.}
    \label{tab:assertmethod}
\end{table}
\newpage
\begin{table}[h!t]
    \begin{tabularx}{\textwidth}{|l|X|l|}
        \hline
        Method & Checks that & New in\\
        \hline
        \href{https://docs.python.org/3/library/unittest.html\#unittest.TestCase.assertRaises}
             {\small \texttt{assertRaises(exc, fun, *args, **kwds)}} & \texttt{fun} raises exc & \\
        \hline
        \href{https://docs.python.org/3/library/unittest.html\#unittest.TestCase.assertRaisesRegex}
             {\small \texttt{assertRaisesRegex(exc, r, fun, *args, **kwds)}} & \texttt{fun} raises exc and the message matches regex r & 3. 1\\
        \hline
        \href{https://docs.python.org/3/library/unittest.html\#unittest.TestCase.assertWarns}
             {\small \texttt{assertWarns(warn, fun, *args, **kwds)}} & \texttt{fun} raises warn & 3.2 \\
        \hline
        \href{https://docs.python.org/3/library/unittest.html\#unittest.TestCase.assertWarnsRegex}
             {\small \texttt{assertWarnsRegex(warn, r, fun, *args, **kwds)}} & \texttt{fun} raises warn and the message matches regex r & 3.2 \\
        \hline
        \href{https://docs.python.org/3/library/unittest.html\#unittest.TestCase.assertLogs}
             {\small \texttt{assertLogs(logger, level)}} & The with block logs on logger with minimum level & 3.4 \\
        \hline
        \href{https://docs.python.org/3/library/unittest.html\#unittest.TestCase.assertNoLogs}
                {\small \texttt{assertNoLogs(logger, level)}} & The with block does not log on logger with minimum level & 3.10 \\
        \hline
    \end{tabularx}
    \caption{TestCase methods to check the production of exceptions.}
    \label{tab:exceptionassertmethods}
\end{table}

\begin{table}[h!t]
    \begin{tabularx}{\textwidth}{|X|X|l|}
        \hline
        Method & Checks that & New in \\
        \hline
        \href{https://docs.python.org/3/library/unittest.html\#unittest.TestCase.assertAlmostEqual}
             {\texttt{assertAlmostEqual(a, b)}} & \highlighttt{\texttt{round(a-b, 7)}} $==$ 0 & \\
        \hline
        \href{https://docs.python.org/3/library/unittest.html\#unittest.TestCase.assertNotAlmostEqual}
             {\texttt{assertNotAlmostEqual(a, b)}} & \highlighttt{\texttt{round(a-b, 7)}} $!=$ 0 & \\
        \hline
        \href{https://docs.python.org/3/library/unittest.html\#unittest.TestCase.assertGreater}
             {\texttt{assertGreater(a, b)}} & \highlighttt{a $>$ b} & 3.1 \\
        \hline
        \href{https://docs.python.org/3/library/unittest.html\#unittest.TestCase.assertGreaterEqual}
             {\texttt{assertGreaterEqual(a, b)}} & \highlighttt{a $>=$ b} & 3.1 \\
        \hline
        \href{https://docs.python.org/3/library/unittest.html\#unittest.TestCase.assertLess}
             {\texttt{assertLess(a, b)}} & \highlighttt{a $<$ b} & 3.1 \\
        \hline
        \href{https://docs.python.org/3/library/unittest.html\#unittest.TestCase.assertLessEqual}
             {\texttt{assertLessEqual(a, b)}} & \highlighttt{a $<=$ b} & 3.1 \\
        \hline
        \href{https://docs.python.org/3/library/unittest.html\#unittest.TestCase.assertRegex}
             {\texttt{assertRegex(s, r)}} & \highlighttt{\texttt{r.search(s)}} & 3.1 \\
        \hline
        \href{https://docs.python.org/3/library/unittest.html\#unittest.TestCase.assertNotRegex}
             {\texttt{assertNotRegex(s, r)}} & \highlighttt{\texttt{not r.search(s)}} & 3.2 \\
        \hline
        \href{https://docs.python.org/3/library/unittest.html\#unittest.TestCase.assertCountEqual}
             {\texttt{assertCountEqual(a, b)}} & a and b have the \highlighttt{same elements} 
        in the \highlighttt{same number}, regardless of their order. & 3.2 \\
        \hline
    \end{tabularx}
    \caption{TestCase methods to perform more specific checks}
    \label{tab:morespecificassertmethods}
\end{table}

\newpage
\begin{table}[h!t]
    \begin{tabularx}{\textwidth}{|l|X|X|}
        \hline
        Method & Used to compare & New in\\
        \hline
        \href{https://docs.python.org/3/library/unittest.html\#unittest.TestCase.assertMultiLineEqual}
             {\texttt{assertMultiLineEqual(a, b)}} & strings & 3.1 \\
        \hline
        \href{https://docs.python.org/3/library/unittest.html\#unittest.TestCase.assertSequenceEqual}
             {\texttt{assertSequenceEqual(a, b)}} & sequences & 3.1 \\
        \hline
        \href{https://docs.python.org/3/library/unittest.html\#unittest.TestCase.assertListEqual}
             {\texttt{assertListEqual(a, b)}} & lists & 3.1 \\
        \hline
        \href{https://docs.python.org/3/library/unittest.html\#unittest.TestCase.assertTupleEqual}
             {\texttt{assertTupleEqual(a, b)}} & tuples & 3.1 \\
        \hline
        \href{https://docs.python.org/3/library/unittest.html\#unittest.TestCase.assertSetEqual}
             {\texttt{assertSetEqual(a, b)}} & sets or frozensets & 3.1 \\
        \hline
        \href{https://docs.python.org/3/library/unittest.html\#unittest.TestCase.assertDictEqual}
             {\texttt{assertDictEqual(a, b)}} & dicts & 3.1 \\
        \hline
    \end{tabularx}
    \caption{TesCase type-specific methods automatically used by \texttt{assertEqual}}
    \label{tab:typespecificassertmethods}
\end{table}

\subsubsection{Organizing test code}

\nparagraph{\href{https://docs.python.org/3/library/unittest.html\#unittest.TestCase.setUp}
           {\texttt{unittest.TestCase.setUp()}}}
In case of a considerable number of tests to run, it is possible to 
factor out the set-up using .

The \highlighttt{TestCase.setUp} method does nothing on default, but 
can be overridden by child classes in order to prepare the \highlight{test fixture}.

Any exception different than \href{https://docs.python.org/3/library/exceptions.html\#AssertionError}{\texttt{AssertionError}} 
or \href{https://docs.python.org/3/library/unittest.html\#unittest.SkipTest}{\texttt{SkipTest}} 
raised by this method is interpreted as an error rather than a test failure.

\nparagraph{\href{https://docs.python.org/3/library/unittest.html\#unittest.TestCase.tearDown}
           {\texttt{unittest.TestCase.tearDown()}}} 
The \texttt{TestCase.tearDown()} method is called immediately after 
the test results are recorded, even if the test method raised an exception. 
The only requirement for this method to be called is the success of \texttt{setUp()} call.

\subsubsection{Grouping Tests}

\subsection{\texttt{doctest}}
\href{https://docs.python.org/3/library/doctest.html#module-doctest}{doctest}


\section{C++ Unit Testing}
\subsection{\texttt{gMock}}
\href{http://google.github.io/googletest/gmock_for_dummies.html}{gMock for Dummies}

\href{http://google.github.io/googletest/gmock_cook_book.html}{gMock Cookbook}

%~ \section{Integration Test and Regression Test}
\section{Integration Test}
Integration testing \highlightit{checks the communication and data flow between 
two modules}
\newline 
\highlightit{of the same system after integrating them}. \\
A software consists of \highlightit{several small modules} that carry certain features 
and come together to build the system. 
Sometimes even when these modules pass the quality test independently, 
they \highlightit{might encounter data flow issues upon integration}. 
\newline
Integration testing verifies the \highlightit{compatibility between these individual 
parts} and ensures they work in unison \highlightit{without any issues after integration}. 
It can be a black-box testing or white-box testing technique.
\newline
\highlightit{An example} of integration testing would be checking the logical connection 
between the ‘Proceed to Payment’ page and the payment gateway for an 
e-commerce application.

\section{Regression Test}
Regression testing \highlightit{re-checks 
software or an application after new changes are introduced}. 
This testing process runs \highlightit{functional and non-functional checks on the 
existing features}
\newline
\highlightit{of the software to ensure that no issues arise due 
to new code changes}.
\newline
The goal of regression testing is \highlightit{to identify 
any defects in the present system that may}
\newline
\highlightit{occur due to updates/changes}. 
Regression testing can either be white box testing or black box testing.
\newline
You can \highlightit{re-use the test cases} in regression testing to verify the existing 
functionalities. This includes running the same test cases and scripts 
to get the same output that validates the present features.
\newline
\highlightit{An example} of regression testing is checking the Camera, Phone, and other 
applications in your mobile after a system update. Smartphones often receive 
regular system updates that can hamper the ongoing functions of the phone. 
Regression testing would assess if any OS or parts upgrade impacted your phone.

\section{Behavior-Driven Development}

\section{terms from the youtube video}
\begin{itemize}
    \item decision coverage
    \item predicate coverage
    \item all path coverage
    \item boundary coverage
    \item black box testing
    \item white box testing
\end{itemize}

\section{Continous Integration Systems}
\begin{itemize}
    \item \href{https://buildbot.net/}{BuildBot}
    \item \href{https://www.jenkins.io/}{Jenkins}
    \item \href{https://github.com/features/actions}{GitHub Actions}
    \item \href{https://www.appveyor.com/}{\texttt{AppVeyor}}
\end{itemize}
